### 2022-04-18 Project_draft.Rmd
Add files via upload
Imported data and changed column names

---
title: "Project draft"
author: "Alison Cheek"
date: "4/13/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(dplyr)
library(ggplot2)

```

#Import air pollution deaths csv and tidy

```{r}

deaths_df <- data.frame(read.csv("death-rates-from-air-pollution.csv", stringsAsFactors = FALSE))

glimpse(deaths_df)

colnames(deaths_df)

colnames(deaths_df) <- c("country", "acronym", "year", "total_deaths", 
                         "indoor_deaths", "outdoor_deaths", "ozone_deaths")

```

#Explore data
```{r}
#glimpse(deaths_df)

deaths_df %>% filter(country=="United States") %>%
  select(year, indoor_deaths) %>% 
  ggplot() + 
  geom_point(aes(x=year, y=indoor_deaths))

deaths_df %>% filter(country=="United States") %>%
  select(year, outdoor_deaths) %>% 
  ggplot() + 
  geom_point(aes(x=year, y=outdoor_deaths))

deaths_df %>% filter(country=="United States") %>%
  select(year, ozone_deaths) %>% 
  ggplot() + 
  geom_point(aes(x=year, y=ozone_deaths))

```

#
```{r}
deaths_df %>% 
  select(total_deaths, indoor_deaths, outdoor_deaths, ozone_deaths) %>% 
  filter(deaths_df$acronym == "AFG") %>% 
  summarize(mean(total_deaths))

deaths_df %>% 
  select(total_deaths, indoor_deaths, outdoor_deaths, ozone_deaths) %>% 
  filter(deaths_df$acronym == "USA") %>% 
  summarize(mean(total_deaths))

```

#

#

#

#

#

#

### 2022-04-18 Draft2.Rmd
The data is loaded in and I added some questions to consider

---
title: "Draft2"
output: slidy_presentation
date: '2022-04-18'
---

# Load libraries 
```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
```

# Add in Data 
```{r}
read_csv('death-rates-from-air-pollution.csv')
```

# Questions to Consider

## Which country has the greatest number of deaths? The least?

## Which type of air pollution has the greatest number of deaths (indoor, outdoor, particulate, or outdoor ozone)?

## Which year had the worst air pollution (average the three diff types)?

## Which year had the worst indoor? Outdoor particulate? Outdoor ozone?

### 2022-04-18 Draft2.Rmd
the variables are 7 now
made a table for increments of 5 years from 2007-2017. having issues with the graph

@@
 ```{r}
-read_csv('death-rates-from-air-pollution.csv')
+deaths_df <- data.frame(read.csv("death-rates-from-air-pollution.csv"))
+deaths_df
+
+colnames(deaths_df) <- c("country", "acronym", "year", "total_deaths", 
+                         "indoor_deaths", "outdoor_deaths", "ozone_deaths")
+
+glimpse(deaths_df)
 ```
@@
 
-## Which country has the greatest number of deaths? The least?
+## Which country has the greatest number of deaths? The least? (dependent on population)
+
+```{r}
+# Want the range 2007-2017 
+# took this out c(2007, 2008, 2009, 2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017)
+
+#Maybe try increments...?
+#When looking at it you can see a decrease among the three years 
+death_07<- deaths_df %>% 
+  group_by(country) %>% 
+  filter(year == 2007) %>% 
+  select(total_deaths) %>% 
+  summarize(average_death_07 = mean(total_deaths))
+
+death_12<- deaths_df %>% 
+  group_by(country) %>% 
+  filter(year == 2012) %>% 
+  select(total_deaths) %>% 
+  summarize(average_death_12 = mean(total_deaths))
+
+death_17<- deaths_df %>% 
+  group_by(country) %>% 
+  filter(year == 2017) %>% 
+  select(total_deaths) %>% 
+  summarize(average_death_17 = mean(total_deaths))
+
+#try to have one cohesive table so it's easier to visualize 
+combo <- death_07 %>% 
+  mutate(death_12) %>% 
+  mutate(death_17)
+
+combo
+
+#Then graph it 
+#Just realized there's 200+ countries....the x-axis looks terrible 
+#Would be nice to color by continent but then that would mean going through each country
+ggplot(death_07, aes(x= country, y=average_death_07)) +
+  geom_point() +
+  theme(axis.text.x = element_text(angle = 15)) #yah..no rotation made it worse
+
+#Not sure if we want to use the selected few countries we looked at earlier?
+#Or if we wanted to use this as a overview of the data
+```
 
@@
 
+```{r}
+
+```
+
 ## Which year had the worst air pollution (average the three diff types)?
 
+```{r}
+
+````
+
 ## Which year had the worst indoor? Outdoor particulate? Outdoor ozone?
+
+```{r}
+
+````
+
+# Notes from first meeting
+
+## Hone in to only one year. Find data that is only population, continent data. Choose a couple countries/continents to work on
+
+## Look at industrialized vs non-industrialized countries

### 2022-04-18 Project_draft.Rmd
made a table for increments of 5 years from 2007-2017. having issues with the graph

@@
   geom_point(aes(x=year, y=ozone_deaths))
-
 ```
@@
   summarize(mean(total_deaths))
-
 ```

### 2022-04-19 Draft2.Rmd
Refined to 6 countries the tables are okay but still questions on graph

--- 2022-04-18 Draft2.Rmd
@@
 
-## Which country has the greatest number of deaths? The least?
+## Which country has the greatest number of deaths? The least? (dependent on population)
+
+```{r}
+# Want the range 2007-2017 
+# took this out c(2007, 2008, 2009, 2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017)
+
+#Maybe try increments...?
+#When looking at it you can see a decrease among the three years 
+#death_07<- deaths_df %>% 
+  #group_by(country) %>% 
+  #filter(year == 2007) %>% 
+  #select(total_deaths) %>% 
+  #summarize(average_death_07 = mean(total_deaths))
+
+#death_12<- deaths_df %>% 
+  #group_by(country) %>% 
+  #filter(year == 2012) %>% 
+  #select(total_deaths) %>% 
+  #summarize(average_death_12 = mean(total_deaths))
+
+#death_17<- deaths_df %>% 
+  #group_by(country) %>% 
+  #filter(year == 2017) %>% 
+  #select(total_deaths) %>% 
+  #summarize(average_death_17 = mean(total_deaths))
+
+#try to have one cohesive table so it's easier to visualize 
+#combo <- death_07 %>% 
+  #mutate(death_12) %>% 
+  #mutate(death_17)
+
+#combo
+
+#Then graph it 
+#Just realized there's 200+ countries....the x-axis looks terrible 
+#Would be nice to color by continent but then that would mean going through each country
+#ggplot(death_07, aes(x= country, y=average_death_07)) +
+  #geom_point() +
+  #theme(axis.text.x = element_text(angle = 15)) #yah..no rotation made it worse
+
+#SCRATCH WORK ABOVE
+
+#Not sure if we want to use the selected few countries we looked at earlier?
+#Or if we wanted to use this as a overview of the data
+
+#separation by high vs low population among continents 
+death_07_high <- deaths_df %>% 
+  filter(year == 2007, country %in% c('United States', 'Brazil', 'Ethiopia', 'Russia', 'China', 'Australia')) %>% 
+  group_by(country) %>% 
+  select(total_deaths) %>% 
+  summarize(average_death_07_high = mean(total_deaths))
+
+death_07_low<- deaths_df %>% 
+  filter(year == 2007, country %in% c('Canada', 'Argentina', 'Seychelles', 'Romania', 'Israel', 'Tonga')) %>% 
+  group_by(country) %>% 
+  select(total_deaths) %>% 
+  summarize(average_death_07_low = mean(total_deaths))
+
+#Combo table 2007
+combo_07<- death_07_high %>% 
+  mutate(death_07_low)
+combo_07
+
+death_12_high <- deaths_df %>% 
+  filter(year == 2012, country %in% c('United States', 'Brazil', 'Ethiopia', 'Russia', 'China', 'Australia')) %>% 
+  group_by(country) %>% 
+  select(total_deaths) %>% 
+  summarize(average_death_12_high = mean(total_deaths))
+
+death_12_low<- deaths_df %>% 
+  filter(year == 2012, country %in% c('Canada', 'Argentina', 'Seychelles', 'Romania', 'Israel', 'Tonga')) %>% 
+  group_by(country) %>% 
+  select(total_deaths) %>% 
+  summarize(average_death_12_low = mean(total_deaths))
+
+#Combo table 2012
+combo_12<- death_12_high %>% 
+  mutate(death_12_low)
+combo_12
+
+death_17_high <- deaths_df %>% 
+  filter(year == 2017, country %in% c('United States', 'Brazil', 'Ethiopia', 'Russia', 'China', 'Australia')) %>% 
+  group_by(country) %>% 
+  select(total_deaths) %>% 
+  summarize(average_death_17_high = mean(total_deaths))
+
+death_17_low<- deaths_df %>% 
+  filter(year == 2017, country %in% c('Canada', 'Argentina', 'Seychelles', 'Romania', 'Israel', 'Tonga')) %>% 
+  group_by(country) %>% 
+  select(total_deaths) %>% 
+  summarize(average_death_17_low = mean(total_deaths))
+
+#Combo table 2017
+combo_17<- death_17_high %>% 
+  mutate(death_17_low)
+combo_17
+
+#Graph selected countries 
+#Is there a way to color by low and high?
+combo_07 %>% 
+  ggplot(aes(x = country, y = average_death_07_low)) + 
+  geom_point()
+```
 

### 2022-04-19 Draft2.Rmd
the graphs are printed for 07. need to see how to combine them together

@@
 
+```
+
+```{r}
 #SCRATCH WORK ABOVE
@@
 #Graph selected countries 
-#Is there a way to color by low and high?
+#Is there a way to color by low and high and combine the two graphs below?
 combo_07 %>% 
   ggplot(aes(x = country, y = average_death_07_low)) + 
+  geom_point()
+
+combo_07 %>% 
+  ggplot(aes(x = country, y = average_death_07_high)) + 
   geom_point()

### 2022-04-20 Draft2.Rmd
tried organizing the countries based on air pollutant type..

@@
 ```{r}
+glimpse(deaths_df)
 
+#This is the average of all countries by type of air pollutant death
+deaths_df %>% 
+  group_by(country) %>% 
+  select(indoor_deaths, outdoor_deaths, ozone_deaths) %>% 
+  summarize(avg_indoor = mean(indoor_deaths), avg_outdoor = mean(outdoor_deaths), avg_ozone = mean(ozone_deaths))
+
+#Sort descending to see which country had the highest of each...(not sure if this is right)
+deaths_df %>% 
+  group_by(country) %>% 
+  select(indoor_deaths, outdoor_deaths, ozone_deaths) %>% 
+  summarize(avg_indoor = mean(indoor_deaths), avg_outdoor = mean(outdoor_deaths), avg_ozone = mean(ozone_deaths)) %>% 
+  arrange(desc(avg_indoor), desc(avg_outdoor), desc(avg_ozone))
 ```

### 2022-04-20 Draft2.Rmd
the sorting part got tricky. might have to separate the air pollutant types, save them into diff objects, and then mutate for them all to be in one table??

--- 2022-04-19 Draft2.Rmd
@@
 ```{r}
+glimpse(deaths_df)
 
+#This is the average of all countries by type of air pollutant death
+deaths_df %>% 
+  group_by(country) %>% 
+  select(indoor_deaths, outdoor_deaths, ozone_deaths) %>% 
+  summarize(avg_indoor = mean(indoor_deaths), avg_outdoor = mean(outdoor_deaths), avg_ozone = mean(ozone_deaths))
+
+#Sort descending to see which country had the highest of each...(not sure if this is right)
+#does separating the arrange individually fix it??
+deaths_df %>% 
+  group_by(country) %>% 
+  select(indoor_deaths, outdoor_deaths, ozone_deaths) %>% 
+  summarize(avg_indoor = mean(indoor_deaths), avg_outdoor = mean(outdoor_deaths), avg_ozone = mean(ozone_deaths)) %>% 
+  arrange(desc(avg_indoor), desc(avg_outdoor), desc(avg_ozone))
 ```
@@
 ```{r}
-
+deaths_df %>% 
+  group_by(country, year) %>% 
+  summarize(avg_indoor = mean(indoor_deaths), avg_outdoor = mean(outdoor_deaths), avg_ozone = mean(ozone_deaths)) %>% 
+  arrange(desc(avg_indoor)) %>% 
+  arrange(desc(avg_outdoor)) %>% 
+  arrange(desc(avg_ozone))
 ````
@@
 ```{r}
+#Probably need to fix the above to answer this one
 

### 2022-04-20 Draft2.Rmd
filter to selected north american countries

--- 2022-04-19 Draft2.Rmd
@@
 ```{r}
-deaths_df <- data.frame(read.csv("death-rates-from-air-pollution.csv"))
-deaths_df
-
-colnames(deaths_df) <- c("country", "acronym", "year", "total_deaths", 
-                         "indoor_deaths", "outdoor_deaths", "ozone_deaths")
-
-glimpse(deaths_df)
+read_csv('death-rates-from-air-pollution.csv')
 ```
@@
 
-## Which country has the greatest number of deaths? The least? (dependent on population)
-
-```{r}
-# Want the range 2007-2017 
-# took this out c(2007, 2008, 2009, 2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017)
-
-#Maybe try increments...?
-#When looking at it you can see a decrease among the three years 
-#death_07<- deaths_df %>% 
-  #group_by(country) %>% 
-  #filter(year == 2007) %>% 
-  #select(total_deaths) %>% 
-  #summarize(average_death_07 = mean(total_deaths))
-
-#death_12<- deaths_df %>% 
-  #group_by(country) %>% 
-  #filter(year == 2012) %>% 
-  #select(total_deaths) %>% 
-  #summarize(average_death_12 = mean(total_deaths))
-
-#death_17<- deaths_df %>% 
-  #group_by(country) %>% 
-  #filter(year == 2017) %>% 
-  #select(total_deaths) %>% 
-  #summarize(average_death_17 = mean(total_deaths))
-
-#try to have one cohesive table so it's easier to visualize 
-#combo <- death_07 %>% 
-  #mutate(death_12) %>% 
-  #mutate(death_17)
-
-#combo
-
-#Then graph it 
-#Just realized there's 200+ countries....the x-axis looks terrible 
-#Would be nice to color by continent but then that would mean going through each country
-#ggplot(death_07, aes(x= country, y=average_death_07)) +
-  #geom_point() +
-  #theme(axis.text.x = element_text(angle = 15)) #yah..no rotation made it worse
-
-```
-
-```{r}
-#SCRATCH WORK ABOVE
-
-#Not sure if we want to use the selected few countries we looked at earlier?
-#Or if we wanted to use this as a overview of the data
-
-#separation by high vs low population among continents 
-death_07_high <- deaths_df %>% 
-  filter(year == 2007, country %in% c('United States', 'Brazil', 'Ethiopia', 'Russia', 'China', 'Australia')) %>% 
-  group_by(country) %>% 
-  select(total_deaths) %>% 
-  summarize(average_death_07_high = mean(total_deaths))
-
-death_07_low<- deaths_df %>% 
-  filter(year == 2007, country %in% c('Canada', 'Argentina', 'Seychelles', 'Romania', 'Israel', 'Tonga')) %>% 
-  group_by(country) %>% 
-  select(total_deaths) %>% 
-  summarize(average_death_07_low = mean(total_deaths))
-
-#Combo table 2007
-combo_07<- death_07_high %>% 
-  mutate(death_07_low)
-combo_07
-
-death_12_high <- deaths_df %>% 
-  filter(year == 2012, country %in% c('United States', 'Brazil', 'Ethiopia', 'Russia', 'China', 'Australia')) %>% 
-  group_by(country) %>% 
-  select(total_deaths) %>% 
-  summarize(average_death_12_high = mean(total_deaths))
-
-death_12_low<- deaths_df %>% 
-  filter(year == 2012, country %in% c('Canada', 'Argentina', 'Seychelles', 'Romania', 'Israel', 'Tonga')) %>% 
-  group_by(country) %>% 
-  select(total_deaths) %>% 
-  summarize(average_death_12_low = mean(total_deaths))
-
-#Combo table 2012
-combo_12<- death_12_high %>% 
-  mutate(death_12_low)
-combo_12
-
-death_17_high <- deaths_df %>% 
-  filter(year == 2017, country %in% c('United States', 'Brazil', 'Ethiopia', 'Russia', 'China', 'Australia')) %>% 
-  group_by(country) %>% 
-  select(total_deaths) %>% 
-  summarize(average_death_17_high = mean(total_deaths))
-
-death_17_low<- deaths_df %>% 
-  filter(year == 2017, country %in% c('Canada', 'Argentina', 'Seychelles', 'Romania', 'Israel', 'Tonga')) %>% 
-  group_by(country) %>% 
-  select(total_deaths) %>% 
-  summarize(average_death_17_low = mean(total_deaths))
-
-#Combo table 2017
-combo_17<- death_17_high %>% 
-  mutate(death_17_low)
-combo_17
-
-#Graph selected countries 
-#Is there a way to color by low and high and combine the two graphs below?
-combo_07 %>% 
-  ggplot(aes(x = country, y = average_death_07_low)) + 
-  geom_point()
-
-combo_07 %>% 
-  ggplot(aes(x = country, y = average_death_07_high)) + 
-  geom_point()
-```
+## Which country has the greatest number of deaths? The least?
 
@@
 
-```{r}
-
-```
-
 ## Which year had the worst air pollution (average the three diff types)?
-
-```{r}
-
-````
 
@@
 
-```{r}
+##Focus in on some countries/ continents
 
-````
+##Pull in more data on populations for the selected countries
 
-# Notes from first meeting
+##Industrial and non industrial zones, what countries have coal plants? oil plants?
 
-## Hone in to only one year. Find data that is only population, continent data. Choose a couple countries/continents to work on
+##Years to focus on: 2007 to 2017,
 
-## Look at industrialized vs non-industrialized countries
+##Countries:
+#US - north america - high pop
+#Canada - north america - low pop
+
+#Brazil - S america - high pop
+#Argentina - S America - low pop
+
+#Ethiopia - Africa - high pop
+#Seychellas - Africa - low pop
+
+#Russia - Europe - high pop
+#Romania - Europe - low pop
+
+#China - Asia - high pop 
+#Israel - Asia - low pop
+
+#Australis - Oceania - high pop
+#Tonga - Oceania - low pop

### 2022-04-20 Project_draft.Rmd
filter to selected north american countries

--- 2022-04-18 Project_draft.Rmd
@@
 knitr::opts_chunk$set(echo = TRUE)
+```
+
+```{r}
 
@@
 
-#
+#filter to north american countries american and canada 
+#in the years 2007 - 2017
+```{r}
+america <- deaths_df %>% filter(country == "United States" & year > 1999)
+canada <- deaths_df %>% filter(country == "Canada" & year > 1999)
+
+```
 

### 2022-04-20 Draft2.Rmd
made tables with kable

--- 2022-04-20 Draft2.Rmd
@@
 
-```{r}
+```{r, include=FALSE}
 # Want the range 2007-2017 
@@
 ```
-
-```{r}
+# Separation by Population
+```{r, echo=FALSE}
 #SCRATCH WORK ABOVE
@@
 #Or if we wanted to use this as a overview of the data
+library(knitr)
+#install.packages('kableExtra')
+library(kableExtra)
 
@@
   summarize(average_death_07_high = mean(total_deaths))
+#death_07_high
 
@@
   summarize(average_death_07_low = mean(total_deaths))
+#death_07_low
 
-#Combo table 2007
-combo_07<- death_07_high %>% 
-  mutate(death_07_low)
-combo_07
+#Side by side table 
+#kable(list(death_07_high, death_07_low))
+kable(death_07_high) %>% 
+  kable_styling(full_width = FALSE, position = 'float_left')
+kable(death_07_low) %>% 
+  kable_styling(full_width = FALSE, position = 'left')
 
@@
   summarize(average_death_12_high = mean(total_deaths))
+#death_12_high
 
@@
   summarize(average_death_12_low = mean(total_deaths))
+#death_12_low
 
 #Combo table 2012
-combo_12<- death_12_high %>% 
-  mutate(death_12_low)
-combo_12
+#kable(list(death_12_high, death_12_low))
+kable(death_12_high) %>% 
+  kable_styling(full_width = FALSE, position = 'float_left')
+kable(death_12_low) %>% 
+  kable_styling(full_width = FALSE, position = 'left')
 
@@
   summarize(average_death_17_high = mean(total_deaths))
+#death_17_high
 
@@
   summarize(average_death_17_low = mean(total_deaths))
+#death_17_low
 
 #Combo table 2017
-combo_17<- death_17_high %>% 
-  mutate(death_17_low)
-combo_17
+#kable(list(death_17_high, death_17_low))
+kable(death_17_high) %>% 
+  kable_styling(full_width = FALSE, position = 'float_left')
+kable(death_17_low) %>% 
+  kable_styling(full_width = FALSE, position = 'left')
 
@@
 #Is there a way to color by low and high and combine the two graphs below?
-combo_07 %>% 
-  ggplot(aes(x = country, y = average_death_07_low)) + 
-  geom_point()
+#I removed combo_07 so this might not work...
 
-combo_07 %>% 
-  ggplot(aes(x = country, y = average_death_07_high)) + 
-  geom_point()
+#combo_07 %>% 
+  #ggplot(aes(x = country, y = average_death_07_low)) + 
+  #geom_point()
+
+#combo_07 %>% 
+  #ggplot(aes(x = country, y = average_death_07_high)) + 
+  #geom_point()
 ```

### 2022-04-23 population_data.Rmd
Added population data

##Population data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(dplyr)
library(ggplot2)

```

```{r}
pop_data_stats <- read.csv("pop_worldometer_data.csv")

glimpse(pop_data_stats)

world_pop <- read.csv("population_total_long.csv")

glimpse(world_pop)

```

### 2022-04-25 population_data.Rmd
issues i ran into when loading file
population data

@@
 
+high_pop_countries <- world_pop %>% 
+  filter(Year > 1999 & Country.Name %in% c('United States', 'Brazil', 'Ethiopia', 'Russia', 'China', 'Australia')) %>% 
+  group_by(Year)
+
+high_pop_countries
+
+low_pop_countries <- world_pop %>% 
+  filter(Year > 1999 & Country.Name %in% c('United States', 'Brazil', 'Ethiopia', 'Russia', 'China', 'Australia'))%>% 
+  group_by(Year)
+
+low_pop_countries
+
 ```

### 2022-04-27 outline.Rmd

---
title: "rough outline"
output: slidy_presentation
date: '2022-04-25'
---

# Introduction 
* Air pollution can be detrimental to both our health and the climate
  + Pollutants of major public health concern include carbon monoxide, ozone, nitrogen oxide, and sulfur oxide
  + Outdoor and indoor air pollution cause respiratory and other diseases. 
  + Air pollution kills about 7 million people worldwide every year

* We want to look at death rates of air pollution world wide

* Why?
  + Hopefully this information can showcase the importance of air pollution and that we should be more mindful about our planet

* Overview
  + We will see how different air pollution types affect the population 
  + See if countries have implemented better techniques to reduce air pollution

# Packages Required

```{r, include=FALSE}
#This will allow us to filter through our data 
library(tidyverse)
library(dplyr)
#This will help us plot figures to showcase our findings
library(ggplot2)
```

# Data Preparation 
The data we will be looking at:

1. **Deaths due to Air Pollution**
[https://www.kaggle.com/datasets/akshat0giri/death-due-to-air-pollution-19902017]

* Author: Akshat Giri
* This data consists the deaths due to air pollution from 1990-2017 around the world
* The dataset was recently updated 2 years ago
* The author wanted to gain more awareness about air pollution and forecast the future

2. **World Population(pending)**

First let's import the Deaths by Air Pollution data.
```{r}
deaths_df <- data.frame(read.csv("death-rates-from-air-pollution.csv"))
```

We are going to rename a few of the columns and view the first 6 rows of the data
```{r}
colnames(deaths_df) <- c("country", "acronym", "year", "total_deaths", "indoor_deaths", "outdoor_deaths", "ozone_deaths")

head(deaths_df)
```

Variables that interest us here include:
* country
* total_deaths
* indoor_deaths
* outdoor_deaths
* ozone_deaths

Now, we are going to import the World Population data 
```{r}
world_pop <- read.csv("population_total_long.csv")
```

Let's view the first 6 rows of the data
```{r}
head(world_pop)
```
# Exploratory Data Analysis

# Summary

### 2022-04-27 outline.Rmd
Moved some graphs and tables from the Draft2rmd into the outline

@@
 library(ggplot2)
+#This will help us organize and display our data as necessary 
+library(knitr)
+library(kableExtra)
+#This expands our plot uses 
+library(plotly)
 ```
@@
 Let's view the first 6 rows of the data
+This dataset covers more than we need, and so we can trim it to the years we need.
 ```{r}
-head(world_pop)
+world_pop2 <- world_pop %>% filter(Year >= 1990 & Year <= 2017)
+glimpse(world_pop2)
+
 ```
+After selecting our desired years, we've cut the population data set nearly in half.
+
+To get a general idea of the data, let's make some plots to see what's happening.
+
+```{r}
+p <- ggplot(deaths_df, aes(x = indoor_deaths, y = outdoor_deaths, text = paste0(country, ", ", year) )) + geom_point()
+ggplotly(p)
+```
+
+This is a mess, and so we chose two countries from each continent (a high-population and a low-population country) to graph.
+
+```{r}
+#Mean total deaths from 1990-2017 of high-population countries
+deaths_highpop_countries <- deaths_df %>% 
+  filter(country %in% c('United States', 'Brazil', 'Ethiopia', 'Russia', 'China', 'Australia')) %>% 
+  group_by(country) %>% 
+  select(total_deaths) %>% 
+  summarize(average_death_high = mean(total_deaths))
+
+#deaths_highpop_countries
+
+#Mean total deaths from 1990-2017 of high-population countries
+deaths_lowpop_countries<- deaths_df %>% 
+  filter(country %in% c('Canada', 'Argentina', 'Seychelles', 'Romania', 'Israel', 'Tonga')) %>% 
+  group_by(country) %>% 
+  select(total_deaths) %>% 
+  summarize(average_death_low = mean(total_deaths))
+
+#death_lowpop_countries
+kable(list(deaths_highpop_countries, deaths_lowpop_countries))
+kable(deaths_highpop_countries) %>% 
+  kable_styling(full_width = FALSE, position = 'float_left')
+kable(deaths_lowpop_countries) %>% 
+  kable_styling(full_width = FALSE, position = 'left')
+
+ggplot(deaths_highpop_countries)+
+  geom_col(mapping = aes(x=country, y=average_death_high))+
+             xlab("Country")+
+             ylab("Average deaths (per 100,000)")+
+             ggtitle("Average total deaths in high-population countries")+
+  coord_flip()
+
+ggplot(deaths_lowpop_countries)+
+  geom_col(mapping = aes(x=country, y=average_death_low))+
+             xlab("Country")+
+             ylab("Average deaths (per 100,000)")+
+             ggtitle("Average total deaths in low-population countries")+
+  coord_flip()
+
+```
+This shows us the deaths due to pollution, but what about the average population of those countries at that time?
+
+```{r}
+hp_countries_population <- world_pop2 %>% 
+  filter(Country.Name %in% c('United States', 'Brazil', 'Ethiopia', 'Russia', 'China', 'Australia')) %>% 
+  group_by(Country.Name) %>% 
+  select(Count) %>% 
+  summarize(average_population = mean(Count))
+
+#hp_countries_population
+
+lp_countries_population <- world_pop2 %>% 
+  filter(Country.Name %in% c('Canada', 'Argentina', 'Seychelles', 'Romania', 'Israel', 'Tonga')) %>% 
+  group_by(Country.Name) %>% 
+  select(Count) %>% 
+  summarize(average_population = mean(Count))
+
+#lp_countries_population
+
+kable(list(hp_countries_population, lp_countries_population))
+kable(hp_countries_population) %>% 
+  kable_styling(full_width = FALSE, position = 'float_left')
+kable(lp_countries_population) %>% 
+  kable_styling(full_width = FALSE, position = 'left')
+
+```
+
 # Exploratory Data Analysis

### 2022-04-27 outline.Rmd
Joined deaths_df and world_pop so we can make more graphs

@@
 
+#Join the data sets so we can overlay the two graph or do a stacked barchart?
+
+- `full_join(x, y, by = )`:return all rows and all columns from both x and y. Where there are not matching values, returns NA for the one missing.
+- full join deaths_df and world_pop2
+
+```{r}
+joined_df <- full_join(deaths_df, world_pop2, by = c("country" = "Country.Name"))
+glimpse(joined_df)
+
+```
+```{r}
+ggplot(joined_df, aes(fill=Count, y=country, x=year)) + 
+  geom_bar(position="stack", stat="identity")
+
+```
+
 # Exploratory Data Analysis

### 2022-04-29 population_data.Rmd
changes were made after second meeting
These is the population data with countries and years we want.

--- 2022-04-23 population_data.Rmd
@@
 glimpse(world_pop)
+head(world_pop)
 
 ```
+
+The above is 1960/2020 and our data is from 1990-2017
+This is a different file but won't open (resolved)
+```{r}
+#This is the data 
+pops_by_country <- read.csv("world_population_update.csv")
+head(pops_by_country)
+
+#The columns that we want 
+pops_by_country %>% filter(Country.Name %in% c('United States', 'Brazil', 'Ethiopia','Russian Federation','China', 'Australia','Canada','Argentina','Seychelles','Romania','Israel', 'Tonga')) %>% select(Country.Name, Country.Code, X1990, X1991, X1992, X1993, X1994, X1995, X1996, X1997, X1998, X1999, X2000, X2001, X2002, X2003, X2004, X2005, X2006, X2007, X2008, X2009, X2010, X2011, X2012, X2013, X2014, X2015, X2016, X2017)
+```

### 2022-04-29 population_data.Rmd
revised columns and deleted lines

--- 2022-04-25 population_data.Rmd
@@
 
+head(world_pop)
+
 ```
+
+The above is 1960/2020 and our data is from 1990-2017
+This is a different file but won't open (resolved)
+```{r}
+#This is the data 
+pops_by_country <- read.csv("world_population_update.csv")
+head(pops_by_country)
+
+#The columns that we want 
+selected_pop <- pops_by_country %>% filter(Country.Name %in% c('United States', 'Brazil', 'Ethiopia','Russian Federation','China', 'Australia','Canada','Argentina','Seychelles','Romania','Israel', 'Tonga')) %>% select(Country.Name, Country.Code, X1990, X1991, X1992, X1993, X1994, X1995, X1996, X1997, X1998, X1999, X2000, X2001, X2002, X2003, X2004, X2005, X2006, X2007, X2008, X2009, X2010, X2011, X2012, X2013, X2014, X2015, X2016, X2017)
+
+#colnames(selected_pop) <- c()
+
+edit <- selected_pop %>% 
+  select(Country.Name, Country.Code, X1990, X1991, X1992, X1993, X1994, X1995, X1996, X1997, X1998, X1999, X2000, X2001, X2002, X2003, X2004, X2005, X2006, X2007, X2008, X2009, X2010, X2011, X2012, X2013, X2014, X2015, X2016, X2017) %>%
+  rename(c('1990' = X1990, '1991'=X1991, '1992'=X1992, '1993'=X1993, '1994'=X1994, '1995'=X1995, '1996'=X1996, '1997'=X1997, '1998'=X1998, '1999'=X1999, '2000'=X2000, '2001'=X2001, '2002'=X2002, '2003'=X2003, '2004'=X2004, '2005'=X2005, '2006'=X2006, '2007'=X2007, '2008'=X2008, '2009'=X2009, '2010'=X2010, '2011'=X2011, '2012'=X2012, '2013'=X2013, '2014'=X2014, '2015'=X2015, '2016'=X2016, '2017'=X2017))
+
+edit
+```

### 2022-04-29 population_data.Rmd
Joined dataframe

@@
 ```
+```{r}
+country_deaths <- deaths_df %>% 
+    filter(country %in% c('Canada', 'Argentina', 'Seychelles', 'Romania', 
+                          'Israel', 'Tonga','United States', 'Brazil', 
+                          'Ethiopia', 'Russia', 'China', 'Australia'))
+
+#joined_country_pop <- full_join(country_deaths, edit, by = c("country" = "Country.Name"))
+#glimpse(joined_country_pop)
+#head(joined_country_pop, 15)
+
+joined_country_pop <- full_join(country_deaths, edit, by = c("country" = "Country.Name"))
+```

### 2022-04-29 outline.Rmd
A little clean up
Changed countries

@@
 deaths_highpop_countries <- deaths_df %>% 
-  filter(country %in% c('United States', 'Brazil', 'Ethiopia', 'Russia', 'China', 'Australia')) %>% 
+  filter(country %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
   group_by(country) %>% 
@@
 deaths_lowpop_countries<- deaths_df %>% 
-  filter(country %in% c('Canada', 'Argentina', 'Seychelles', 'Romania', 'Israel', 'Tonga')) %>% 
+  filter(country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand')) %>% 
   group_by(country) %>% 
@@
 ```{r}
-joined_df <- full_join(deaths_df, world_pop2, by = c("country" = "Country.Name"))
-glimpse(joined_df)
-
-```
-```{r}
-ggplot(joined_df, aes(fill=Count, y=country, x=year)) + 
-  geom_bar(position="stack", stat="identity")
 

### 2022-04-29 population_data.Rmd
A little clean up
New population data set

@@
 high_pop_countries <- world_pop %>% 
-  filter(Year > 1999 & Country.Name %in% c('United States', 'Brazil', 'Ethiopia', 'Russia', 'China', 'Australia')) %>% 
+  filter(Year > 1999 & Country.Name %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
   group_by(Year)
@@
 low_pop_countries <- world_pop %>% 
-  filter(Year > 1999 & Country.Name %in% c('United States', 'Brazil', 'Ethiopia', 'Russia', 'China', 'Australia'))%>% 
+  filter(Year > 1999 & Country.Name %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand'))%>% 
   group_by(Year)
@@
 joined_country_pop <- full_join(country_deaths, edit, by = c("country" = "Country.Name"))
+
+#ggplot(joined_df, aes(fill=Count, y=country, x=year)) + 
+#  geom_bar(position="stack", stat="identity")
+
+#pops_by_country$Country.Name[[Indicator.Name("Population, total" >= 130000000)]
+#rownames(table_name)[rownames(table_name) == "old label"] = "new label"
 ```

### 2022-04-30 outline.Rmd
transfering info on outline

@@
 world_pop <- read.csv("population_total_long.csv")
+glimpse(world_pop)
 ```
 
-Let's view the first 6 rows of the data
-This dataset covers more than we need, and so we can trim it to the years we need.
+Let's split the data into high and low population based on country.
+- Low population = high population * .10
 ```{r}
-world_pop2 <- world_pop %>% filter(Year >= 1990 & Year <= 2017)
-glimpse(world_pop2)
+#Let's compare the first decade 1996-2006 with second decade 2007-2017 
+high_pop_countries <- world_pop %>% 
+  filter(Year > 1996 & Country.Name %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
+  group_by(Year)
+
+high_pop_countries
+
+low_pop_countries <- world_pop %>% 
+  filter(Year > 1996 & Country.Name %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand'))%>% 
+  group_by(Year)
+
+low_pop_countries
 
@@
 ```{r}
-p <- ggplot(deaths_df, aes(x = indoor_deaths, y = outdoor_deaths, text = paste0(country, ", ", year) )) + geom_point()
-ggplotly(p)
+d <- ggplot(deaths_df, aes(x = indoor_deaths, y = outdoor_deaths, text = paste0(country, ", ", year) )) + geom_point()
+ggplotly(d)
+
+#p <- ggplot(world_pop, aes(x = Country.Name, y = Year,  text = paste0(Country.Name, ", ", Year) )) + geom_density()
+#ggplotly(p)
 ```
@@
 ```{r}
-hp_countries_population <- world_pop2 %>% 
+hp_countries_population <- world_pop %>% 
   filter(Country.Name %in% c('United States', 'Brazil', 'Ethiopia', 'Russia', 'China', 'Australia')) %>% 
@@
 
-lp_countries_population <- world_pop2 %>% 
+lp_countries_population <- world_pop %>% 
   filter(Country.Name %in% c('Canada', 'Argentina', 'Seychelles', 'Romania', 'Israel', 'Tonga')) %>% 

### 2022-04-30 outline.Rmd
updates on outline

@@
 hp_countries_population <- world_pop %>% 
-  filter(Country.Name %in% c('United States', 'Brazil', 'Ethiopia', 'Russia', 'China', 'Australia')) %>% 
+  filter(Country.Name %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
   group_by(Country.Name) %>% 
@@
 lp_countries_population <- world_pop %>% 
-  filter(Country.Name %in% c('Canada', 'Argentina', 'Seychelles', 'Romania', 'Israel', 'Tonga')) %>% 
+  filter(Country.Name %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand')) %>% 
   group_by(Country.Name) %>% 
@@
 
+ggplot(hp_countries_population)+
+  geom_col(mapping = aes(x=Country.Name, y=average_population))+
+             xlab("Country")+
+             ylab("Average Population")+
+             ggtitle("Average high-population countries")+
+  coord_flip()
+
+ggplot(lp_countries_population)+
+  geom_col(mapping = aes(x=Country.Name, y=average_population))+
+             xlab("Country")+
+             ylab("Average Population")+
+             ggtitle("Average low-population countries")+
+  coord_flip()
 ```

### 2022-04-30 outline.Rmd
updates on outline

@@
 library(plotly)
+#Adjusting numbers 
+format(4.470862e-01, scientific = F, digits = 3)
 ```

### 2022-04-30 outline.Rmd
Made changes to the slidy presentation

@@
 ---
-title: "rough outline"
+title: "Global deaths due to air ollution"
+author: "Elizabeth Bekele, Alison Cheek"
 output: slidy_presentation
-date: '2022-04-25'
+date: '2022-05-03'
 ---
@@
 * Air pollution can be detrimental to both our health and the climate
-  + Pollutants of major public health concern include carbon monoxide, ozone, nitrogen oxide, and sulfur oxide
-  + Outdoor and indoor air pollution cause respiratory and other diseases. 
+  + Outdoor and indoor air pollution cause chronic pain, respiratory diseases, shortened lifespan
   + Air pollution kills about 7 million people worldwide every year 
-
-* We want to look at death rates of air pollution world wide
-
-* Why?
   + Hopefully this information can showcase the importance of air pollution and that we should be more mindful about our planet
@@
   + We will see how different air pollution types affect the population 
-  + See if countries have implemented better techniques to reduce air pollution
+  + compare past and present population numbers
+  + determine which air pollutant type has the highest associated death rate
 
 # Packages Required
+
+# Data Details 
+*Air pollution deaths 
+  +Kaggle: https://www.kaggle.com/datasets/akshat0giri/death-due-to-air-pollution-19902017 
+  + Author: Akshat Giri
+  + Last update: 2 years ago
 

### 2022-04-30 outline.Rmd
re-arranged chunks in outline

@@
 ---
-
+```{r, include=FALSE}
+#This will allow us to filter through our data 
+library(tidyverse)
+library(dplyr)
+#This will help us plot figures to showcase our findings
+library(ggplot2)
+#This will help us organize and display our data as necessary 
+library(knitr)
+library(kableExtra)
+#This expands our plot uses 
+library(plotly)
+```
 # Introduction 
@@
 *Air pollution deaths 
-  +Kaggle: https://www.kaggle.com/datasets/akshat0giri/death-due-to-air-pollution-19902017 
+  +Kaggle: [https://www.kaggle.com/datasets/akshat0giri/death-due-to-air-pollution-19902017] 
   + Author: Akshat Giri
@@
 
-```{r, include=FALSE}
-#This will allow us to filter through our data 
-library(tidyverse)
-library(dplyr)
-#This will help us plot figures to showcase our findings
-library(ggplot2)
-#This will help us organize and display our data as necessary 
-library(knitr)
-library(kableExtra)
-#This expands our plot uses 
-library(plotly)
-#Adjusting numbers 
-format(4.470862e-01, scientific = F, digits = 3)
-```
-
-# Data Preparation 
-The data we will be looking at:
-
-1. **Deaths due to Air Pollution**
-[https://www.kaggle.com/datasets/akshat0giri/death-due-to-air-pollution-19902017]
-
-* Author: Akshat Giri
-* This data consists the deaths due to air pollution from 1990-2017 around the world
-* The dataset was recently updated 2 years ago
-* The author wanted to gain more awareness about air pollution and forecast the future
-
-2. **World Population(pending)**
-
-First let's import the Deaths by Air Pollution data.
+Import the Deaths by Air Pollution data
 ```{r}
@@
 world_pop <- read.csv("population_total_long.csv")
-glimpse(world_pop)
+head(world_pop)
+```
+To get a general idea of the data, let's make some plots to see what's happening.
+
+```{r}
+d <- ggplot(deaths_df, aes(x = indoor_deaths, y = outdoor_deaths, text = paste0(country, ", ", year) )) + geom_point()
+ggplotly(d)
 ```
 
-Let's split the data into high and low population based on country.
-- Low population = high population * .10
+This is a mess, and so we chose two countries from each continent (a high-population and a low-population country) to graph.
+
+# Exploratory Data Analysis
+
+Let's split the data into high and low population based on country
+
+  +Low population = high population * .10
+
 ```{r}
-#Let's compare the first decade 1996-2006 with second decade 2007-2017 
+#selecting high-population countries from the world population data frame 
 high_pop_countries <- world_pop %>% 
@@
 
+#selecting low-population countries from the world population data frame 
 low_pop_countries <- world_pop %>% 
@@
 ```
-After selecting our desired years, we've cut the population data set nearly in half.
-
-To get a general idea of the data, let's make some plots to see what's happening.
-
-```{r}
-d <- ggplot(deaths_df, aes(x = indoor_deaths, y = outdoor_deaths, text = paste0(country, ", ", year) )) + geom_point()
-ggplotly(d)
-
-#p <- ggplot(world_pop, aes(x = Country.Name, y = Year,  text = paste0(Country.Name, ", ", Year) )) + geom_density()
-#ggplotly(p)
-```
-
-This is a mess, and so we chose two countries from each continent (a high-population and a low-population country) to graph.
 
@@
 
-# Exploratory Data Analysis
-
 # Summary

### 2022-04-30 outline.Rmd
re-arranged the outline

@@
 ---
-title: "Global deaths due to air ollution"
+title: "Global Deaths Due to Air Pollution"
 author: "Elizabeth Bekele, Alison Cheek"
@@
 # Data Details 
-*Air pollution deaths 
-  +Kaggle: [https://www.kaggle.com/datasets/akshat0giri/death-due-to-air-pollution-19902017] 
+* Deaths from air pollution:
+  + Kaggle: [https://www.kaggle.com/datasets/akshat0giri/death-due-to-air-pollution-19902017] 
   + Author: Akshat Giri
@@
 
-Import the Deaths by Air Pollution data
+Import the deaths-due-to-air-pollution data
 ```{r}
@@
 
-We are going to rename a few of the columns and view the first 6 rows of the data
+We are going to rename a few of the columns and glimpse the data
 ```{r}
@@
 
-head(deaths_df)
+glimpse(deaths_df)
 ```
 
-Variables that interest us here include:
-* country
-* total_deaths
-* indoor_deaths
-* outdoor_deaths
-* ozone_deaths
+Variables that interest us here include: country, total_deaths, indoor_deaths, outdoor_deaths, ozone_deaths
 
-Now, we are going to import the World Population data 
+Now, we are going to import the world-popultaion data and glimpse
 ```{r}
 world_pop <- read.csv("population_total_long.csv")
-head(world_pop)
+glimpse(world_pop)
 ```
-To get a general idea of the data, let's make some plots to see what's happening.
+To get a general idea of 'deaths-dataframe' we made, let's make a plots to see what's happening. This is a plot of indoor x outdoor deaths around the world by country.
 
@@
 
-Let's split the data into high and low population based on country
+First, we split the data into high and low population based on country
 

### 2022-05-01 outline.Rmd
lots of updates

@@
 ---
-```{r, include=FALSE}
-#This will allow us to filter through our data 
-library(tidyverse)
-library(dplyr)
-#This will help us plot figures to showcase our findings
-library(ggplot2)
-#This will help us organize and display our data as necessary 
-library(knitr)
-library(kableExtra)
-#This expands our plot uses 
-library(plotly)
-```
 # Introduction 
@@
 # Packages Required 
+```{r, message=FALSE}
+#This will allow us to filter through our data 
+library(tidyverse)
+library(dplyr)
+#This will help us plot figures to showcase our findings
+library(ggplot2)
+#This will help us organize and display our data as necessary 
+library(knitr)
+library(kableExtra)
+#This expands our plot uses 
+library(plotly)
+```
 
@@
 
-Now, we are going to import the world-popultaion data and glimpse
+# Data Continued
+
+Now, let's take a look at the population data.
 ```{r}
@@
 ```
+
 To get a general idea of 'deaths-dataframe' we made, let's make a plots to see what's happening. This is a plot of indoor x outdoor deaths around the world by country.
@@
 ```{r}
-d <- ggplot(deaths_df, aes(x = indoor_deaths, y = outdoor_deaths, text = paste0(country, ", ", year) )) + geom_point()
+d <- ggplot(deaths_df, aes(x = indoor_deaths, y = outdoor_deaths, text = paste0(country, ", ", year) )) + geom_point() +
+  ggtitle("Outdoor Deaths vs Indoor Deaths")
 ggplotly(d)
@@
 
-# Exploratory Data Analysis
+# Combine Data Sets
+
+First let's look at a table of the high and low populated countries using the world population data set.
+```{r}
+#selecting high-population countries from the world population data frame 
+high_pop_countries <- world_pop %>% 
+  filter(Year > 1996 & Country.Name %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
+  group_by(Year)
+
+head(high_pop_countries)
+
+#selecting low-population countries from the world population data frame 
+low_pop_countries <- world_pop %>% 
+  filter(Year > 1996 & Country.Name %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand'))%>% 
+  group_by(Year)
+
+head(low_pop_countries)
+```
+
+Next, we are going to see the death count for high and low populated countries using the deaths dataframe.
+```{r}
+#selecting high-population deaths from death dataframe 
+high_pop_death <- deaths_df %>% 
+  filter(year > 1996 & country %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
+  group_by(year)
+
+head(high_pop_death)
+
+#selecting low-population deaths from death dataframe 
+low_pop_death <- deaths_df %>% 
+  filter(year > 1996 & country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand')) %>% 
+  group_by(year)
+
+head(low_pop_death)
+
+```
+
+Lastly, we will join the population and and deaths with its respected country.
+```{r}
+#Combined High Population with Pollution Death
+joined_high <- right_join(high_pop_death, high_pop_countries, by= c('country' = 'Country.Name', 'year' = 'Year'))
+head(joined_high)
+
+#Combined Low Population with Pollution Death
+joined_low <-right_join(low_pop_death, low_pop_countries, by= c('country' = 'Country.Name', 'year' = 'Year'))
+head(joined_low)
+
+```
+
+# Which country has the highest death count?
+
+Let's make a table depicting the high and low populated countries and their respected death count due to pollution.
+
+```{r}
+#Mean total deaths of high-population countries
+deaths_highpop_countries <- deaths_df %>% 
+  filter(country %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
+  group_by(country) %>% 
+  select(total_deaths) %>% 
+  summarize(average_death_high = mean(total_deaths))
+
+#Mean total deaths of high-population countries
+deaths_lowpop_countries<- deaths_df %>% 
+  filter(country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand')) %>% 
+  group_by(country) %>% 
+  select(total_deaths) %>% 
+  summarize(average_death_low = mean(total_deaths))
+
+#Side-by-side Table to view deaths 
+kable(list(deaths_highpop_countries, deaths_lowpop_countries))
+
+```
+
+Here's a graph to clearly visualize the previous table
+```{r}
+#Plot High Population Deaths (average)
+ggplot(deaths_highpop_countries)+
+  geom_col(mapping = aes(x=country, y=average_death_high))+
+             xlab("Country")+
+             ylab("Average deaths (per 100,000)")+
+             ggtitle("Average total deaths in high-population countries")+
+  coord_flip()
+
+#Plot of Low Population Deaths (average)
+ggplot(deaths_lowpop_countries)+
+  geom_col(mapping = aes(x=country, y=average_death_low))+
+             xlab("Country")+
+             ylab("Average deaths (per 100,000)")+
+             ggtitle("Average total deaths in low-population countries")+
+  coord_flip()
+```
+
+# Which type of pollution has the greatest number of deaths?
+# Pollution Over Time 
+## Which year had the worst pollution?
+## Which year had the worst indoor? Outdoor particulate? Outdoor ozone? 
+# Which is worse - outdoor or indoor pollution?
 
@@
 
-  +Low population = high population * .10
+Low population = high population * .10
 
@@
 ```{r}
-#Mean total deaths from 1990-2017 of high-population countries
+#Mean total deaths from 1996-2017 of high-population countries
 deaths_highpop_countries <- deaths_df %>% 
@@
   summarize(average_death_high = mean(total_deaths))
-
-#deaths_highpop_countries
 
@@
 kable(list(deaths_highpop_countries, deaths_lowpop_countries))
-kable(deaths_highpop_countries) %>% 
-  kable_styling(full_width = FALSE, position = 'float_left')
-kable(deaths_lowpop_countries) %>% 
-  kable_styling(full_width = FALSE, position = 'left')
 
@@
 ```
+
 This shows us the deaths due to pollution, but what about the average population of those countries at that time?
@@
 hp_countries_population <- world_pop %>% 
-  filter(Country.Name %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
+  filter(Country.Name %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia'), Year > 1996) %>% 
   group_by(Country.Name) %>% 
@@
 lp_countries_population <- world_pop %>% 
-  filter(Country.Name %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand')) %>% 
+  filter(Country.Name %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand'), Year > 1996) %>% 
   group_by(Country.Name) %>% 
@@
 
-#lp_countries_population
+#Population Average Table
+kable(list(hp_countries_population, lp_countries_population))
 
-kable(list(hp_countries_population, lp_countries_population))
-kable(hp_countries_population) %>% 
-  kable_styling(full_width = FALSE, position = 'float_left')
-kable(lp_countries_population) %>% 
-  kable_styling(full_width = FALSE, position = 'left')
-
+#Graph of Population Average
 ggplot(hp_countries_population)+

### 2022-05-01 outline.Rmd
up to the first question. need to add info on population data source

@@
 # Packages Required 
-```{r, message=FALSE}
+```{r, message=FALSE, warning=FALSE}
 #This will allow us to filter through our data 
@@
 
-# Data Details 
-* Deaths from air pollution:
+# Pollution Data
+* Deaths from pollution:
   + Kaggle: [https://www.kaggle.com/datasets/akshat0giri/death-due-to-air-pollution-19902017] 
@@
 
-Variables that interest us here include: country, total_deaths, indoor_deaths, outdoor_deaths, ozone_deaths
+# Data Variables
+Variables that interest us here include:
 
-# Data Continued
+* __country__ 
+* __total_deaths__: per 100,000
+* __indoor_deaths__: Indoor air pollution is considered pollution that occurs in the household. Cooking with solid fuels:
+  + Wood
+  + Crop waste, dung
+  + Charcoal, coal
+* __outdoor_deaths__: Outdoor air or ambient air are emissions caused by combustion processes from motor vehicles, solid fuel burning and industry 
+  + Ozone (O3)
+  + Particulate matter (PM10 and PM2.5)
+  + Nitrogen dioxide (NO2)
+  + Carbon monoxide (CO)
+  + Sulfur dioxide (SO2)
+* __ozone_deaths__: Ozone is a gas that occurs both in Earth's upper atmosphere and at ground level.
+  + Pollutants emitted by cars
+  + Power plants, industrial boilers, refineries, chemical plants
+
+# World Population Data 
+* World Population:
+  + Kaggle: [] 
+  + Author: 
+  + Last update:
+
+***
 
@@
 
+***
+
 Next, we are going to see the death count for high and low populated countries using the deaths dataframe.
@@
 ```
+
+***
 

### 2022-05-01 outline.Rmd
made the plots interactive for average death count

@@
 
-# Which country has the highest death count?
+# Death Count
+> Which country has the highest death count?
 
@@
 
+***
+
 Here's a graph to clearly visualize the previous table
@@
 #Plot High Population Deaths (average)
-ggplot(deaths_highpop_countries)+
+h<- ggplot(deaths_highpop_countries)+
   geom_col(mapping = aes(x=country, y=average_death_high))+
@@
 
+ggplotly(h)
+
 #Plot of Low Population Deaths (average)
-ggplot(deaths_lowpop_countries)+
+l <- ggplot(deaths_lowpop_countries)+
   geom_col(mapping = aes(x=country, y=average_death_low))+
@@
   coord_flip()
+
+ggplotly(l)
 ```
 
+***
+
 # Which type of pollution has the greatest number of deaths?
+
 # Pollution Over Time 

### 2022-05-01 outline.Rmd
question 2 complete

@@
 
-```{r}
+```{r, echo=FALSE, warning=FALSE}
 #Mean total deaths of high-population countries
@@
 Here's a graph to clearly visualize the previous table
-```{r}
+```{r, echo=FALSE}
 #Plot High Population Deaths (average)
@@
 # Which type of pollution has the greatest number of deaths?
+
+```{r}
+#High Population Pollutant Averages
+high_poll <- deaths_df %>% 
+  group_by(country) %>% 
+  filter(country %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
+  select(country, indoor_deaths, outdoor_deaths, ozone_deaths) %>% 
+  summarize(avg_indoor = mean(indoor_deaths), avg_outdoor = mean(outdoor_deaths), avg_ozone = mean(ozone_deaths))
+
+high_poll
+
+#Plot High Population
+#Indoor Air Pollution
+h_indoor<- ggplot(high_poll, aes(x=country, y = avg_indoor)) + 
+  geom_point() +
+  ggtitle("Average Death due to Indoor Air Pollution for High Population")
+ggplotly(h_indoor)
+
+#Outdoor Air Pollution
+h_outdoor <- ggplot(high_poll, aes(x=country, y = avg_outdoor, color = avg_ozone)) +
+  geom_point() +
+  labs(title = "Average Death due to Outdoor Air Pollution for High Population") +
+  xlab("Country") + 
+  ylab("Average Outdoor Air Pollution Deaths")
+ggplotly(h_outdoor)
+
+#Low Population Pollutant Averages 
+low_poll <- deaths_df %>% 
+  group_by(country) %>% 
+  filter(country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand')) %>% 
+  select(country, indoor_deaths, outdoor_deaths, ozone_deaths) %>% 
+  summarize(avg_indoor = mean(indoor_deaths), avg_outdoor = mean(outdoor_deaths), avg_ozone = mean(ozone_deaths))
+low_poll
+
+#Plot Low Population
+#Indoor Air Pollution
+l_indoor<- ggplot(low_poll, aes(x=country, y = avg_indoor)) + 
+  geom_point() +
+  ggtitle("Average Death due to Indoor Air Pollution for Low Population")
+ggplotly(l_indoor)
+
+#Outdoor Air Pollution
+l_outdoor <- ggplot(low_poll, aes(x=country, y = avg_outdoor, color = avg_ozone)) +
+  geom_point() +
+  labs(title = "Average Death due to Outdoor Air Pollution for Low Population", subtitle = "Colored by Average Ozone Air Pollution Deaths")
+ggplotly(l_outdoor)
+```
+
+***
 
@@
 
-#Join the data sets so we can overlay the two graph or do a stacked barchart?
-
-- `full_join(x, y, by = )`:return all rows and all columns from both x and y. Where there are not matching values, returns NA for the one missing.
-- full join deaths_df and world_pop2
-
-```{r}
-
-```
-
 # Summary

### 2022-05-01 outline.Rmd
disabled scientific notation

@@
 library(plotly)
+#Scientific Notation Disabled 
+options(scipen=999)
 ```
@@
 
-# Which type of pollution has the greatest number of deaths?
+# Pollution Types
+> Which type of pollution has the greatest number of deaths?
 
@@
 # Pollution Over Time 
-## Which year had the worst pollution?
+> Which year had the worst pollution?
+
+```{r}
+
+```
+
+***
+
 ## Which year had the worst indoor? Outdoor particulate? Outdoor ozone? 
@@
 deaths_lowpop_countries<- deaths_df %>% 
-  filter(country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand')) %>% 
+  filter(year> 1995 & country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand')) %>% 
   group_by(country) %>% 
@@
   coord_flip()
-
 ```

### 2022-05-01 outline.Rmd
made tables for the 2 decades need to make a graph for it

@@
 # Pollution Over Time 
-> Which year had the worst pollution?
+> Let's look at the previous two decades and compare the death count
+> Has there been a change?
 
+This is the first decade 1996-2006
+```{r, echo=FALSE, message=FALSE}
+# Look at the trend of deaths within the last decade from low and high populated countries
+death_96_high<- deaths_df %>% 
+  group_by(country) %>% 
+  filter(year == 1996, country %in% c('United States', 'Brazil','Nigeria', 'Germany', 'Pakistan', 'Australia')) %>%
+  select(total_deaths) %>% 
+  summarize(High_Deaths_96 = mean(total_deaths))
+
+death_96_low<- deaths_df %>% 
+  group_by(country) %>% 
+  filter(year == 1996, country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'Tonga')) %>%
+  select(total_deaths) %>% 
+  summarize(Low_Deaths_96 = mean(total_deaths))
+
+death_01_high<- deaths_df %>% 
+  group_by(country) %>% 
+  filter(year == 2001, country %in% c('United States', 'Brazil','Nigeria', 'Germany', 'Pakistan', 'Australia')) %>%
+  select(total_deaths) %>% 
+  summarize(High_Deaths_01 = mean(total_deaths))
+
+death_01_low<- deaths_df %>% 
+  group_by(country) %>% 
+  filter(year == 2001, country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'Tonga')) %>%
+  select(total_deaths) %>% 
+  summarize(Low_Deaths_01 = mean(total_deaths))
+
+death_06_high<- deaths_df %>% 
+  group_by(country) %>% 
+  filter(year == 2006, country %in% c('United States', 'Brazil','Nigeria', 'Germany', 'Pakistan', 'Australia')) %>%
+  select(total_deaths) %>% 
+  summarize(High_Deaths_06 = mean(total_deaths))
+
+death_06_low<- deaths_df %>% 
+  group_by(country) %>% 
+  filter(year == 2006, country %in% c('United States', 'Brazil','Nigeria', 'Germany', 'Pakistan', 'Australia')) %>%
+  select(total_deaths) %>% 
+  summarize(Low_Deaths_06 = mean(total_deaths))
+
+#cohesive table so it's easier to visualize 
+#Low Population Table 2007-2017
+all_low_first <- death_96_low %>% 
+  mutate(death_01_low) %>% 
+  mutate(death_06_low)
+
+#High Population Table 2007-2017
+all_high_first <- death_96_high %>% 
+  mutate(death_01_high) %>% 
+  mutate(death_06_high)
+
+kable(list(all_high_first))
+kable(list(all_low_first))
+```
+***
+
+This is the second decade 2007-2017
+```{r, echo=FALSE, message=FALSE}
+# Look at the trend of deaths within the last decade from low and high populated countries
+death_07_high<- deaths_df %>% 
+  group_by(country) %>% 
+  filter(year == 2007, country %in% c('United States', 'Brazil','Nigeria', 'Germany', 'Pakistan', 'Australia')) %>%
+  select(total_deaths) %>% 
+  summarize(High_Deaths_07 = mean(total_deaths))
+
+death_07_low<- deaths_df %>% 
+  group_by(country) %>% 
+  filter(year == 2007, country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'Tonga')) %>%
+  select(total_deaths) %>% 
+  summarize(Low_Deaths_07 = mean(total_deaths))
+
+death_12_high<- deaths_df %>% 
+  group_by(country) %>% 
+  filter(year == 2012, country %in% c('United States', 'Brazil','Nigeria', 'Germany', 'Pakistan', 'Australia')) %>%
+  select(total_deaths) %>% 
+  summarize(High_Deaths_12 = mean(total_deaths))
+
+death_12_low<- deaths_df %>% 
+  group_by(country) %>% 
+  filter(year == 2012, country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'Tonga')) %>%
+  select(total_deaths) %>% 
+  summarize(Low_Deaths_12 = mean(total_deaths))
+
+death_17_high<- deaths_df %>% 
+  group_by(country) %>% 
+  filter(year == 2017, country %in% c('United States', 'Brazil','Nigeria', 'Germany', 'Pakistan', 'Australia')) %>%
+  select(total_deaths) %>% 
+  summarize(High_Deaths_17 = mean(total_deaths))
+
+death_17_low<- deaths_df %>% 
+  group_by(country) %>% 
+  filter(year == 2017, country %in% c('United States', 'Brazil','Nigeria', 'Germany', 'Pakistan', 'Australia')) %>%
+  select(total_deaths) %>% 
+  summarize(Low_Deaths_17 = mean(total_deaths))
+
+#cohesive table so it's easier to visualize 
+#Low Population Table 2007-2017
+all_low_second <- death_07_low %>% 
+  mutate(death_12_low) %>% 
+  mutate(death_17_low)
+
+#High Population Table 2007-2017
+all_high_second <- death_07_high %>% 
+  mutate(death_12_high) %>% 
+  mutate(death_17_high)
+
+kable(list(all_high_second))
+kable(list(all_low_second))
+```
+
+***
+
+Let's graph the previous tables!
 ```{r}
@@
 
-#hp_countries_population
-
 lp_countries_population <- world_pop %>% 
@@
   coord_flip()
+
+#Percentage of those affected 
+affected_high<- joined_high %>% 
+  mutate(percent_high = total_deaths/Count)
+
+percent_high<- affected_high %>% ggplot(aes(x = year, y = percent_high, color = country))+
+  geom_point() +
+  labs(title = "Percentage of High Populated Countries Affected by Air Pollution") +
+  xlab("Year")+
+  ylab("Percent (total_deaths/Count)")
+ggplotly(percent_high)
+
+affected_low <- joined_low %>% 
+  mutate(percent_low = total_deaths/Count)
+
+percent_low <- affected_low %>% ggplot(aes(x = year, y = percent_low, color = country))+
+  geom_point() +
+  labs(title = "Percentage of Low Populated Countries Affected by Air Pollution") +
+  xlab("Year")+
+  ylab("Percent (total_deaths/Count)")
+ggplotly(percent_low)
 ```

### 2022-05-01 outline.Rmd
I was able to graph 2007 and 2012 for the high and low deaths but couldn't add another year w/o an error

@@
 ```{r}
+#High Population Deaths 2007 and 20012
+p_high <- all_high_second %>%
+  group_by(country) %>% 
+  ggplot(aes(x= High_Deaths_07, y= High_Deaths_12, color= country)) +
+  geom_point()
+
+interact_high<- p_high + labs(title="High Populated Countries", x="2007 Deaths", y="2012 Deaths")
+
+ggplotly(interact_high)
+
+#Low Population Deaths 2007 and 2012
+p_low <- all_low_second %>%
+  group_by(country) %>% 
+  ggplot(aes(x= Low_Deaths_07, y= Low_Deaths_12, color= country)) +
+  geom_point()
+
+interact_low<- p_low + labs(title="Low Populated Countries", x="2007 Deaths", y="2012 Deaths")
+
+ggplotly(interact_low)
 

### 2022-05-01 outline.Rmd
Figured out how to do all three years!

--- 2022-05-01 outline.Rmd
@@
 Import the deaths-due-to-air-pollution data
-```{r}
+```{r, echo=FALSE}
 deaths_df <- data.frame(read.csv("death-rates-from-air-pollution.csv"))
@@
 We are going to rename a few of the columns and glimpse the data
-```{r}
+```{r, echo=FALSE}
 colnames(deaths_df) <- c("country", "acronym", "year", "total_deaths", "indoor_deaths", "outdoor_deaths", "ozone_deaths")
@@
 Now, let's take a look at the population data.
-```{r}
+```{r, echo=FALSE}
 world_pop <- read.csv("population_total_long.csv")
@@
 
-```{r}
+```{r, echo=FALSE}
 d <- ggplot(deaths_df, aes(x = indoor_deaths, y = outdoor_deaths, text = paste0(country, ", ", year) )) + geom_point() +
@@
 First let's look at a table of the high and low populated countries using the world population data set.
-```{r}
+```{r, echo=FALSE}
 #selecting high-population countries from the world population data frame 
@@
 Next, we are going to see the death count for high and low populated countries using the deaths dataframe.
-```{r}
+```{r, echo=FALSE}
 #selecting high-population deaths from death dataframe 
@@
 Lastly, we will join the population and and deaths with its respected country.
-```{r}
+```{r, message=FALSE, echo=FALSE}
 #Combined High Population with Pollution Death
@@
 
-```{r, echo=FALSE, warning=FALSE}
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
 #Mean total deaths of high-population countries
@@
 Here's a graph to clearly visualize the previous table
-```{r, echo=FALSE}
+```{r, echo=FALSE, message=FALSE}
 #Plot High Population Deaths (average)
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE}
 #High Population Pollutant Averages
@@
 Let's graph the previous tables!
-```{r}
+```{r, echo=FALSE}
+#High Population Deaths 2007 and 20012
+p_high <- all_high_second %>%
+  group_by(country) %>% 
+  ggplot(aes(x= High_Deaths_07, y= High_Deaths_12, color= High_Deaths_17)) +
+  geom_point() +
+  facet_wrap(~country)
+
+interact_high<- p_high + labs(title="High Populated Countries", x="2007 Deaths", y="2012 Deaths")
+
+ggplotly(interact_high)
+
+#Low Population Deaths 2007 and 2012
+p_low <- all_low_second %>%
+  group_by(country) %>% 
+  ggplot(aes(x= Low_Deaths_07, y= Low_Deaths_12, color= Low_Deaths_17)) +
+  geom_point() +
+  facet_wrap(~country)
+
+interact_low<- p_low + labs(title="Low Populated Countries", x="2007 Deaths", y="2012 Deaths")
+
+ggplotly(interact_low)
 
@@
 
-```{r}
+```{r, echo=FALSE}
 #selecting high-population countries from the world population data frame 

### 2022-05-01 outline.Rmd
so i didnt change the country name for one year and it printed duplicates. its fixed now

@@
 Import the deaths-due-to-air-pollution data
-```{r, echo=FALSE}
+```{r}
 deaths_df <- data.frame(read.csv("death-rates-from-air-pollution.csv"))
@@
 We are going to rename a few of the columns and glimpse the data
-```{r, echo=FALSE}
+```{r}
 colnames(deaths_df) <- c("country", "acronym", "year", "total_deaths", "indoor_deaths", "outdoor_deaths", "ozone_deaths")
@@
 Now, let's take a look at the population data.
-```{r, echo=FALSE}
+```{r}
 world_pop <- read.csv("population_total_long.csv")

### 2022-05-01 outline.Rmd
Pollution types are done

@@
   group_by(country) %>% 
-  filter(year == 2017, country %in% c('United States', 'Brazil','Nigeria', 'Germany', 'Pakistan', 'Australia')) %>%
+  filter(year == 2017, country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'Tonga')) %>%
   select(total_deaths) %>% 
@@
 Let's graph the previous tables!
+
+The first decade.
 ```{r, echo=FALSE}
-#High Population Deaths 2007 and 20012
-p_high <- all_high_second %>%
+#High Population Deaths 1996-2006 
+p_high_first <- all_high_first %>%
+  group_by(country) %>% 
+  ggplot(aes(x= High_Deaths_96, y= High_Deaths_01, color= High_Deaths_06)) +
+  geom_point() +
+  facet_wrap(~country)
+
+interact_high_first<- p_high_first + labs(title="High Populated Countries", x="1996 Deaths", y="2001 Deaths")
+
+ggplotly(interact_high_first)
+
+#Low Population Deaths 1996-2006
+p_low_first <- all_low_first %>%
+  group_by(country) %>% 
+  ggplot(aes(x= Low_Deaths_96, y= Low_Deaths_01, color= Low_Deaths_06)) +
+  geom_point() +
+  facet_wrap(~country)
+
+interact_low_first<- p_low_first + labs(title="Low Populated Countries", x="1996 Deaths", y="2001 Deaths")
+
+ggplotly(interact_low_first)
+```
+
+***
+
+This shows the second decade.
+```{r, echo=FALSE}
+#High Population Deaths 2007-2017
+p_high_second <- all_high_second %>%
   group_by(country) %>% 
@@
 
-interact_high<- p_high + labs(title="High Populated Countries", x="2007 Deaths", y="2012 Deaths")
+interact_high_second<- p_high_second + labs(title="High Populated Countries", x="2007 Deaths", y="2012 Deaths")
 
-ggplotly(interact_high)
+ggplotly(interact_high_second)
 
-#Low Population Deaths 2007 and 2012
-p_low <- all_low_second %>%
+#Low Population Deaths 2007-2017
+p_low_second <- all_low_second %>%
   group_by(country) %>% 
@@
 
-interact_low<- p_low + labs(title="Low Populated Countries", x="2007 Deaths", y="2012 Deaths")
+interact_low_second<- p_low_second + labs(title="Low Populated Countries", x="2007 Deaths", y="2012 Deaths")
 
-ggplotly(interact_low)
+ggplotly(interact_low_second)
 
@@
 
-## Which year had the worst indoor? Outdoor particulate? Outdoor ozone? 
-# Which is worse - outdoor or indoor pollution?
+> Which year had the worst indoor? Outdoor particulate? Outdoor ozone?
 
+Indoor Deaths
+```{r, echo=FALSE}
+#Indoor 1996-2017
+indoor_high<- deaths_df %>% 
+    filter(year> 1995 & country %in% c('United States', 'Brazil','Nigeria', 
+                                       'Germany', 'Pakistan', 'Australia')) %>% 
+  group_by(country, year, indoor_deaths) %>% 
+  ggplot()+
+  geom_point(aes(year, indoor_deaths, color = country))+
+  ggtitle("High Populated Indoor Death 1996-2017")+
+  xlab("Year")+
+  ylab("Deaths per 100,000")
+ggplotly(indoor_high)
+
+indoor_low<- deaths_df %>% 
+    filter(year> 1995 & country %in% c('Canada', 'Chile', 'Malawi', 
+                                       'Serbia', 'Sri Lanka', 'Tonga')) %>% 
+  group_by(country, year, indoor_deaths) %>% 
+  ggplot()+
+  geom_point(aes(year, indoor_deaths, color = country))+
+  ggtitle("Low Populated Indoor Death 1996-2017")+
+  xlab("Year")+
+  ylab("Deaths per 100,000")
+ggplotly(indoor_low)
+
+#Pakistan and Malawi were both mainly affected in 1996
+indoor_top<- deaths_df %>% 
+    filter(year> 1995 & country %in% c('Pakistan', 'Malawi')) %>% 
+  group_by(country, year, indoor_deaths) %>% 
+  ggplot()+
+  geom_point(aes(year, indoor_deaths, color = country))+
+  ggtitle("Greatest Indoor Death 1996-2017")+
+  xlab("Year")+
+  ylab("Deaths per 100,000")
+ggplotly(indoor_top)
+```
+
+***
+
+Outdoor Deaths
+```{r, echo= FALSE}
+#Outdoor Particulate 1996-2017
+outdoor_high<- deaths_df %>% 
+    filter(year> 1995 & country %in% c('United States', 'Brazil','Nigeria', 
+                                       'Germany', 'Pakistan', 'Australia')) %>% 
+  group_by(country, year, outdoor_deaths) %>% 
+  ggplot()+
+  geom_point(aes(year, outdoor_deaths, color = country))+
+  ggtitle("High Populated Outdoor Death 1996-2017")+
+  xlab("Year")+
+  ylab("Deaths per 100,000")
+ggplotly(outdoor_high)
+
+outdoor_low<- deaths_df %>% 
+    filter(year> 1995 & country %in% c('Canada', 'Chile', 'Malawi', 
+                                       'Serbia', 'Sri Lanka', 'Tonga')) %>% 
+  group_by(country, year, outdoor_deaths) %>% 
+  ggplot()+
+  geom_point(aes(year, outdoor_deaths, color = country))+
+  ggtitle("Low Populated Outdoor Death 1996-2017")+
+  xlab("Year")+
+  ylab("Deaths per 100,000")
+ggplotly(outdoor_low)
+
+#Serbia and Pakistan were the top countries affected. 
+#2011 Pakistan had the greatest amount of Outdoor Deaths
+outdoor_top<- deaths_df %>% 
+    filter(year> 1995 & country %in% c('Pakistan', 'Serbia')) %>% 
+  group_by(country, year, outdoor_deaths) %>% 
+  ggplot()+
+  geom_point(aes(year, outdoor_deaths, color = country))+
+  ggtitle("Greatest Outdoor Death 1996-2017")+
+  xlab("Year")+
+  ylab("Deaths per 100,000")
+ggplotly(outdoor_top)
+```
+
+***
+Ozone Deaths 
+```{r, echo=FALSE}
+#Outdoor ozone 1996-2017
+ozone_high<- deaths_df %>% 
+    filter(year> 1995 & country %in% c('United States', 'Brazil','Nigeria', 
+                                       'Germany', 'Pakistan', 'Australia')) %>% 
+  group_by(country, year, ozone_deaths) %>% 
+  ggplot()+
+  geom_point(aes(year, ozone_deaths, color = country))+
+  ggtitle("High Populated Ozone Death 1996-2017")+
+  xlab("Year")+
+  ylab("Deaths per 100,000")
+ggplotly(ozone_high)
+
+ozone_low<- deaths_df %>% 
+    filter(year> 1995 & country %in% c('Canada', 'Chile', 'Malawi', 
+                                       'Serbia', 'Sri Lanka', 'Tonga')) %>% 
+  group_by(country, year, ozone_deaths) %>% 
+  ggplot()+
+  geom_point(aes(year, ozone_deaths, color = country))+
+  ggtitle("Low Populated Ozone Death 1996-2017")+
+  xlab("Year")+
+  ylab("Deaths per 100,000")
+ggplotly(ozone_low)
+
+#Malawi and Pakistan were the top countries affected 
+#1997 had the greatest amount of ozone death in Pakistan
+ozone_top<- deaths_df %>% 
+    filter(year> 1995 & country %in% c('Pakistan', 'Malawi')) %>% 
+  group_by(country, year, ozone_deaths) %>% 
+  ggplot()+
+  geom_point(aes(year, ozone_deaths, color = country))+
+  ggtitle("Greatest Ozone Death 1996-2017")+
+  xlab("Year")+
+  ylab("Deaths per 100,000")
+ggplotly(ozone_top)
+```
+
+# Which is worse? 
+> outdoor or indoor pollution?
+```{r}
+
+```
+
+# Reminder: Include this in the beginning 
 First, we split the data into high and low population based on country
@@
 ```
+
+# Include Earlier to See who's been affected
 

### 2022-05-01 outline.Rmd
All questions have been answered just need to do some reorganization, summary, link sources

@@
   geom_point() +
-  labs(title = "Average Death due to Outdoor Air Pollution for Low Population", subtitle = "Colored by Average Ozone Air Pollution Deaths")
+  labs(title = "Average Death due to Outdoor Air Pollution for Low Population")
 ggplotly(l_outdoor)
@@
 > outdoor or indoor pollution?
-```{r}
 
+Let's reintroduce a graph we looked at earlier. Instead this time we will combine the pollutant types together.
+```{r, echo=FALSE}
+#Plot High Population
+h_pollutant<- ggplot(high_poll, aes(x=avg_indoor, y = avg_outdoor, color= avg_ozone)) + 
+  geom_point() +
+  ggtitle("High Population Pollutant Type Death Counts")
+ggplotly(h_pollutant)
+
+#Plot Low Population
+l_pollutant<- ggplot(low_poll, aes(x=avg_indoor, y = avg_outdoor, color= avg_ozone)) + 
+  geom_point() +
+  ggtitle("Low Population Pollutant Type Death Counts")
+ggplotly(l_pollutant)
 ```
+
+We cannot conclude which is worse.
+
+* Low Populated Countries: 
+  + it's half-and-half
+
+* High Populated Countries: 
+  + Outdoor pollution seems to be more detrimental with the exception of two countries in this sample set.
 

### 2022-05-01 outline.Rmd
rearranged the percentage info to be in the beginning

@@
 This is a mess, and so we chose two countries from each continent (a high-population and a low-population country) to graph.
+
+***
+We selected a high population from each continent and used the formula below to determine the low population.
+
+Low population = high population * .10
+
+```{r, echo=FALSE}
+#selecting high-population countries from the world population data frame 
+high_pop_countries <- world_pop %>% 
+  filter(Year > 1996 & Country.Name %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
+  group_by(Year)
+
+glimpse(high_pop_countries)
+
+#selecting low-population countries from the world population data frame 
+low_pop_countries <- world_pop %>% 
+  filter(Year > 1996 & Country.Name %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand'))%>% 
+  group_by(Year)
+
+glimpse(low_pop_countries)
+```
 
@@
 *** 
+So we've looked at the deaths due to pollution, but what percentage of the population was affected?
+
+```{r, echo=FALSE, message=FALSE}
+hp_countries_population <- world_pop %>% 
+  filter(Country.Name %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia'), Year > 1996) %>% 
+  group_by(Country.Name) %>% 
+  select(Count) %>% 
+  summarize(average_population = mean(Count))
+
+lp_countries_population <- world_pop %>% 
+  filter(Country.Name %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand'), Year > 1996) %>% 
+  group_by(Country.Name) %>% 
+  select(Count) %>% 
+  summarize(average_population = mean(Count))
+
+#Population Average Table
+kable(list(hp_countries_population, lp_countries_population))
+
+#Graph of Population Average
+ggplot(hp_countries_population)+
+  geom_col(mapping = aes(x=Country.Name, y=average_population))+
+             xlab("Country")+
+             ylab("Average Population")+
+             ggtitle("Average high-population countries")+
+  coord_flip()
+
+ggplot(lp_countries_population)+
+  geom_col(mapping = aes(x=Country.Name, y=average_population))+
+             xlab("Country")+
+             ylab("Average Population")+
+             ggtitle("Average low-population countries")+
+  coord_flip()
+
+#Percentage of those affected 
+affected_high<- joined_high %>% 
+  mutate(percent_high = total_deaths/Count)
+
+percent_high<- affected_high %>% ggplot(aes(x = year, y = percent_high, color = country))+
+  geom_point() +
+  labs(title = "Percentage of High Populated Countries Affected by Air Pollution") +
+  xlab("Year")+
+  ylab("Percent (total_deaths/Count)")
+ggplotly(percent_high)
+
+affected_low <- joined_low %>% 
+  mutate(percent_low = total_deaths/Count)
+
+percent_low <- affected_low %>% ggplot(aes(x = year, y = percent_low, color = country))+
+  geom_point() +
+  labs(title = "Percentage of Low Populated Countries Affected by Air Pollution") +
+  xlab("Year")+
+  ylab("Percent (total_deaths/Count)")
+ggplotly(percent_low)
+```
 
@@
 
-# Reminder: Include this in the beginning 
-First, we split the data into high and low population based on country
-
-Low population = high population * .10
-
-```{r, echo=FALSE}
-#selecting high-population countries from the world population data frame 
-high_pop_countries <- world_pop %>% 
-  filter(Year > 1996 & Country.Name %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
-  group_by(Year)
-
-high_pop_countries
-
-#selecting low-population countries from the world population data frame 
-low_pop_countries <- world_pop %>% 
-  filter(Year > 1996 & Country.Name %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand'))%>% 
-  group_by(Year)
-
-low_pop_countries
-
-```
-
-# Include Earlier to See who's been affected
+# We have this included already
 
@@
 
-This shows us the deaths due to pollution, but what about the average population of those countries at that time?
+# Summary
 
-```{r}
-hp_countries_population <- world_pop %>% 
-  filter(Country.Name %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia'), Year > 1996) %>% 
-  group_by(Country.Name) %>% 
-  select(Count) %>% 
-  summarize(average_population = mean(Count))
-
-lp_countries_population <- world_pop %>% 
-  filter(Country.Name %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand'), Year > 1996) %>% 
-  group_by(Country.Name) %>% 
-  select(Count) %>% 
-  summarize(average_population = mean(Count))
-
-#Population Average Table
-kable(list(hp_countries_population, lp_countries_population))
-
-#Graph of Population Average
-ggplot(hp_countries_population)+
-  geom_col(mapping = aes(x=Country.Name, y=average_population))+
-             xlab("Country")+
-             ylab("Average Population")+
-             ggtitle("Average high-population countries")+
-  coord_flip()
-
-ggplot(lp_countries_population)+
-  geom_col(mapping = aes(x=Country.Name, y=average_population))+
-             xlab("Country")+
-             ylab("Average Population")+
-             ggtitle("Average low-population countries")+
-  coord_flip()
-
-#Percentage of those affected 
-affected_high<- joined_high %>% 
-  mutate(percent_high = total_deaths/Count)
-
-percent_high<- affected_high %>% ggplot(aes(x = year, y = percent_high, color = country))+
-  geom_point() +
-  labs(title = "Percentage of High Populated Countries Affected by Air Pollution") +
-  xlab("Year")+
-  ylab("Percent (total_deaths/Count)")
-ggplotly(percent_high)
-
-affected_low <- joined_low %>% 
-  mutate(percent_low = total_deaths/Count)
-
-percent_low <- affected_low %>% ggplot(aes(x = year, y = percent_low, color = country))+
-  geom_point() +
-  labs(title = "Percentage of Low Populated Countries Affected by Air Pollution") +
-  xlab("Year")+
-  ylab("Percent (total_deaths/Count)")
-ggplotly(percent_low)
-```
-
-# Summary
+# Sources

### 2022-05-02 outline.Rmd
Added world pop data, and a boxplot of pop data (still needs work)

@@
 ---
+
 # Introduction 
@@
 
-# Pollution Data
+# Deaths Data
 * Deaths from pollution:
@@
 * __ozone_deaths__: Ozone is a gas that occurs both in Earth's upper atmosphere and at ground level.
+Ozone in the atmosphere is an important and helpful greenhouse gas, but ground-level ozone is created by extensive use of fossil fuels:
   + Pollutants emitted by cars
@@
 * World Population:
-  + Kaggle: [] 
-  + Author: 
-  + Last update:
-
-***
+  + Kaggle: [https://www.kaggle.com/datasets/imdevskp/world-population-19602018] 
+  + Author: Devakumar K. P. 
+  + Last update: 2 Years ago
 
@@
 
-glimpse(high_pop_countries)
+ggplot(high_pop_countries, aes(x=Country.Name, y=Count))+
+  geom_boxplot()
+
+#glimpse(high_pop_countries)
 
@@
 
-glimpse(low_pop_countries)
+ggplot(low_pop_countries, aes(x=Country.Name, y=Count))+
+  geom_boxplot()
+
+#glimpse(low_pop_countries)
+
+kable(list(high_pop_countries, low_pop_countries))
+
 ```

### 2022-05-02 outline.Rmd
Added title and lable to boxplot

@@
 
-ggplot(high_pop_countries, aes(x=Country.Name, y=Count))+
-  geom_boxplot()
+ggplot(high_pop_countries, aes(x=Country.Name, y=Count, color=Country.Name))+
+  geom_boxplot()+
+  labs(title="Selection of high-population countries", x="Country", y="Population (in millions)")
 
@@
 
-ggplot(low_pop_countries, aes(x=Country.Name, y=Count))+
-  geom_boxplot()
+ggplot(low_pop_countries, aes(x=Country.Name, y=Count, color=Country.Name))+
+  geom_boxplot()+
+  labs(title="Selection of low-population countries", x="Country", y="Population (in millions)")
 

### 2022-05-02 outline.Rmd
organized tables using kable

@@
 
-#glimpse(high_pop_countries)
-
 #selecting low-population countries from the world population data frame 
@@
 
-#glimpse(low_pop_countries)
-
-kable(list(high_pop_countries, low_pop_countries))
+kable(head(list(high_pop_countries[1:6,], low_pop_countries[1:6,])))
 
@@
 ```{r, echo=FALSE}
-#selecting high-population countries from the world population data frame 
-high_pop_countries <- world_pop %>% 
-  filter(Year > 1996 & Country.Name %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
-  group_by(Year)
-
-head(high_pop_countries)
-
-#selecting low-population countries from the world population data frame 
-low_pop_countries <- world_pop %>% 
-  filter(Year > 1996 & Country.Name %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand'))%>% 
-  group_by(Year)
-
-head(low_pop_countries)
+kable(head(list(high_pop_countries[1:6,], low_pop_countries[1:6,])))
 ```
-
-***
 
@@
 
-head(high_pop_death)
-
 #selecting low-population deaths from death dataframe 
@@
 
-head(low_pop_death)
-
+kable(head(list(high_pop_death[1:22,], low_pop_death[1:22,])))
 ```
-
-***
 
@@
 joined_high <- right_join(high_pop_death, high_pop_countries, by= c('country' = 'Country.Name', 'year' = 'Year'))
-head(joined_high)
 
@@
 joined_low <-right_join(low_pop_death, low_pop_countries, by= c('country' = 'Country.Name', 'year' = 'Year'))
-head(joined_low)
 
+kable(head(list(joined_high[1:22,], joined_low[1:22,])))
 ```

### 2022-05-02 outline.Rmd
more updates

@@
 
+***
+
+```{r, echo=FALSE}
+ggplot(data=joined_high, aes(x=year, y=Count, color=country, size = total_deaths/Count)) +
+  geom_line() +
+  labs(title="Percentage of Death Count High Population", subtitle="Bar width is equal to percentage of Deaths")+
+  xlab("Year")+
+  ylab("Population")
+
+ggplot(data=joined_low, aes(x=year, y=Count, color=country, size = total_deaths/Count)) +
+  geom_line() +
+  labs(title="Percentage of Death Count Low Population", subtitle="Bar width is equal to percentage of Deaths")+
+  xlab("Year")+
+  ylab("Population")
+```
+
 # Death Count
@@
 h<- ggplot(deaths_highpop_countries)+
-  geom_col(mapping = aes(x=country, y=average_death_high))+
+  geom_col(mapping = aes(x=country, y=average_death_high, color=country))+
              xlab("Country")+
@@
 l <- ggplot(deaths_lowpop_countries)+
-  geom_col(mapping = aes(x=country, y=average_death_low))+
+  geom_col(mapping = aes(x=country, y=average_death_low, color=country))+
              xlab("Country")+
@@
 deaths_lowpop_countries<- deaths_df %>% 
-  filter(year> 1995 & country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand')) %>% 
+  filter(country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand')) %>% 
   group_by(country) %>% 
@@
 
-#death_lowpop_countries
+#Average Death of both High and Low Populated countries  
 kable(list(deaths_highpop_countries, deaths_lowpop_countries))

### 2022-05-02 outline.Rmd
some graphs and changes

@@
   geom_line() +
-  labs(title="Percentage of Death Count High Population", subtitle="Bar width is equal to percentage of Deaths")+
+  labs(title="Percentage of Death Count in High Population Countries", subtitle="Bar width is equal to percentage of Deaths")+
   xlab("Year")+
@@
   geom_line() +
-  labs(title="Percentage of Death Count Low Population", subtitle="Bar width is equal to percentage of Deaths")+
+  labs(title="Percentage of Death Count in Low Population Countries", subtitle="Bar width is equal to percentage of Deaths")+
   xlab("Year")+
@@
   select(total_deaths) %>% 
-  summarize(average_death_high = mean(total_deaths))
+  summarize(hp_average_death = mean(total_deaths))
 
@@
   select(total_deaths) %>% 
-  summarize(average_death_low = mean(total_deaths))
+  summarize(lp_average_death = mean(total_deaths))
 
@@
 h<- ggplot(deaths_highpop_countries)+
-  geom_col(mapping = aes(x=country, y=average_death_high, color=country))+
+  geom_col(mapping = aes(x=country, y=hp_average_death, color=country))+
              xlab("Country")+
              ylab("Average deaths (per 100,000)")+
-             ggtitle("Average total deaths in high-population countries")+
+             ggtitle("Average deaths in high-population countries")+
   coord_flip()
@@
 l <- ggplot(deaths_lowpop_countries)+
-  geom_col(mapping = aes(x=country, y=average_death_low, color=country))+
+  geom_col(mapping = aes(x=country, y=lp_average_death, color=country))+
              xlab("Country")+
              ylab("Average deaths (per 100,000)")+
-             ggtitle("Average total deaths in low-population countries")+
+             ggtitle("Average deaths in low-population countries")+
   coord_flip()
@@
 
-#Graph of Population Average
-ggplot(hp_countries_population)+
-  geom_col(mapping = aes(x=Country.Name, y=average_population))+
-             xlab("Country")+
-             ylab("Average Population")+
-             ggtitle("Average high-population countries")+
-  coord_flip()
-
-ggplot(lp_countries_population)+
-  geom_col(mapping = aes(x=Country.Name, y=average_population))+
-             xlab("Country")+
-             ylab("Average Population")+
-             ggtitle("Average low-population countries")+
-  coord_flip()
-
 #Percentage of those affected 
 affected_high<- joined_high %>% 
-  mutate(percent_high = total_deaths/Count)
+  mutate(percent_high = total_deaths/mean(Count))
 
@@
   geom_point() +
-  labs(title = "Percentage of High Populated Countries Affected by Air Pollution") +
+  labs(title = "High Population Countries Affected by Air Pollution") +
   xlab("Year")+
-  ylab("Percent (total_deaths/Count)")
+  ylab("Percent (total_deaths/average pop)")
 ggplotly(percent_high)
@@
 affected_low <- joined_low %>% 
-  mutate(percent_low = total_deaths/Count)
+  mutate(percent_low = total_deaths/mean(Count))
 
@@
   geom_point() +
-  labs(title = "Percentage of Low Populated Countries Affected by Air Pollution") +
+  labs(title = "Low Populated Countries Affected by Air Pollution") +
   xlab("Year")+
-  ylab("Percent (total_deaths/Count)")
+  ylab("Percent (total_deaths/average pop)")
 ggplotly(percent_low)
@@
   summarize(avg_indoor = mean(indoor_deaths), avg_outdoor = mean(outdoor_deaths), avg_ozone = mean(ozone_deaths))
-low_poll
+kable(low_poll)
 

### 2022-05-02 outline.Rmd
updates

@@
   select(country, indoor_deaths, outdoor_deaths, ozone_deaths) %>% 
-  summarize(avg_indoor = mean(indoor_deaths), avg_outdoor = mean(outdoor_deaths), avg_ozone = mean(ozone_deaths))
+  summarize(avg_indoor = mean(indoor_deaths), avg_outdoor = mean(outdoor_deaths), avg_ozone = mean(ozone_deaths)) %>% 
+  arrange(desc(avg_indoor),desc(avg_outdoor),desc(avg_ozone))
 
-high_poll
-
-#Plot High Population
-#Indoor Air Pollution
-h_indoor<- ggplot(high_poll, aes(x=country, y = avg_indoor)) + 
-  geom_point() +
-  ggtitle("Average Death due to Indoor Air Pollution for High Population")
-ggplotly(h_indoor)
-
-#Outdoor Air Pollution
-h_outdoor <- ggplot(high_poll, aes(x=country, y = avg_outdoor, color = avg_ozone)) +
-  geom_point() +
-  labs(title = "Average Death due to Outdoor Air Pollution for High Population") +
-  xlab("Country") + 
-  ylab("Average Outdoor Air Pollution Deaths")
-ggplotly(h_outdoor)
+kable(high_poll)
 
@@
 kable(low_poll)
+```
 
-#Plot Low Population
-#Indoor Air Pollution
-l_indoor<- ggplot(low_poll, aes(x=country, y = avg_indoor)) + 
+***
+```{r, echo=FALSE}
+
+#Plot High Population
+# High Indoor Air Pollution
+h_indoor<- ggplot(high_poll, aes(x=country, y = avg_indoor, cex=5)) + 
   geom_point() +
-  ggtitle("Average Death due to Indoor Air Pollution for Low Population")
+  ggtitle("Indoor Air Pollution Deaths in High Population Countries")+
+  xlab("Country")+
+  ylab("Average Deaths")
+ggplotly(h_indoor)
+
+#Low Indoor Air Pollution
+l_indoor<- ggplot(low_poll, aes(x=country, y = avg_indoor, cex=5)) + 
+  geom_point() +
+  ggtitle("Indoor Air Pollution Deaths in Low Population Countries")+
+  xlab("Country")+
+  ylab("Average Deaths")
 ggplotly(l_indoor)
 
-#Outdoor Air Pollution
-l_outdoor <- ggplot(low_poll, aes(x=country, y = avg_outdoor, color = avg_ozone)) +
+# High Outdoor Air Pollution
+h_outdoor <- ggplot(high_poll, aes(x=country, y = avg_outdoor, color = avg_ozone, cex=5)) +
   geom_point() +
-  labs(title = "Average Death due to Outdoor Air Pollution for Low Population")
+  labs(title = "Outdoor Air Pollution Deaths in High Population Countries") +
+  xlab("Country") + 
+  ylab("Average Deaths")
+ggplotly(h_outdoor)
+
+#Low Outdoor Air Pollution
+l_outdoor <- ggplot(low_poll, aes(x = country, y= avg_outdoor, color = avg_ozone, cex=5)) +
+  geom_point() +
+  labs(title = "Outdoor Air Pollution Deaths in Low Population Countries")+
+  xlab("Country")+
+  ylab("Average Deaths")
 ggplotly(l_outdoor)
@@
 
-> Which year had the worst indoor? Outdoor particulate? Outdoor ozone?
+> By comparing each pollutant type, we can determine which year and country had the highest numbers of deaths
 

### 2022-05-02 outline.Rmd
changes were made

@@
 
-# We have this included already
-
-```{r}
-#Mean total deaths from 1996-2017 of high-population countries
-deaths_highpop_countries <- deaths_df %>% 
-  filter(country %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
-  group_by(country) %>% 
-  select(total_deaths) %>% 
-  summarize(average_death_high = mean(total_deaths))
-
-#Mean total deaths from 1990-2017 of high-population countries
-deaths_lowpop_countries<- deaths_df %>% 
-  filter(country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand')) %>% 
-  group_by(country) %>% 
-  select(total_deaths) %>% 
-  summarize(average_death_low = mean(total_deaths))
-
-#Average Death of both High and Low Populated countries  
-kable(list(deaths_highpop_countries, deaths_lowpop_countries))
-
-ggplot(deaths_highpop_countries)+
-  geom_col(mapping = aes(x=country, y=average_death_high))+
-             xlab("Country")+
-             ylab("Average deaths (per 100,000)")+
-             ggtitle("Average total deaths in high-population countries")+
-  coord_flip()
-
-ggplot(deaths_lowpop_countries)+
-  geom_col(mapping = aes(x=country, y=average_death_low))+
-             xlab("Country")+
-             ylab("Average deaths (per 100,000)")+
-             ggtitle("Average total deaths in low-population countries")+
-  coord_flip()
-```
-
 # Summary

### 2022-05-02 outline.Rmd
?

--- 2022-05-02 outline.Rmd
@@
   summarize(avg_indoor = mean(indoor_deaths), avg_outdoor = mean(outdoor_deaths), avg_ozone = mean(ozone_deaths))
-kable(low_poll)
+kable(low_poll, ascending=TRUE)
 

### 2022-05-02 outline.Rmd
summary is done

--- 2022-05-02 outline.Rmd
@@
 
+* Which country has the highest average death count?
+  + High Population: Pakistan
+  + Low Population: Malawi
+* Has the percentage of the affected population decreased or increased over time?
+  + Generally it is decreasing for both High and Low populated countries 
+* Which pollutant type has the greatest number of deaths?
+  + High Population: Indoor Pollution
+  + Low Population: Indoor Pollution
+* How has the death count changed over the past two decades?
+  + 1996-2006: 
+    - High Population: Decreases
+    - Low Population: Decreases
+    - Outliers: Nigeria, Pakistan
+  + 2007-2017: 
+    - High Population: Decreases 
+    - Low Population: Decreases
+    - Outlliers: N/A
+* Which year and country had the highest number of deaths per pollutant type?
+  + We looked at years 1996-2017 
+  + Indoor: Pakistan and Malawi were mainly affected in 1996
+  + Outdoor: Serbia and Pakistan were the top countries. 2011 was the worst for Pakistan. 
+  + Outdoor Ozone: Pakistan and Malawi were the top countries. 1997 was the worst for Pakistan. 
+* Which is worse - outdoor or indoor pollution?
+  + Inconclusive
+
 # Sources

### 2022-05-03 outline.Rmd
Sources

@@
 # Death Count
-> Which country has the highest death count?
+> Which country has the highest average death count?
 
@@
 # Sources
+* [https://www.kaggle.com/datasets/akshat0giri/death-due-to-air-pollution-19902017 ]
+
+* [https://www.worldometers.info/population/world/]
+
+* [https://www.who.int/health-topics/air-pollution#tab=tab_1]
+
+* [https://www.epa.gov/ground-level-ozone-pollution/ground-level-ozone-basics]
+
+* [https://www.health.nsw.gov.au/environment/air/Pages/outdoor-air-pollution.aspx]
+
+* [https://www.mana.md/indoor-air-vs-outdoor-air/]
+
+* [https://www.kaggle.com/datasets/imdevskp/world-population-19602018]

### 2022-05-03 outline.Rmd
updated changes

@@
 deaths_df <- data.frame(read.csv("death-rates-from-air-pollution.csv"))
+glimpse(deaths_df)
 ```
@@
   group_by(country) %>% 
-  filter(year == 2006, country %in% c('United States', 'Brazil','Nigeria', 'Germany', 'Pakistan', 'Australia')) %>%
+  filter(year == 2006, country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'Tonga')) %>%
   select(total_deaths) %>% 
@@
 ```{r, echo=FALSE}
-#Indoor 1996-2017
+#Indoor 1997-2017
 indoor_high<- deaths_df %>% 
-    filter(year> 1995 & country %in% c('United States', 'Brazil','Nigeria', 
+    filter(year> 1996 & country %in% c('United States', 'Brazil','Nigeria', 
                                        'Germany', 'Pakistan', 'Australia')) %>% 
@@
   geom_point(aes(year, indoor_deaths, color = country))+
-  ggtitle("High Populated Indoor Death 1996-2017")+
+  ggtitle("High Populated Indoor Death 1997-2017")+
   xlab("Year")+
@@
 indoor_low<- deaths_df %>% 
-    filter(year> 1995 & country %in% c('Canada', 'Chile', 'Malawi', 
+    filter(year> 1996 & country %in% c('Canada', 'Chile', 'Malawi', 
                                        'Serbia', 'Sri Lanka', 'Tonga')) %>% 
@@
   geom_point(aes(year, indoor_deaths, color = country))+
-  ggtitle("Low Populated Indoor Death 1996-2017")+
+  ggtitle("Low Populated Indoor Death 1997-2017")+
   xlab("Year")+
@@
 indoor_top<- deaths_df %>% 
-    filter(year> 1995 & country %in% c('Pakistan', 'Malawi')) %>% 
+    filter(year> 1996 & country %in% c('Pakistan', 'Malawi')) %>% 
   group_by(country, year, indoor_deaths) %>% 
@@
   geom_point(aes(year, indoor_deaths, color = country))+
-  ggtitle("Greatest Indoor Death 1996-2017")+
+  ggtitle("Greatest Indoor Death 1997-2017")+
   xlab("Year")+
@@
 ```{r, echo= FALSE}
-#Outdoor Particulate 1996-2017
+#Outdoor Particulate 1997-2017
 outdoor_high<- deaths_df %>% 
-    filter(year> 1995 & country %in% c('United States', 'Brazil','Nigeria', 
+    filter(year> 1996 & country %in% c('United States', 'Brazil','Nigeria', 
                                        'Germany', 'Pakistan', 'Australia')) %>% 
@@
   geom_point(aes(year, outdoor_deaths, color = country))+
-  ggtitle("High Populated Outdoor Death 1996-2017")+
+  ggtitle("High Populated Outdoor Death 1997-2017")+
   xlab("Year")+
@@
 outdoor_low<- deaths_df %>% 
-    filter(year> 1995 & country %in% c('Canada', 'Chile', 'Malawi', 
+    filter(year> 1996 & country %in% c('Canada', 'Chile', 'Malawi', 
                                        'Serbia', 'Sri Lanka', 'Tonga')) %>% 
@@
   geom_point(aes(year, outdoor_deaths, color = country))+
-  ggtitle("Low Populated Outdoor Death 1996-2017")+
+  ggtitle("Low Populated Outdoor Death 1997-2017")+
   xlab("Year")+
@@
 outdoor_top<- deaths_df %>% 
-    filter(year> 1995 & country %in% c('Pakistan', 'Serbia')) %>% 
+    filter(year> 1997 & country %in% c('Pakistan', 'Serbia')) %>% 
   group_by(country, year, outdoor_deaths) %>% 
@@
   geom_point(aes(year, outdoor_deaths, color = country))+
-  ggtitle("Greatest Outdoor Death 1996-2017")+
+  ggtitle("Greatest Outdoor Death 1997-2017")+
   xlab("Year")+
@@
 ```{r, echo=FALSE}
-#Outdoor ozone 1996-2017
+#Outdoor ozone 1997-2017
 ozone_high<- deaths_df %>% 
-    filter(year> 1995 & country %in% c('United States', 'Brazil','Nigeria', 
+    filter(year> 1996 & country %in% c('United States', 'Brazil','Nigeria', 
                                        'Germany', 'Pakistan', 'Australia')) %>% 
@@
   geom_point(aes(year, ozone_deaths, color = country))+
-  ggtitle("High Populated Ozone Death 1996-2017")+
+  ggtitle("High Populated Ozone Death 1997-2017")+
   xlab("Year")+
@@
 ozone_low<- deaths_df %>% 
-    filter(year> 1995 & country %in% c('Canada', 'Chile', 'Malawi', 
+    filter(year> 1996 & country %in% c('Canada', 'Chile', 'Malawi', 
                                        'Serbia', 'Sri Lanka', 'Tonga')) %>% 
@@
   geom_point(aes(year, ozone_deaths, color = country))+
-  ggtitle("Low Populated Ozone Death 1996-2017")+
+  ggtitle("Low Populated Ozone Death 1997-2017")+
   xlab("Year")+
@@
 ozone_top<- deaths_df %>% 
-    filter(year> 1995 & country %in% c('Pakistan', 'Malawi')) %>% 
+    filter(year> 1996 & country %in% c('Pakistan', 'Malawi')) %>% 
   group_by(country, year, ozone_deaths) %>% 
@@
   geom_point(aes(year, ozone_deaths, color = country))+
-  ggtitle("Greatest Ozone Death 1996-2017")+
+  ggtitle("Greatest Ozone Death 1997-2017")+
   xlab("Year")+
@@
     - Low Population: Decreases
-    - Outliers: Nigeria, Pakistan
   + 2007-2017: 
@@
     - Low Population: Decreases
-    - Outlliers: N/A
 * Which year and country had the highest number of deaths per pollutant type?
@@
   + Indoor: Pakistan and Malawi were mainly affected in 1996
-  + Outdoor: Serbia and Pakistan were the top countries. 2011 was the worst for Pakistan. 
+  + Outdoor: Serbia and Pakistan were the top countries. 
+    - 2011 was the worst for Pakistan. 
+    - 2005 was the worst for Serbia 
+    - Sri Lanka and Tonga increased, but Sri Lanka had a steep decrease after 2015 
   + Outdoor Ozone: Pakistan and Malawi were the top countries. 1997 was the worst for Pakistan. 
+    - United States was the second highest amount of deaths among the higher populated countries
+    - Pakistan decreased and then slightly increased
 * Which is worse - outdoor or indoor pollution?

### 2022-05-03 outline.Rmd
boxplot

@@
 high_pop_countries <- world_pop %>% 
-  filter(Year > 1996 & Country.Name %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
+  filter(Country.Name %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
   group_by(Year)
 
-ggplot(high_pop_countries, aes(x=Country.Name, y=Count, color=Country.Name))+
+high<- ggplot(high_pop_countries, aes(x=Country.Name, y=Count, color=Country.Name))+
   geom_boxplot()+
   labs(title="Selection of high-population countries", x="Country", y="Population (in millions)")
+ggplotly(high)
 
@@
 low_pop_countries <- world_pop %>% 
-  filter(Year > 1996 & Country.Name %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand'))%>% 
+  filter(Country.Name %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand'))%>% 
   group_by(Year)
 
-ggplot(low_pop_countries, aes(x=Country.Name, y=Count, color=Country.Name))+
+low<- ggplot(low_pop_countries, aes(x=Country.Name, y=Count, color=Country.Name))+
   geom_boxplot()+
   labs(title="Selection of low-population countries", x="Country", y="Population (in millions)")
+ggplotly(low)
 

### 2022-05-03 outline.Rmd
minor changes

@@
 high_pop_countries <- world_pop %>% 
-  filter(Country.Name %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
+  filter(Year> 1995 & Country.Name %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
   group_by(Year)
@@
 low_pop_countries <- world_pop %>% 
-  filter(Country.Name %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand'))%>% 
+  filter(Year> 1995 & Country.Name %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand'))%>% 
   group_by(Year)
@@
 high_pop_death <- deaths_df %>% 
-  filter(year > 1996 & country %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
+  filter(year > 1995 & country %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
   group_by(year)
@@
 low_pop_death <- deaths_df %>% 
-  filter(year > 1996 & country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand')) %>% 
+  filter(year > 1995 & country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand')) %>% 
   group_by(year)
@@
 hp_countries_population <- world_pop %>% 
-  filter(Country.Name %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia'), Year > 1996) %>% 
+  filter(Country.Name %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia'), Year > 1995) %>% 
   group_by(Country.Name) %>% 
@@
 lp_countries_population <- world_pop %>% 
-  filter(Country.Name %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand'), Year > 1996) %>% 
+  filter(Country.Name %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand'), Year > 1995) %>% 
   group_by(Country.Name) %>% 
@@
 > Let's look at the previous two decades and compare the death count
-> Has there been a change?
+
+> has there been a change?
 
@@
 
-The first decade.
+The first decade 1996-2006.
 ```{r, echo=FALSE}
@@
 
-This shows the second decade.
+This shows the second decade 2007-2017.
 ```{r, echo=FALSE}
@@
 indoor_high<- deaths_df %>% 
-    filter(year> 1996 & country %in% c('United States', 'Brazil','Nigeria', 
+    filter(year> 1995 & country %in% c('United States', 'Brazil','Nigeria', 
                                        'Germany', 'Pakistan', 'Australia')) %>% 
@@
 indoor_low<- deaths_df %>% 
-    filter(year> 1996 & country %in% c('Canada', 'Chile', 'Malawi', 
+    filter(year> 1995 & country %in% c('Canada', 'Chile', 'Malawi', 
                                        'Serbia', 'Sri Lanka', 'Tonga')) %>% 
@@
 indoor_top<- deaths_df %>% 
-    filter(year> 1996 & country %in% c('Pakistan', 'Malawi')) %>% 
+    filter(year> 1995 & country %in% c('Pakistan', 'Malawi')) %>% 
   group_by(country, year, indoor_deaths) %>% 
@@
 outdoor_high<- deaths_df %>% 
-    filter(year> 1996 & country %in% c('United States', 'Brazil','Nigeria', 
+    filter(year> 1995 & country %in% c('United States', 'Brazil','Nigeria', 
                                        'Germany', 'Pakistan', 'Australia')) %>% 
@@
 outdoor_low<- deaths_df %>% 
-    filter(year> 1996 & country %in% c('Canada', 'Chile', 'Malawi', 
+    filter(year> 1995 & country %in% c('Canada', 'Chile', 'Malawi', 
                                        'Serbia', 'Sri Lanka', 'Tonga')) %>% 
@@
 outdoor_top<- deaths_df %>% 
-    filter(year> 1997 & country %in% c('Pakistan', 'Serbia')) %>% 
+    filter(year> 1995 & country %in% c('Pakistan', 'Serbia')) %>% 
   group_by(country, year, outdoor_deaths) %>% 
@@
 ozone_high<- deaths_df %>% 
-    filter(year> 1996 & country %in% c('United States', 'Brazil','Nigeria', 
+    filter(year> 1995 & country %in% c('United States', 'Brazil','Nigeria', 
                                        'Germany', 'Pakistan', 'Australia')) %>% 
@@
 ozone_low<- deaths_df %>% 
-    filter(year> 1996 & country %in% c('Canada', 'Chile', 'Malawi', 
+    filter(year> 1995 & country %in% c('Canada', 'Chile', 'Malawi', 
                                        'Serbia', 'Sri Lanka', 'Tonga')) %>% 
@@
 ozone_top<- deaths_df %>% 
-    filter(year> 1996 & country %in% c('Pakistan', 'Malawi')) %>% 
+    filter(year> 1995 & country %in% c('Pakistan', 'Malawi')) %>% 
   group_by(country, year, ozone_deaths) %>% 
@@
     - Sri Lanka and Tonga increased, but Sri Lanka had a steep decrease after 2015 
-  + Outdoor Ozone: Pakistan and Malawi were the top countries. 1997 was the worst for Pakistan. 
+  + Outdoor Ozone: Pakistan and Malawi were the top countries. 
+    - 1997 was the worst for Pakistan. 
+    - 1998 was the worst for Malawi
     - United States was the second highest amount of deaths among the higher populated countries

### 2022-05-03 outline.Rmd
nothing to see here

@@
 percent_high<- affected_high %>% ggplot(aes(x = year, y = percent_high, color = country))+
-  geom_point() +
+  geom_line() +
   labs(title = "High Population Countries Affected by Air Pollution") +

### 2022-05-03 outline.Rmd
serbia year changed

--- 2022-05-03 outline.Rmd
@@
     - 2011 was the worst for Pakistan. 
-    - 2005 was the worst for Serbia 
+    - 1997 was the worst for Serbia 
     - Sri Lanka and Tonga increased, but Sri Lanka had a steep decrease after 2015 

### 2022-05-03 outline.Rmd
changed high and low ordering

@@
 deaths_df <- data.frame(read.csv("death-rates-from-air-pollution.csv"))
-glimpse(deaths_df)
 ```
@@
 high_pop_countries <- world_pop %>% 
-  filter(Year> 1995 & Country.Name %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
+  filter(Year > 1996 & Country.Name %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
   group_by(Year)
 
-high<- ggplot(high_pop_countries, aes(x=Country.Name, y=Count, color=Country.Name))+
+ggplot(high_pop_countries, aes(x=Country.Name, y=Count, color=Country.Name))+
   geom_boxplot()+
   labs(title="Selection of high-population countries", x="Country", y="Population (in millions)")
-ggplotly(high)
 
@@
 low_pop_countries <- world_pop %>% 
-  filter(Year> 1995 & Country.Name %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand'))%>% 
+  filter(Year > 1996 & Country.Name %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand'))%>% 
   group_by(Year)
 
-low<- ggplot(low_pop_countries, aes(x=Country.Name, y=Count, color=Country.Name))+
+ggplot(low_pop_countries, aes(x=Country.Name, y=Count, color=Country.Name))+
   geom_boxplot()+
   labs(title="Selection of low-population countries", x="Country", y="Population (in millions)")
-ggplotly(low)
 
@@
 high_pop_death <- deaths_df %>% 
-  filter(year > 1995 & country %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
+  filter(year > 1996 & country %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
   group_by(year)
@@
 low_pop_death <- deaths_df %>% 
-  filter(year > 1995 & country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand')) %>% 
+  filter(year > 1996 & country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand')) %>% 
   group_by(year)
@@
 # Death Count
-> Which country has the highest average death count?
+> Which country has the highest death count?
 
@@
 hp_countries_population <- world_pop %>% 
-  filter(Country.Name %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia'), Year > 1995) %>% 
+  filter(Country.Name %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia'), Year > 1996) %>% 
   group_by(Country.Name) %>% 
@@
 lp_countries_population <- world_pop %>% 
-  filter(Country.Name %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand'), Year > 1995) %>% 
+  filter(Country.Name %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand'), Year > 1996) %>% 
   group_by(Country.Name) %>% 
@@
   summarize(avg_indoor = mean(indoor_deaths), avg_outdoor = mean(outdoor_deaths), avg_ozone = mean(ozone_deaths))
+
+kable(low_poll, ascending=TRUE)
+
 kable(low_poll)
@@
 > Let's look at the previous two decades and compare the death count
-
-> has there been a change?
+> Has there been a change?
 
@@
   group_by(country) %>% 
-  filter(year == 2006, country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'Tonga')) %>%
+  filter(year == 2006, country %in% c('United States', 'Brazil','Nigeria', 'Germany', 'Pakistan', 'Australia')) %>%
   select(total_deaths) %>% 
@@
 
-The first decade 1996-2006.
+The first decade.
 ```{r, echo=FALSE}
@@
 
-This shows the second decade 2007-2017.
+This shows the second decade.
 ```{r, echo=FALSE}
@@
 ```{r, echo=FALSE}
-#Indoor 1997-2017
+#Indoor 1996-2017
 indoor_high<- deaths_df %>% 
@@
   geom_point(aes(year, indoor_deaths, color = country))+
-  ggtitle("High Populated Indoor Death 1997-2017")+
+  ggtitle("High Populated Indoor Death 1996-2017")+
   xlab("Year")+
@@
   geom_point(aes(year, indoor_deaths, color = country))+
-  ggtitle("Low Populated Indoor Death 1997-2017")+
+  ggtitle("Low Populated Indoor Death 1996-2017")+
   xlab("Year")+
@@
   geom_point(aes(year, indoor_deaths, color = country))+
-  ggtitle("Greatest Indoor Death 1997-2017")+
+  ggtitle("Greatest Indoor Death 1996-2017")+
   xlab("Year")+
@@
 ```{r, echo= FALSE}
-#Outdoor Particulate 1997-2017
+#Outdoor Particulate 1996-2017
 outdoor_high<- deaths_df %>% 
@@
   geom_point(aes(year, outdoor_deaths, color = country))+
-  ggtitle("High Populated Outdoor Death 1997-2017")+
+  ggtitle("High Populated Outdoor Death 1996-2017")+
   xlab("Year")+
@@
   geom_point(aes(year, outdoor_deaths, color = country))+
-  ggtitle("Low Populated Outdoor Death 1997-2017")+
+  ggtitle("Low Populated Outdoor Death 1996-2017")+
   xlab("Year")+
@@
   geom_point(aes(year, outdoor_deaths, color = country))+
-  ggtitle("Greatest Outdoor Death 1997-2017")+
+  ggtitle("Greatest Outdoor Death 1996-2017")+
   xlab("Year")+
@@
 ```{r, echo=FALSE}
-#Outdoor ozone 1997-2017
+#Outdoor ozone 1996-2017
 ozone_high<- deaths_df %>% 
@@
   geom_point(aes(year, ozone_deaths, color = country))+
-  ggtitle("High Populated Ozone Death 1997-2017")+
+  ggtitle("High Populated Ozone Death 1996-2017")+
   xlab("Year")+
@@
   geom_point(aes(year, ozone_deaths, color = country))+
-  ggtitle("Low Populated Ozone Death 1997-2017")+
+  ggtitle("Low Populated Ozone Death 1996-2017")+
   xlab("Year")+
@@
   geom_point(aes(year, ozone_deaths, color = country))+
-  ggtitle("Greatest Ozone Death 1997-2017")+
+  ggtitle("Greatest Ozone Death 1996-2017")+
   xlab("Year")+
@@
 
-* Which country has the highest average death count?
-  + High Population: Pakistan
-  + Low Population: Malawi
-* Has the percentage of the affected population decreased or increased over time?
-  + Generally it is decreasing for both High and Low populated countries 
-* Which pollutant type has the greatest number of deaths?
-  + High Population: Indoor Pollution
-  + Low Population: Indoor Pollution
-* How has the death count changed over the past two decades?
-  + 1996-2006: 
-    - High Population: Decreases
-    - Low Population: Decreases
-  + 2007-2017: 
-    - High Population: Decreases 
-    - Low Population: Decreases
-* Which year and country had the highest number of deaths per pollutant type?
-  + We looked at years 1996-2017 
-  + Indoor: Pakistan and Malawi were mainly affected in 1996
-  + Outdoor: Serbia and Pakistan were the top countries. 
-    - 2011 was the worst for Pakistan. 
-    - 1997 was the worst for Serbia 
-    - Sri Lanka and Tonga increased, but Sri Lanka had a steep decrease after 2015 
-  + Outdoor Ozone: Pakistan and Malawi were the top countries. 
-    - 1997 was the worst for Pakistan. 
-    - 1998 was the worst for Malawi
-    - United States was the second highest amount of deaths among the higher populated countries
-    - Pakistan decreased and then slightly increased
-* Which is worse - outdoor or indoor pollution?
-  + Inconclusive
-
 # Sources
-* [https://www.kaggle.com/datasets/akshat0giri/death-due-to-air-pollution-19902017 ]
-
-* [https://www.worldometers.info/population/world/]
-
-* [https://www.who.int/health-topics/air-pollution#tab=tab_1]
-
-* [https://www.epa.gov/ground-level-ozone-pollution/ground-level-ozone-basics]
-
-* [https://www.health.nsw.gov.au/environment/air/Pages/outdoor-air-pollution.aspx]
-
-* [https://www.mana.md/indoor-air-vs-outdoor-air/]
-
-* [https://www.kaggle.com/datasets/imdevskp/world-population-19602018]

### 2022-05-03 outline.Rmd

--- 2022-05-03 outline.Rmd
@@
 
+* High Populated Countries: 
+  + Outdoor pollution seems to be more detrimental with the exception of two countries in this sample set.
+
 * Low Populated Countries: 
   + it's half-and-half
-
-* High Populated Countries: 
-  + Outdoor pollution seems to be more detrimental with the exception of two countries in this sample set.
 

### 2022-05-06 Final Report.Rmd
final report file

---
title: "Final Report"
author: "Alison Cheek & Elizabeth Bekele"
date: '2022-05-06'
output: html_document
---

### 2022-05-06 Final Report.Rmd
copy+pasted from outline

@@
 ---
+
+# Introduction 
+* Air pollution can be detrimental to both our health and the climate
+  + Outdoor and indoor air pollution cause chronic pain, respiratory diseases, shortened lifespan
+  + Air pollution kills about 7 million people worldwide every year 
+  + Hopefully this information can showcase the importance of air pollution and that we should be more mindful about our planet
+
+* Overview
+  + We will see how different air pollution types affect the population 
+  + compare past and present population numbers
+  + determine which air pollutant type has the highest associated death rate
+
+# Packages Required 
+```{r, message=FALSE, warning=FALSE}
+#This will allow us to filter through our data 
+library(tidyverse)
+library(dplyr)
+#This will help us plot figures to showcase our findings
+library(ggplot2)
+#This will help us organize and display our data as necessary 
+library(knitr)
+library(kableExtra)
+#This expands our plot uses 
+library(plotly)
+#Scientific Notation Disabled 
+options(scipen=999)
+```
+
+# Deaths Data
+* Deaths from pollution:
+  + Kaggle: [https://www.kaggle.com/datasets/akshat0giri/death-due-to-air-pollution-19902017] 
+  + Author: Akshat Giri
+  + Last update: 2 years ago
+
+Import the deaths-due-to-air-pollution data
+```{r}
+deaths_df <- data.frame(read.csv("death-rates-from-air-pollution.csv"))
+glimpse(deaths_df)
+```
+
+We are going to rename a few of the columns and glimpse the data
+```{r}
+colnames(deaths_df) <- c("country", "acronym", "year", "total_deaths", "indoor_deaths", "outdoor_deaths", "ozone_deaths")
+
+glimpse(deaths_df)
+```
+
+# Data Variables
+Variables that interest us here include:
+
+* __country__ 
+* __total_deaths__: per 100,000
+* __indoor_deaths__: Indoor air pollution is considered pollution that occurs in the household. Cooking with solid fuels:
+  + Wood
+  + Crop waste, dung
+  + Charcoal, coal
+* __outdoor_deaths__: Outdoor air or ambient air are emissions caused by combustion processes from motor vehicles, solid fuel burning and industry 
+  + Ozone (O3)
+  + Particulate matter (PM10 and PM2.5)
+  + Nitrogen dioxide (NO2)
+  + Carbon monoxide (CO)
+  + Sulfur dioxide (SO2)
+* __ozone_deaths__: Ozone is a gas that occurs both in Earth's upper atmosphere and at ground level.
+Ozone in the atmosphere is an important and helpful greenhouse gas, but ground-level ozone is created by extensive use of fossil fuels:
+  + Pollutants emitted by cars
+  + Power plants, industrial boilers, refineries, chemical plants
+
+# World Population Data 
+* World Population:
+  + Kaggle: [https://www.kaggle.com/datasets/imdevskp/world-population-19602018] 
+  + Author: Devakumar K. P. 
+  + Last update: 2 Years ago
+
+Now, let's take a look at the population data.
+```{r}
+world_pop <- read.csv("population_total_long.csv")
+glimpse(world_pop)
+```
+
+***
+
+To get a general idea of 'deaths-dataframe' we made, let's make a plots to see what's happening. This is a plot of indoor x outdoor deaths around the world by country.
+
+```{r, echo=FALSE}
+d <- ggplot(deaths_df, aes(x = indoor_deaths, y = outdoor_deaths, text = paste0(country, ", ", year) )) + geom_point() +
+  ggtitle("Outdoor Deaths vs Indoor Deaths")
+ggplotly(d)
+```
+
+This is a mess, and so we chose two countries from each continent (a high-population and a low-population country) to graph.
+
+***
+We selected a high population from each continent and used the formula below to determine the low population.
+
+Low population = high population * .10
+
+```{r, echo=FALSE}
+#selecting high-population countries from the world population data frame 
+high_pop_countries <- world_pop %>% 
+  filter(Year> 1995 & Country.Name %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
+  group_by(Year)
+
+high<- ggplot(high_pop_countries, aes(x=Country.Name, y=Count, color=Country.Name))+
+  geom_boxplot()+
+  labs(title="Selection of high-population countries", x="Country", y="Population (in millions)")
+ggplotly(high)
+
+#selecting low-population countries from the world population data frame 
+low_pop_countries <- world_pop %>% 
+  filter(Year> 1995 & Country.Name %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand'))%>% 
+  group_by(Year)
+
+low<- ggplot(low_pop_countries, aes(x=Country.Name, y=Count, color=Country.Name))+
+  geom_boxplot()+
+  labs(title="Selection of low-population countries", x="Country", y="Population (in millions)")
+ggplotly(low)
+
+kable(head(list(high_pop_countries[1:6,], low_pop_countries[1:6,])))
+
+```
+
+# Combine Data Sets
+
+First let's look at a table of the high and low populated countries using the world population data set.
+```{r, echo=FALSE}
+kable(head(list(high_pop_countries[1:6,], low_pop_countries[1:6,])))
+```
+
+Next, we are going to see the death count for high and low populated countries using the deaths dataframe.
+```{r, echo=FALSE}
+#selecting high-population deaths from death dataframe 
+high_pop_death <- deaths_df %>% 
+  filter(year > 1995 & country %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
+  group_by(year)
+
+#selecting low-population deaths from death dataframe 
+low_pop_death <- deaths_df %>% 
+  filter(year > 1995 & country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand')) %>% 
+  group_by(year)
+
+kable(head(list(high_pop_death[1:22,], low_pop_death[1:22,])))
+```
+
+Lastly, we will join the population and and deaths with its respected country.
+```{r, message=FALSE, echo=FALSE}
+#Combined High Population with Pollution Death
+joined_high <- right_join(high_pop_death, high_pop_countries, by= c('country' = 'Country.Name', 'year' = 'Year'))
+
+#Combined Low Population with Pollution Death
+joined_low <-right_join(low_pop_death, low_pop_countries, by= c('country' = 'Country.Name', 'year' = 'Year'))
+
+kable(head(list(joined_high[1:22,], joined_low[1:22,])))
+```
+
+***
+
+```{r, echo=FALSE}
+ggplot(data=joined_high, aes(x=year, y=Count, color=country, size = total_deaths/Count)) +
+  geom_line() +
+  labs(title="Percentage of Death Count in High Population Countries", subtitle="Bar width is equal to percentage of Deaths")+
+  xlab("Year")+
+  ylab("Population")
+
+ggplot(data=joined_low, aes(x=year, y=Count, color=country, size = total_deaths/Count)) +
+  geom_line() +
+  labs(title="Percentage of Death Count in Low Population Countries", subtitle="Bar width is equal to percentage of Deaths")+
+  xlab("Year")+
+  ylab("Population")
+```
+
+# Death Count
+> Which country has the highest average death count?
+
+Let's make a table depicting the high and low populated countries and their respected death count due to pollution.
+
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
+#Mean total deaths of high-population countries
+deaths_highpop_countries <- deaths_df %>% 
+  filter(country %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
+  group_by(country) %>% 
+  select(total_deaths) %>% 
+  summarize(hp_average_death = mean(total_deaths))
+
+#Mean total deaths of high-population countries
+deaths_lowpop_countries<- deaths_df %>% 
+  filter(country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand')) %>% 
+  group_by(country) %>% 
+  select(total_deaths) %>% 
+  summarize(lp_average_death = mean(total_deaths))
+
+#Side-by-side Table to view deaths 
+kable(list(deaths_highpop_countries, deaths_lowpop_countries))
+
+```
+
+***
+
+Here's a graph to clearly visualize the previous table
+```{r, echo=FALSE, message=FALSE}
+#Plot High Population Deaths (average)
+h<- ggplot(deaths_highpop_countries)+
+  geom_col(mapping = aes(x=country, y=hp_average_death, color=country))+
+             xlab("Country")+
+             ylab("Average deaths (per 100,000)")+
+             ggtitle("Average deaths in high-population countries")+
+  coord_flip()
+
+ggplotly(h)
+
+#Plot of Low Population Deaths (average)
+l <- ggplot(deaths_lowpop_countries)+
+  geom_col(mapping = aes(x=country, y=lp_average_death, color=country))+
+             xlab("Country")+
+             ylab("Average deaths (per 100,000)")+
+             ggtitle("Average deaths in low-population countries")+
+  coord_flip()
+
+ggplotly(l)
+```
+
+*** 
+So we've looked at the deaths due to pollution, but what percentage of the population was affected?
+
+```{r, echo=FALSE, message=FALSE}
+hp_countries_population <- world_pop %>% 
+  filter(Country.Name %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia'), Year > 1995) %>% 
+  group_by(Country.Name) %>% 
+  select(Count) %>% 
+  summarize(average_population = mean(Count))
+
+lp_countries_population <- world_pop %>% 
+  filter(Country.Name %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand'), Year > 1995) %>% 
+  group_by(Country.Name) %>% 
+  select(Count) %>% 
+  summarize(average_population = mean(Count))
+
+#Population Average Table
+kable(list(hp_countries_population, lp_countries_population))
+
+#Percentage of those affected 
+affected_high<- joined_high %>% 
+  mutate(percent_high = total_deaths/mean(Count))
+
+percent_high<- affected_high %>% ggplot(aes(x = year, y = percent_high, color = country))+
+  geom_point() +
+  labs(title = "High Population Countries Affected by Air Pollution") +
+  xlab("Year")+
+  ylab("Percent (total_deaths/average pop)")
+ggplotly(percent_high)
+
+affected_low <- joined_low %>% 
+  mutate(percent_low = total_deaths/mean(Count))
+
+percent_low <- affected_low %>% ggplot(aes(x = year, y = percent_low, color = country))+
+  geom_point() +
+  labs(title = "Low Populated Countries Affected by Air Pollution") +
+  xlab("Year")+
+  ylab("Percent (total_deaths/average pop)")
+ggplotly(percent_low)
+```
+
+# Pollution Types
+> Which type of pollution has the greatest number of deaths?
+
+```{r, echo=FALSE, message=FALSE}
+#High Population Pollutant Averages
+high_poll <- deaths_df %>% 
+  group_by(country) %>% 
+  filter(country %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
+  select(country, indoor_deaths, outdoor_deaths, ozone_deaths) %>% 
+  summarize(avg_indoor = mean(indoor_deaths), avg_outdoor = mean(outdoor_deaths), avg_ozone = mean(ozone_deaths)) %>% 
+  arrange(desc(avg_indoor),desc(avg_outdoor),desc(avg_ozone))
+
+kable(high_poll)
+
+#Low Population Pollutant Averages 
+low_poll <- deaths_df %>% 
+  group_by(country) %>% 
+  filter(country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand')) %>% 
+  select(country, indoor_deaths, outdoor_deaths, ozone_deaths) %>% 
+  summarize(avg_indoor = mean(indoor_deaths), avg_outdoor = mean(outdoor_deaths), avg_ozone = mean(ozone_deaths))
+
+kable(low_poll, ascending=TRUE)
+
+kable(low_poll)
+```
+
+***
+```{r, echo=FALSE}
+
+#Plot High Population
+# High Indoor Air Pollution
+h_indoor<- ggplot(high_poll, aes(x=country, y = avg_indoor, cex=5)) + 
+  geom_point() +
+  ggtitle("Indoor Air Pollution Deaths in High Population Countries")+
+  xlab("Country")+
+  ylab("Average Deaths")
+ggplotly(h_indoor)
+
+#Low Indoor Air Pollution
+l_indoor<- ggplot(low_poll, aes(x=country, y = avg_indoor, cex=5)) + 
+  geom_point() +
+  ggtitle("Indoor Air Pollution Deaths in Low Population Countries")+
+  xlab("Country")+
+  ylab("Average Deaths")
+ggplotly(l_indoor)
+
+# High Outdoor Air Pollution
+h_outdoor <- ggplot(high_poll, aes(x=country, y = avg_outdoor, color = avg_ozone, cex=5)) +
+  geom_point() +
+  labs(title = "Outdoor Air Pollution Deaths in High Population Countries") +
+  xlab("Country") + 
+  ylab("Average Deaths")
+ggplotly(h_outdoor)
+
+#Low Outdoor Air Pollution
+l_outdoor <- ggplot(low_poll, aes(x = country, y= avg_outdoor, color = avg_ozone, cex=5)) +
+  geom_point() +
+  labs(title = "Outdoor Air Pollution Deaths in Low Population Countries")+
+  xlab("Country")+
+  ylab("Average Deaths")
+ggplotly(l_outdoor)
+```
+
+***
+
+# Pollution Over Time 
+> Let's look at the previous two decades and compare the death count
+
+> has there been a change?
+
+This is the first decade 1996-2006
+```{r, echo=FALSE, message=FALSE}
+# Look at the trend of deaths within the last decade from low and high populated countries
+death_96_high<- deaths_df %>% 
+  group_by(country) %>% 
+  filter(year == 1996, country %in% c('United States', 'Brazil','Nigeria', 'Germany', 'Pakistan', 'Australia')) %>%
+  select(total_deaths) %>% 
+  summarize(High_Deaths_96 = mean(total_deaths))
+
+death_96_low<- deaths_df %>% 
+  group_by(country) %>% 
+  filter(year == 1996, country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'Tonga')) %>%
+  select(total_deaths) %>% 
+  summarize(Low_Deaths_96 = mean(total_deaths))
+
+death_01_high<- deaths_df %>% 
+  group_by(country) %>% 
+  filter(year == 2001, country %in% c('United States', 'Brazil','Nigeria', 'Germany', 'Pakistan', 'Australia')) %>%
+  select(total_deaths) %>% 
+  summarize(High_Deaths_01 = mean(total_deaths))
+
+death_01_low<- deaths_df %>% 
+  group_by(country) %>% 
+  filter(year == 2001, country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'Tonga')) %>%
+  select(total_deaths) %>% 
+  summarize(Low_Deaths_01 = mean(total_deaths))
+
+death_06_high<- deaths_df %>% 
+  group_by(country) %>% 
+  filter(year == 2006, country %in% c('United States', 'Brazil','Nigeria', 'Germany', 'Pakistan', 'Australia')) %>%
+  select(total_deaths) %>% 
+  summarize(High_Deaths_06 = mean(total_deaths))
+
+death_06_low<- deaths_df %>% 
+  group_by(country) %>% 
+  filter(year == 2006, country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'Tonga')) %>%
+  select(total_deaths) %>% 
+  summarize(Low_Deaths_06 = mean(total_deaths))
+
+#cohesive table so it's easier to visualize 
+#Low Population Table 2007-2017
+all_low_first <- death_96_low %>% 
+  mutate(death_01_low) %>% 
+  mutate(death_06_low)
+
+#High Population Table 2007-2017
+all_high_first <- death_96_high %>% 
+  mutate(death_01_high) %>% 
+  mutate(death_06_high)
+
+kable(list(all_high_first))
+kable(list(all_low_first))
+```
+***
+
+This is the second decade 2007-2017
+```{r, echo=FALSE, message=FALSE}
+# Look at the trend of deaths within the last decade from low and high populated countries
+death_07_high<- deaths_df %>% 
+  group_by(country) %>% 
+  filter(year == 2007, country %in% c('United States', 'Brazil','Nigeria', 'Germany', 'Pakistan', 'Australia')) %>%
+  select(total_deaths) %>% 
+  summarize(High_Deaths_07 = mean(total_deaths))
+
+death_07_low<- deaths_df %>% 
+  group_by(country) %>% 
+  filter(year == 2007, country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'Tonga')) %>%
+  select(total_deaths) %>% 
+  summarize(Low_Deaths_07 = mean(total_deaths))
+
+death_12_high<- deaths_df %>% 
+  group_by(country) %>% 
+  filter(year == 2012, country %in% c('United States', 'Brazil','Nigeria', 'Germany', 'Pakistan', 'Australia')) %>%
+  select(total_deaths) %>% 
+  summarize(High_Deaths_12 = mean(total_deaths))
+
+death_12_low<- deaths_df %>% 
+  group_by(country) %>% 
+  filter(year == 2012, country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'Tonga')) %>%
+  select(total_deaths) %>% 
+  summarize(Low_Deaths_12 = mean(total_deaths))
+
+death_17_high<- deaths_df %>% 
+  group_by(country) %>% 
+  filter(year == 2017, country %in% c('United States', 'Brazil','Nigeria', 'Germany', 'Pakistan', 'Australia')) %>%
+  select(total_deaths) %>% 
+  summarize(High_Deaths_17 = mean(total_deaths))
+
+death_17_low<- deaths_df %>% 
+  group_by(country) %>% 
+  filter(year == 2017, country %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'Tonga')) %>%
+  select(total_deaths) %>% 
+  summarize(Low_Deaths_17 = mean(total_deaths))
+
+#cohesive table so it's easier to visualize 
+#Low Population Table 2007-2017
+all_low_second <- death_07_low %>% 
+  mutate(death_12_low) %>% 
+  mutate(death_17_low)
+
+#High Population Table 2007-2017
+all_high_second <- death_07_high %>% 
+  mutate(death_12_high) %>% 
+  mutate(death_17_high)
+
+kable(list(all_high_second))
+kable(list(all_low_second))
+```
+
+***
+
+Let's graph the previous tables!
+
+The first decade 1996-2006.
+```{r, echo=FALSE}
+#High Population Deaths 1996-2006 
+p_high_first <- all_high_first %>%
+  group_by(country) %>% 
+  ggplot(aes(x= High_Deaths_96, y= High_Deaths_01, color= High_Deaths_06)) +
+  geom_point() +
+  facet_wrap(~country)
+
+interact_high_first<- p_high_first + labs(title="High Populated Countries", x="1996 Deaths", y="2001 Deaths")
+
+ggplotly(interact_high_first)
+
+#Low Population Deaths 1996-2006
+p_low_first <- all_low_first %>%
+  group_by(country) %>% 
+  ggplot(aes(x= Low_Deaths_96, y= Low_Deaths_01, color= Low_Deaths_06)) +
+  geom_point() +
+  facet_wrap(~country)
+
+interact_low_first<- p_low_first + labs(title="Low Populated Countries", x="1996 Deaths", y="2001 Deaths")
+
+ggplotly(interact_low_first)
+```
+
+***
+
+This shows the second decade 2007-2017.
+```{r, echo=FALSE}
+#High Population Deaths 2007-2017
+p_high_second <- all_high_second %>%
+  group_by(country) %>% 
+  ggplot(aes(x= High_Deaths_07, y= High_Deaths_12, color= High_Deaths_17)) +
+  geom_point() +
+  facet_wrap(~country)
+
+interact_high_second<- p_high_second + labs(title="High Populated Countries", x="2007 Deaths", y="2012 Deaths")
+
+ggplotly(interact_high_second)
+
+#Low Population Deaths 2007-2017
+p_low_second <- all_low_second %>%
+  group_by(country) %>% 
+  ggplot(aes(x= Low_Deaths_07, y= Low_Deaths_12, color= Low_Deaths_17)) +
+  geom_point() +
+  facet_wrap(~country)
+
+interact_low_second<- p_low_second + labs(title="Low Populated Countries", x="2007 Deaths", y="2012 Deaths")
+
+ggplotly(interact_low_second)
+
+```
+
+***
+
+> By comparing each pollutant type, we can determine which year and country had the highest numbers of deaths
+
+Indoor Deaths
+```{r, echo=FALSE}
+#Indoor 1997-2017
+indoor_high<- deaths_df %>% 
+    filter(year> 1995 & country %in% c('United States', 'Brazil','Nigeria', 
+                                       'Germany', 'Pakistan', 'Australia')) %>% 
+  group_by(country, year, indoor_deaths) %>% 
+  ggplot()+
+  geom_point(aes(year, indoor_deaths, color = country))+
+  ggtitle("High Populated Indoor Death 1997-2017")+
+  xlab("Year")+
+  ylab("Deaths per 100,000")
+ggplotly(indoor_high)
+
+indoor_low<- deaths_df %>% 
+    filter(year> 1995 & country %in% c('Canada', 'Chile', 'Malawi', 
+                                       'Serbia', 'Sri Lanka', 'Tonga')) %>% 
+  group_by(country, year, indoor_deaths) %>% 
+  ggplot()+
+  geom_point(aes(year, indoor_deaths, color = country))+
+  ggtitle("Low Populated Indoor Death 1997-2017")+
+  xlab("Year")+
+  ylab("Deaths per 100,000")
+ggplotly(indoor_low)
+
+#Pakistan and Malawi were both mainly affected in 1996
+indoor_top<- deaths_df %>% 
+    filter(year> 1995 & country %in% c('Pakistan', 'Malawi')) %>% 
+  group_by(country, year, indoor_deaths) %>% 
+  ggplot()+
+  geom_point(aes(year, indoor_deaths, color = country))+
+  ggtitle("Greatest Indoor Death 1997-2017")+
+  xlab("Year")+
+  ylab("Deaths per 100,000")
+ggplotly(indoor_top)
+```
+
+***
+
+Outdoor Deaths
+```{r, echo= FALSE}
+#Outdoor Particulate 1997-2017
+outdoor_high<- deaths_df %>% 
+    filter(year> 1995 & country %in% c('United States', 'Brazil','Nigeria', 
+                                       'Germany', 'Pakistan', 'Australia')) %>% 
+  group_by(country, year, outdoor_deaths) %>% 
+  ggplot()+
+  geom_point(aes(year, outdoor_deaths, color = country))+
+  ggtitle("High Populated Outdoor Death 1997-2017")+
+  xlab("Year")+
+  ylab("Deaths per 100,000")
+ggplotly(outdoor_high)
+
+outdoor_low<- deaths_df %>% 
+    filter(year> 1995 & country %in% c('Canada', 'Chile', 'Malawi', 
+                                       'Serbia', 'Sri Lanka', 'Tonga')) %>% 
+  group_by(country, year, outdoor_deaths) %>% 
+  ggplot()+
+  geom_point(aes(year, outdoor_deaths, color = country))+
+  ggtitle("Low Populated Outdoor Death 1997-2017")+
+  xlab("Year")+
+  ylab("Deaths per 100,000")
+ggplotly(outdoor_low)
+
+#Serbia and Pakistan were the top countries affected. 
+#2011 Pakistan had the greatest amount of Outdoor Deaths
+outdoor_top<- deaths_df %>% 
+    filter(year> 1995 & country %in% c('Pakistan', 'Serbia')) %>% 
+  group_by(country, year, outdoor_deaths) %>% 
+  ggplot()+
+  geom_point(aes(year, outdoor_deaths, color = country))+
+  ggtitle("Greatest Outdoor Death 1997-2017")+
+  xlab("Year")+
+  ylab("Deaths per 100,000")
+ggplotly(outdoor_top)
+```
+
+***
+Ozone Deaths 
+```{r, echo=FALSE}
+#Outdoor ozone 1997-2017
+ozone_high<- deaths_df %>% 
+    filter(year> 1995 & country %in% c('United States', 'Brazil','Nigeria', 
+                                       'Germany', 'Pakistan', 'Australia')) %>% 
+  group_by(country, year, ozone_deaths) %>% 
+  ggplot()+
+  geom_point(aes(year, ozone_deaths, color = country))+
+  ggtitle("High Populated Ozone Death 1997-2017")+
+  xlab("Year")+
+  ylab("Deaths per 100,000")
+ggplotly(ozone_high)
+
+ozone_low<- deaths_df %>% 
+    filter(year> 1995 & country %in% c('Canada', 'Chile', 'Malawi', 
+                                       'Serbia', 'Sri Lanka', 'Tonga')) %>% 
+  group_by(country, year, ozone_deaths) %>% 
+  ggplot()+
+  geom_point(aes(year, ozone_deaths, color = country))+
+  ggtitle("Low Populated Ozone Death 1997-2017")+
+  xlab("Year")+
+  ylab("Deaths per 100,000")
+ggplotly(ozone_low)
+
+#Malawi and Pakistan were the top countries affected 
+#1997 had the greatest amount of ozone death in Pakistan
+ozone_top<- deaths_df %>% 
+    filter(year> 1995 & country %in% c('Pakistan', 'Malawi')) %>% 
+  group_by(country, year, ozone_deaths) %>% 
+  ggplot()+
+  geom_point(aes(year, ozone_deaths, color = country))+
+  ggtitle("Greatest Ozone Death 1997-2017")+
+  xlab("Year")+
+  ylab("Deaths per 100,000")
+ggplotly(ozone_top)
+```
+
+# Which is worse? 
+> outdoor or indoor pollution?
+
+Let's reintroduce a graph we looked at earlier. Instead this time we will combine the pollutant types together.
+```{r, echo=FALSE}
+#Plot High Population
+h_pollutant<- ggplot(high_poll, aes(x=avg_indoor, y = avg_outdoor, color= avg_ozone)) + 
+  geom_point() +
+  ggtitle("High Population Pollutant Type Death Counts")
+ggplotly(h_pollutant)
+
+#Plot Low Population
+l_pollutant<- ggplot(low_poll, aes(x=avg_indoor, y = avg_outdoor, color= avg_ozone)) + 
+  geom_point() +
+  ggtitle("Low Population Pollutant Type Death Counts")
+ggplotly(l_pollutant)
+```
+
+We cannot conclude which is worse.
+
+* High Populated Countries: 
+  + Outdoor pollution seems to be more detrimental with the exception of two countries in this sample set.
+
+* Low Populated Countries: 
+  + it's half-and-half
+
+# Summary
+
+* Which country has the highest average death count?
+  + High Population: Pakistan
+  + Low Population: Malawi
+* Has the percentage of the affected population decreased or increased over time?
+  + Generally it is decreasing for both High and Low populated countries 
+* Which pollutant type has the greatest number of deaths?
+  + High Population: Indoor Pollution
+  + Low Population: Indoor Pollution
+* How has the death count changed over the past two decades?
+  + 1996-2006: 
+    - High Population: Decreases
+    - Low Population: Decreases
+  + 2007-2017: 
+    - High Population: Decreases 
+    - Low Population: Decreases
+* Which year and country had the highest number of deaths per pollutant type?
+  + We looked at years 1996-2017 
+  + Indoor: Pakistan and Malawi were mainly affected in 1996
+  + Outdoor: Serbia and Pakistan were the top countries. 
+    - 2011 was the worst for Pakistan. 
+    - 1997 was the worst for Serbia 
+    - Sri Lanka and Tonga increased, but Sri Lanka had a steep decrease after 2015 
+  + Outdoor Ozone: Pakistan and Malawi were the top countries. 
+    - 1997 was the worst for Pakistan. 
+    - 1998 was the worst for Malawi
+    - United States was the second highest amount of deaths among the higher populated countries
+    - Pakistan decreased and then slightly increased
+* Which is worse - outdoor or indoor pollution?
+  + Inconclusive
+
+# Sources
+* [https://www.kaggle.com/datasets/akshat0giri/death-due-to-air-pollution-19902017 ]
+
+* [https://www.worldometers.info/population/world/]
+
+* [https://www.who.int/health-topics/air-pollution#tab=tab_1]
+
+* [https://www.epa.gov/ground-level-ozone-pollution/ground-level-ozone-basics]
+
+* [https://www.health.nsw.gov.au/environment/air/Pages/outdoor-air-pollution.aspx]
+
+* [https://www.mana.md/indoor-air-vs-outdoor-air/]
+
+* [https://www.kaggle.com/datasets/imdevskp/world-population-19602018]

### 2022-05-06 Final Report.Rmd
changed decimals on y-axis

@@
 
+In order to get rid of the leading zeros, and clean up the y-axis, we multiplied the 'percent_high' and 'percent_low' calculation by 10,000,000
+
 ```{r, echo=FALSE, message=FALSE}
@@
 affected_high<- joined_high %>% 
-  mutate(percent_high = total_deaths/mean(Count))
+  mutate(percent_high = (total_deaths/mean(Count))*(1000000))
 
@@
 affected_low <- joined_low %>% 
-  mutate(percent_low = total_deaths/mean(Count))
+  mutate(percent_low = (total_deaths/mean(Count))*(1000000))
 

### 2022-05-06 Final Report.Rmd
changed colnames to rename

@@
 ---
+We were excited to do our report over this data because it was relatively tidy and had quite a few categorical variables and options for additional columns to graph.
 

### 2022-05-06 Final Report.Rmd
renames death df cols

@@
 ```{r}
-deaths_df <- data.frame(read.csv("death-rates-from-air-pollution.csv"))
-glimpse(deaths_df)
+deaths_df_old <- data.frame(read.csv("death-rates-from-air-pollution.csv"))
+glimpse(deaths_df_old)
 ```
+Fixed: use rename instead of colnames
 
@@
 ```{r}
-colnames(deaths_df) <- c("country", "acronym", "year", "total_deaths", "indoor_deaths", "outdoor_deaths", "ozone_deaths")
+deaths_df<- deaths_df_old %>% rename(country=Entity, acronym=Code, year=Year, total_deaths=Air.pollution..total...deaths.per.100.000., indoor_deaths=Indoor.air.pollution..deaths.per.100.000., outdoor_deaths=Outdoor.particulate.matter..deaths.per.100.000., ozone_deaths=Outdoor.ozone.pollution..deaths.per.100.000.)
 

### 2022-05-06 Final Report.Rmd
some changes
popultion graphs

@@
 ***
+
+This is a closer view on the population growth over time in both the high and low populated countries that we selected.
+```{r, echo=FALSE}
+pop_high_growth<- ggplot(data=joined_high, aes(x=year, y=Count, color=country)) +
+  geom_line() +
+  labs(title="Population Over Time in High Populated Countries")+
+  xlab("Year")+
+  ylab("Population")
+ggplotly(pop_high_growth)
+
+pop_low_growth<- ggplot(data=joined_low, aes(x=year, y=Count, color=country)) +
+  geom_line() +
+  labs(title="Population Over Time in Low Populated Countries")+
+  xlab("Year")+
+  ylab("Population")
+ggplotly(pop_low_growth)
+```
 

### 2022-05-06 Final Report.Rmd
clearer explanation of 10%

@@
 
+  - We wanted to have a consistent variation between our selection of low and high populated countries. 
+  - Based on our calculation we selected the low populated country that was closest to 10% of the high populated country.
+  - EX: U.S (high populated) = 331002651. If we multiply by .10 we get 33100265.1 and Canada (low populated) was a population of 37742154 which was closest to 10% of the US population in 2020
 ```{r, echo=FALSE}

### 2022-05-06 Final Report.Rmd
organization

--- 2022-05-06 Final Report.Rmd
@@
 
+  - We wanted to have a consistent variation between our selection of low and high populated countries. 
+  - Based on our calculation we selected the low populated country that was closest to 10% of the high populated country.
+  - EX: U.S (high populated) = 331002651. If we multiply by .10 we get 33100265.1 and Canada (low populated) was a population of 37742154 which was closest to 10% of the US population in 2020
+
 ```{r, echo=FALSE}
@@
 First let's look at a table of the high and low populated countries using the world population data set.
+
 ```{r, echo=FALSE}
@@
 ```
+***
 
 Next, we are going to see the death count for high and low populated countries using the deaths dataframe.
+
 ```{r, echo=FALSE}
@@
 ```
+***
 
 Lastly, we will join the population and and deaths with its respected country.
+
 ```{r, message=FALSE, echo=FALSE}
@@
 This is a closer view on the population growth over time in both the high and low populated countries that we selected.
+
 ```{r, echo=FALSE}
@@
 Here's a graph to clearly visualize the previous table
+
 ```{r, echo=FALSE, message=FALSE}
@@
 h<- ggplot(deaths_highpop_countries)+
-  geom_col(mapping = aes(x=country, y=hp_average_death, color=country))+
+  geom_col(mapping = aes(x=country, y=hp_average_death, fill=country))+
              xlab("Country")+
@@
 l <- ggplot(deaths_lowpop_countries)+
-  geom_col(mapping = aes(x=country, y=lp_average_death, color=country))+
+  geom_col(mapping = aes(x=country, y=lp_average_death, fill=country))+
              xlab("Country")+

### 2022-05-06 Final Report.Rmd
i put head so the html didn't run forever but it only shows one of the two countries we can fix this later

@@
 kable(head(list(joined_high[1:22,], joined_low[1:22,])))
+```
+
+***
+
+Combine the data based on continent.
+
+```{r}
+joined_all <- right_join(deaths_df, world_pop, by=c('country' = 'Country.Name', 'year' = 'Year'))
+head(joined_all)
+
+north_america <- joined_all %>% filter(country %in% c("United States", "Canada"))
+head(na.omit(north_america))
+
+south_america <- joined_all %>% filter(country %in% c("Brazil", "Chile"))
+head(na.omit(south_america))
+
+africa <- joined_all %>% filter(country %in% c("Nigeria", "Malawi"))
+head(na.omit(africa))
+
+europe <- joined_all %>% filter(country %in% c("Germany", "Serbia"))
+head(na.omit(europe))
+
+asia <- joined_all %>% filter(country %in% c("Pakistan", "Sri Lanka"))
+head(na.omit(asia))
+
+oceania <- joined_all %>% filter(country %in% c("Australia", "New Zealand"))
+head(na.omit(oceania))
+
 ```

### 2022-05-06 Final Report.Rmd
tables with labels

@@
 
+joined_select <- full_join(joined_high, joined_low)
+
 kable(head(list(joined_high[1:22,], joined_low[1:22,])))
@@
 kable(list(deaths_highpop_countries, deaths_lowpop_countries))
+
+```
+
+> Let's see how this is different from continent to continent
+
+```{r}
+#Mean total deaths for each continent
+deaths_north <- na.omit(north_america)  %>% 
+  group_by(country) %>% 
+  summarize(north_america_deaths = mean(total_deaths))
+
+deaths_south <- na.omit(south_america)  %>% 
+  group_by(country) %>% 
+  summarize(south_america_deaths = mean(total_deaths))
+
+deaths_africa <- na.omit(africa)  %>% 
+  group_by(country) %>% 
+  summarize(africa_deaths = mean(total_deaths))
+
+deaths_europe <- na.omit(europe)  %>% 
+  group_by(country) %>% 
+  summarize(europe_deaths = mean(total_deaths))
+
+deaths_asia <- na.omit(asia)  %>% 
+  group_by(country) %>% 
+  summarize(asia_deaths = mean(total_deaths))
+
+deaths_oceania <- na.omit(oceania)  %>% 
+  group_by(country) %>% 
+  summarize(oceania_deaths = mean(total_deaths))
+
+#Table to view continent deaths 
+kable(deaths_north, caption = "North America Average Death Count")
+kable(deaths_south, caption = "South America Average Death Count")
+kable(deaths_africa, caption = "Africa Average Death Count")
+kable(deaths_asia, caption = "Asia Average Death Count")
+kable(deaths_europe, caption = "Europe Average Death Count")
+kable(deaths_oceania, caption = "Oceania Average Death Count")
 

### 2022-05-07 Final Report.Rmd
more updates

@@
 #Scientific Notation Disabled 
-options(scipen=999)
+options(scipen=T)
 ```
@@
 
-In order to get rid of the leading zeros, and clean up the y-axis, we multiplied the 'percent_high' and 'percent_low' calculation by 10,000,000
+In order to get rid of the leading zeros, and clean up the y-axis, we multiplied the 'percent_high' and 'percent_low' by 100,000 since the data was per 100,000 when calculating deaths.
 
@@
 affected_high<- joined_high %>% 
-  mutate(percent_high = (total_deaths/mean(Count))*(1000000))
+  mutate(percent_high = (total_deaths/mean(Count))*100000)
 
@@
 affected_low <- joined_low %>% 
-  mutate(percent_low = (total_deaths/mean(Count))*(1000000))
+  mutate(percent_low = (total_deaths/mean(Count))*100000)
 

### 2022-05-07 Final Report.Rmd
nicer plots

@@
 
-kable(low_poll, ascending=TRUE)
-
 kable(low_poll)
@@
 ***
-```{r, echo=FALSE}
+```{r, echo=FALSE, warning=FALSE}
 
@@
 h_outdoor <- ggplot(high_poll, aes(x=country, y = avg_outdoor, color = avg_ozone, cex=5)) +
+  scale_color_gradient2(low = "light pink", mid = "pink", high = "violet", aesthetics = "colour") +
   geom_point() +
@@
 l_outdoor <- ggplot(low_poll, aes(x = country, y= avg_outdoor, color = avg_ozone, cex=5)) +
+  scale_color_gradient2(low = "light pink", mid = "pink", high = "violet", aesthetics = "colour") +
   geom_point() +

### 2022-05-07 Final Report.Rmd
north tables

@@
 ***
+Let's see if there is variation by continent.
+
+```{r}
+#1996-2006
+north_96 <- na.omit(north_america)  %>% 
+  group_by(country) %>% 
+  filter(year == 1996) %>% 
+  summarize(avg_deaths_96 = mean(total_deaths))
+
+north_01 <- na.omit(north_america)  %>% 
+  group_by(country) %>% 
+  filter(year == 2001) %>% 
+  summarize(avg_deaths_01 = mean(total_deaths))
+
+north_06 <- na.omit(north_america)  %>% 
+  group_by(country) %>% 
+  filter(year == 2006) %>% 
+  summarize(avg_deaths_06 = mean(total_deaths))
+
+kable(list(north_96,north_01,north_06), caption = "North America Deaths 1996-2006")
+
+north_07 <- na.omit(north_america)  %>% 
+  group_by(country) %>% 
+  filter(year == 2007) %>% 
+  summarize(avg_deaths_07 = mean(total_deaths))
+
+north_12 <- na.omit(north_america)  %>% 
+  group_by(country) %>% 
+  filter(year == 2012) %>% 
+  summarize(avg_deaths_12 = mean(total_deaths))
+
+north_17 <- na.omit(north_america)  %>% 
+  group_by(country) %>% 
+  filter(year == 2017) %>% 
+  summarize(avg_deaths_17 = mean(total_deaths))
+
+kable(list(north_07,north_12,north_17), caption = "North America Deaths 2007-2017")
+
+deaths_south <- na.omit(south_america)  %>% 
+  group_by(country) %>% 
+  summarize(south_america_deaths = mean(total_deaths))
+
+deaths_africa <- na.omit(africa)  %>% 
+  group_by(country) %>% 
+  summarize(africa_deaths = mean(total_deaths))
+
+deaths_europe <- na.omit(europe)  %>% 
+  group_by(country) %>% 
+  summarize(europe_deaths = mean(total_deaths))
+
+deaths_asia <- na.omit(asia)  %>% 
+  group_by(country) %>% 
+  summarize(asia_deaths = mean(total_deaths))
+
+deaths_oceania <- na.omit(oceania)  %>% 
+  group_by(country) %>% 
+  summarize(oceania_deaths = mean(total_deaths))
+```
+***
 

### 2022-05-07 Final Report.Rmd
more tables

@@
 
-deaths_south <- na.omit(south_america)  %>% 
+south_96 <- na.omit(south_america)  %>% 
   group_by(country) %>% 
-  summarize(south_america_deaths = mean(total_deaths))
+  filter(year == 1996) %>% 
+  summarize(avg_deaths_96 = mean(total_deaths))
 
-deaths_africa <- na.omit(africa)  %>% 
+south_01 <- na.omit(south_america)  %>% 
   group_by(country) %>% 
-  summarize(africa_deaths = mean(total_deaths))
+  filter(year == 2001) %>% 
+  summarize(avg_deaths_01 = mean(total_deaths))
+
+south_06 <- na.omit(south_america)  %>% 
+  group_by(country) %>% 
+  filter(year == 2006) %>% 
+  summarize(avg_deaths_06 = mean(total_deaths))
+
+kable(list(south_96,south_01,south_06), caption = "South America Deaths 1996-2006")
+
+south_07 <- na.omit(south_america)  %>% 
+  group_by(country) %>% 
+  filter(year == 2007) %>% 
+  summarize(avg_deaths_07 = mean(total_deaths))
+
+south_12 <- na.omit(south_america)  %>% 
+  group_by(country) %>% 
+  filter(year == 2012) %>% 
+  summarize(avg_deaths_12 = mean(total_deaths))
+
+south_17 <- na.omit(south_america)  %>% 
+  group_by(country) %>% 
+  filter(year == 2017) %>% 
+  summarize(avg_deaths_17 = mean(total_deaths))
+
+kable(list(south_07,south_12,south_17), caption = "South America Deaths 2007-2017")
+
+africa_96 <- na.omit(africa)  %>% 
+  group_by(country) %>% 
+  filter(year == 1996) %>% 
+  summarize(avg_deaths_96 = mean(total_deaths))
+
+africa_01 <- na.omit(africa)  %>% 
+  group_by(country) %>% 
+  filter(year == 2001) %>% 
+  summarize(avg_deaths_01 = mean(total_deaths))
+
+africa_06 <- na.omit(africa)  %>% 
+  group_by(country) %>% 
+  filter(year == 2006) %>% 
+  summarize(avg_deaths_06 = mean(total_deaths))
+
+kable(list(africa_96,africa_01,africa_06), caption = "Africa Deaths 1996-2006")
+
+africa_07 <- na.omit(africa)  %>% 
+  group_by(country) %>% 
+  filter(year == 2007) %>% 
+  summarize(avg_deaths_07 = mean(total_deaths))
+
+africa_12 <- na.omit(africa)  %>% 
+  group_by(country) %>% 
+  filter(year == 2012) %>% 
+  summarize(avg_deaths_12 = mean(total_deaths))
+
+africa_17 <- na.omit(africa)  %>% 
+  group_by(country) %>% 
+  filter(year == 2017) %>% 
+  summarize(avg_deaths_17 = mean(total_deaths))
+
+kable(list(africa_07,africa_12,africa_17), caption = "Africa Deaths 2007-2017")
 

### 2022-05-07 Final Report.Rmd
second half of tables

--- 2022-05-07 Final Report.Rmd
@@
 
+#Europe 1996-2006
+europe_96 <- na.omit(europe)  %>% 
+  group_by(country) %>% 
+  filter(year == 1996) %>% 
+  summarize(avg_deaths_96 = mean(total_deaths))
+
+europe_01 <- na.omit(europe)  %>% 
+  group_by(country) %>% 
+  filter(year == 2001) %>% 
+  summarize(avg_deaths_01 = mean(total_deaths))
+
+europe_06 <- na.omit(europe)  %>% 
+  group_by(country) %>% 
+  filter(year == 2006) %>% 
+  summarize(avg_deaths_06 = mean(total_deaths))
+
+kable(list(europe_96,europe_01,europe_06), caption = "Europe Deaths 1996-2006")
+
+europe_07 <- na.omit(europe)  %>% 
+  group_by(country) %>% 
+  filter(year == 2007) %>% 
+  summarize(avg_deaths_07 = mean(total_deaths))
+
+europe_12 <- na.omit(europe)  %>% 
+  group_by(country) %>% 
+  filter(year == 2012) %>% 
+  summarize(avg_deaths_12 = mean(total_deaths))
+
+europe_17 <- na.omit(europe)  %>% 
+  group_by(country) %>% 
+  filter(year == 2017) %>% 
+  summarize(avg_deaths_17 = mean(total_deaths))
+
+kable(list(europe_07,europe_12,europe_17), caption = "Europe Deaths 2007-2017")
+
+#Asia 1996-2006
+asia_96 <- na.omit(asia)  %>% 
+  group_by(country) %>% 
+  filter(year == 1996) %>% 
+  summarize(avg_deaths_96 = mean(total_deaths))
+
+asia_01 <- na.omit(asia)  %>% 
+  group_by(country) %>% 
+  filter(year == 2001) %>% 
+  summarize(avg_deaths_01 = mean(total_deaths))
+
+asia_06 <- na.omit(asia)  %>% 
+  group_by(country) %>% 
+  filter(year == 2006) %>% 
+  summarize(avg_deaths_06 = mean(total_deaths))
+
+kable(list(asia_96,asia_01,asia_06), caption = "Asia Deaths 1996-2006")
+
+asia_07 <- na.omit(asia)  %>% 
+  group_by(country) %>% 
+  filter(year == 2007) %>% 
+  summarize(avg_deaths_07 = mean(total_deaths))
+
+asia_12 <- na.omit(asia)  %>% 
+  group_by(country) %>% 
+  filter(year == 2012) %>% 
+  summarize(avg_deaths_12 = mean(total_deaths))
+
+asia_17 <- na.omit(asia)  %>% 
+  group_by(country) %>% 
+  filter(year == 2017) %>% 
+  summarize(avg_deaths_17 = mean(total_deaths))
+
+kable(list(asia_07,asia_12,asia_17), caption = "Asia Deaths 2007-2017")
+
+#Oceania 1996-2006
+oceania_96 <- na.omit(oceania)  %>% 
+  group_by(country) %>% 
+  filter(year == 1996) %>% 
+  summarize(avg_deaths_96 = mean(total_deaths))
+
+oceania_01 <- na.omit(oceania)  %>% 
+  group_by(country) %>% 
+  filter(year == 2001) %>% 
+  summarize(avg_deaths_01 = mean(total_deaths))
+
+oceania_06 <- na.omit(oceania)  %>% 
+  group_by(country) %>% 
+  filter(year == 2006) %>% 
+  summarize(avg_deaths_06 = mean(total_deaths))
+
+kable(list(oceania_96,oceania_01,oceania_06), caption = "Oceania Deaths 1996-2006")
+
+oceania_07 <- na.omit(oceania)  %>% 
+  group_by(country) %>% 
+  filter(year == 2007) %>% 
+  summarize(avg_deaths_07 = mean(total_deaths))
+
+oceania_12 <- na.omit(oceania)  %>% 
+  group_by(country) %>% 
+  filter(year == 2012) %>% 
+  summarize(avg_deaths_12 = mean(total_deaths))
+
+oceania_17 <- na.omit(oceania)  %>% 
+  group_by(country) %>% 
+  filter(year == 2017) %>% 
+  summarize(avg_deaths_17 = mean(total_deaths))
+
+kable(list(oceania_07,oceania_12,oceania_17), caption = "Asia Deaths 2007-2017")
+
 deaths_south <- na.omit(south_america)  %>% 

### 2022-05-07 Final Report.Rmd
changes

@@
 ```{r}
-#1996-2006
+#North America 1996-2006
 north_96 <- na.omit(north_america)  %>% 
@@
 kable(list(north_07,north_12,north_17), caption = "North America Deaths 2007-2017")
+
+#South America
+south_96 <- na.omit(south_america)  %>% 
+  group_by(country) %>% 
+  filter(year == 1996) %>% 
+  summarize(avg_deaths_96 = mean(total_deaths))
+
+south_01 <- na.omit(south_america)  %>% 
+  group_by(country) %>% 
+  filter(year == 2001) %>% 
+  summarize(avg_deaths_01 = mean(total_deaths))
+
+south_06 <- na.omit(south_america)  %>% 
+  group_by(country) %>% 
+  filter(year == 2006) %>% 
+  summarize(avg_deaths_06 = mean(total_deaths))
 
@@
 
-deaths_south <- na.omit(south_america)  %>% 
+=======
+kable(list(south_96,south_01,south_06), caption = "South America Deaths 1996-2006")
+
+south_07 <- na.omit(south_america)  %>% 
   group_by(country) %>% 
-  summarize(south_america_deaths = mean(total_deaths))
+  filter(year == 2007) %>% 
+  summarize(avg_deaths_07 = mean(total_deaths))
+>>>>>>> b277bd1eb5761b57f0819d4f702030583bbe3739
 
-deaths_africa <- na.omit(africa)  %>% 
+south_12 <- na.omit(south_america)  %>% 
   group_by(country) %>% 
-  summarize(africa_deaths = mean(total_deaths))
+  filter(year == 2012) %>% 
+  summarize(avg_deaths_12 = mean(total_deaths))
+
+south_17 <- na.omit(south_america)  %>% 
+  group_by(country) %>% 
+  filter(year == 2017) %>% 
+  summarize(avg_deaths_17 = mean(total_deaths))
+
+kable(list(south_07,south_12,south_17), caption = "South America Deaths 2007-2017")
+
+africa_96 <- na.omit(africa)  %>% 
+  group_by(country) %>% 
+  filter(year == 1996) %>% 
+  summarize(avg_deaths_96 = mean(total_deaths))
+
+africa_01 <- na.omit(africa)  %>% 
+  group_by(country) %>% 
+  filter(year == 2001) %>% 
+  summarize(avg_deaths_01 = mean(total_deaths))
+
+africa_06 <- na.omit(africa)  %>% 
+  group_by(country) %>% 
+  filter(year == 2006) %>% 
+  summarize(avg_deaths_06 = mean(total_deaths))
+
+kable(list(africa_96,africa_01,africa_06), caption = "Africa Deaths 1996-2006")
+
+africa_07 <- na.omit(africa)  %>% 
+  group_by(country) %>% 
+  filter(year == 2007) %>% 
+  summarize(avg_deaths_07 = mean(total_deaths))
+
+africa_12 <- na.omit(africa)  %>% 
+  group_by(country) %>% 
+  filter(year == 2012) %>% 
+  summarize(avg_deaths_12 = mean(total_deaths))
+
+africa_17 <- na.omit(africa)  %>% 
+  group_by(country) %>% 
+  filter(year == 2017) %>% 
+  summarize(avg_deaths_17 = mean(total_deaths))
+
+kable(list(africa_07,africa_12,africa_17), caption = "Africa Deaths 2007-2017")
 

### 2022-05-07 Final Report.Rmd
took out weird text

@@
 ```{r}
-#North America 1996-2006
+#1996-2006
 north_96 <- na.omit(north_america)  %>% 
@@
 
-#South America
 south_96 <- na.omit(south_america)  %>% 
@@
 
-=======
 kable(list(south_96,south_01,south_06), caption = "South America Deaths 1996-2006")
@@
   summarize(avg_deaths_07 = mean(total_deaths))
->>>>>>> b277bd1eb5761b57f0819d4f702030583bbe3739
 
@@
 kable(list(africa_07,africa_12,africa_17), caption = "Africa Deaths 2007-2017")
-
-deaths_europe <- na.omit(europe)  %>% 
-  group_by(country) %>% 
-  summarize(europe_deaths = mean(total_deaths))
-
-deaths_asia <- na.omit(asia)  %>% 
-  group_by(country) %>% 
-  summarize(asia_deaths = mean(total_deaths))
-
-deaths_oceania <- na.omit(oceania)  %>% 
-  group_by(country) %>% 
-  summarize(oceania_deaths = mean(total_deaths))
 ```

### 2022-05-07 Final Report.Rmd

@@
 
+#South America
 south_96 <- na.omit(south_america)  %>% 

### 2022-05-07 Final Report.Rmd
all tables

@@
 
+# North America 2007-2017
 north_07 <- na.omit(north_america)  %>% 
@@
 
-#South America
+#South America 1996-2006
 south_96 <- na.omit(south_america)  %>% 
@@
 
+kable(list(south_96,south_01,south_06), caption = "South America Deaths 1996-2006")
+
+# South America 2007-2017
+south_07 <- na.omit(south_america)  %>% 
+  group_by(country) %>% 
+  filter(year == 2007) %>% 
+  summarize(avg_deaths_07 = mean(total_deaths))
+
+south_12 <- na.omit(south_america)  %>% 
+  group_by(country) %>% 
+  filter(year == 2012) %>% 
+  summarize(avg_deaths_12 = mean(total_deaths))
+
+south_17 <- na.omit(south_america)  %>% 
+  group_by(country) %>% 
+  filter(year == 2017) %>% 
+  summarize(avg_deaths_17 = mean(total_deaths))
+
+kable(list(south_07,south_12,south_17), caption = "South America Deaths 2007-2017")
+
+# Africa 1996-2006
+africa_96 <- na.omit(africa)  %>% 
+  group_by(country) %>% 
+  filter(year == 1996) %>% 
+  summarize(avg_deaths_96 = mean(total_deaths))
+
+africa_01 <- na.omit(africa)  %>% 
+  group_by(country) %>% 
+  filter(year == 2001) %>% 
+  summarize(avg_deaths_01 = mean(total_deaths))
+
+africa_06 <- na.omit(africa)  %>% 
+  group_by(country) %>% 
+  filter(year == 2006) %>% 
+  summarize(avg_deaths_06 = mean(total_deaths))
+
+kable(list(africa_96,africa_01,africa_06), caption = "Africa Deaths 1996-2006")
+
+# Africa 2007-2017
+africa_07 <- na.omit(africa)  %>% 
+  group_by(country) %>% 
+  filter(year == 2007) %>% 
+  summarize(avg_deaths_07 = mean(total_deaths))
+
+africa_12 <- na.omit(africa)  %>% 
+  group_by(country) %>% 
+  filter(year == 2012) %>% 
+  summarize(avg_deaths_12 = mean(total_deaths))
+
+africa_17 <- na.omit(africa)  %>% 
+  group_by(country) %>% 
+  filter(year == 2017) %>% 
+  summarize(avg_deaths_17 = mean(total_deaths))
+
+kable(list(africa_07,africa_12,africa_17), caption = "Africa Deaths 2007-2017")
+
 #Europe 1996-2006
@@
 
+#Europe 2007-2017
 europe_07 <- na.omit(europe)  %>% 
@@
 
+#Asia 2007-2017
 asia_07 <- na.omit(asia)  %>% 
@@
 
+#Oceania 2007-2017
 oceania_07 <- na.omit(oceania)  %>% 
@@
 
-kable(list(oceania_07,oceania_12,oceania_17), caption = "Asia Deaths 2007-2017")
-
-kable(list(south_96,south_01,south_06), caption = "South America Deaths 1996-2006")
-
-south_07 <- na.omit(south_america)  %>% 
-  group_by(country) %>% 
-  filter(year == 2007) %>% 
-  summarize(avg_deaths_07 = mean(total_deaths))
-
-south_12 <- na.omit(south_america)  %>% 
-  group_by(country) %>% 
-  filter(year == 2012) %>% 
-  summarize(avg_deaths_12 = mean(total_deaths))
-
-south_17 <- na.omit(south_america)  %>% 
-  group_by(country) %>% 
-  filter(year == 2017) %>% 
-  summarize(avg_deaths_17 = mean(total_deaths))
-
-kable(list(south_07,south_12,south_17), caption = "South America Deaths 2007-2017")
-
-africa_96 <- na.omit(africa)  %>% 
-  group_by(country) %>% 
-  filter(year == 1996) %>% 
-  summarize(avg_deaths_96 = mean(total_deaths))
-
-africa_01 <- na.omit(africa)  %>% 
-  group_by(country) %>% 
-  filter(year == 2001) %>% 
-  summarize(avg_deaths_01 = mean(total_deaths))
-
-africa_06 <- na.omit(africa)  %>% 
-  group_by(country) %>% 
-  filter(year == 2006) %>% 
-  summarize(avg_deaths_06 = mean(total_deaths))
-
-kable(list(africa_96,africa_01,africa_06), caption = "Africa Deaths 1996-2006")
-
-africa_07 <- na.omit(africa)  %>% 
-  group_by(country) %>% 
-  filter(year == 2007) %>% 
-  summarize(avg_deaths_07 = mean(total_deaths))
-
-africa_12 <- na.omit(africa)  %>% 
-  group_by(country) %>% 
-  filter(year == 2012) %>% 
-  summarize(avg_deaths_12 = mean(total_deaths))
-
-africa_17 <- na.omit(africa)  %>% 
-  group_by(country) %>% 
-  filter(year == 2017) %>% 
-  summarize(avg_deaths_17 = mean(total_deaths))
-
-kable(list(africa_07,africa_12,africa_17), caption = "Africa Deaths 2007-2017")
+kable(list(oceania_07,oceania_12,oceania_17), caption = "Oceania Deaths 2007-2017")
 ```

### 2022-05-07 Final Report.Rmd
nice colors

@@
 Let's see if there is variation by continent.
+Here are some tables for the first decade (1996-2006) and second decade (2007-2017) grouped by continent.
 
@@
   ggplot(aes(x= High_Deaths_96, y= High_Deaths_01, color= High_Deaths_06)) +
+  scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
   geom_point() +
@@
   ggplot(aes(x= Low_Deaths_96, y= Low_Deaths_01, color= Low_Deaths_06)) +
+  scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
   geom_point() +
@@
 ggplotly(interact_low_first)
+
+#Joined all 
+#X - year 
+# Avg total deaths 
+# color by continent 
 ```
@@
   ggplot(aes(x= High_Deaths_07, y= High_Deaths_12, color= High_Deaths_17)) +
+  scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
   geom_point() +
@@
   ggplot(aes(x= Low_Deaths_07, y= Low_Deaths_12, color= Low_Deaths_17)) +
+  scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
   geom_point() +

### 2022-05-09 Final Report.Rmd
code

@@
 ```{r, echo=FALSE}
-#High Population Deaths 1996-2006 
+#High Population Deaths 1996 
 p_high_first <- all_high_first %>%
   group_by(country) %>% 
-  ggplot(aes(x= High_Deaths_96, y= High_Deaths_01, color= High_Deaths_06)) +
+  ggplot(aes(x= country, y= High_Deaths_96, fill = country)) +
   scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
-  geom_point() +
-  facet_wrap(~country)
-
-interact_high_first<- p_high_first + labs(title="High Populated Countries", x="1996 Deaths", y="2001 Deaths")
+  geom_col(position = 'dodge')
+
+interact_high_first<- p_high_first + labs(title="High Populated Countries", x="1996 Deaths")
 
 ggplotly(interact_high_first)
+
+#High Population Deaths 1996
 

### 2022-05-09 Final Report.Rmd
first decade done

--- 2022-05-07 Final Report.Rmd
@@
 
+#High Population Deaths 1996 
+p_high_first <- all_high_first %>%
+  group_by(country) %>% 
+  ggplot(aes(x= country, y= High_Deaths_96, fill = country)) +
+  scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
+  geom_col(position = 'dodge')
+
+interact_high_first<- p_high_first + labs(title="High Populated Countries", x="1996 Deaths")
+
+ggplotly(interact_high_first)
+
+#High Population Deaths 2001
+p_high_first <- all_high_first %>%
+  group_by(country) %>% 
+  ggplot(aes(x= country, y= High_Deaths_01, fill = country)) +
+  scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
+  geom_col(position = 'dodge')
+
+interact_high_first<- p_high_first + labs(title="High Populated Countries", x="2001 Deaths")
+
+ggplotly(interact_high_first)
+
+#High Population Deaths 2006
+p_high_first <- all_high_first %>%
+  group_by(country) %>% 
+  ggplot(aes(x= country, y= High_Deaths_06, fill = country)) +
+  scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
+  geom_col(position = 'dodge')
+
+interact_high_first<- p_high_first + labs(title="High Populated Countries", x="2006 Deaths")
+
+ggplotly(interact_high_first)
+
 #Low Population Deaths 1996-2006
@@
 
-#Joined all 
-#X - year 
-# Avg total deaths 
-# color by continent 
+#Low Population Deaths 1996 
+p_low_first <- all_low_first %>%
+  group_by(country) %>% 
+  ggplot(aes(x= country, y= Low_Deaths_96, fill = country)) +
+  scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
+  geom_col(position = 'dodge')
+
+interact_low_first<- p_low_first + labs(title="High Populated Countries", x="1996 Deaths")
+
+ggplotly(interact_low_first)
+
+#Low Population Deaths 2001 
+p_low_first <- all_low_first %>%
+  group_by(country) %>% 
+  ggplot(aes(x= country, y= Low_Deaths_01, fill = country)) +
+  scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
+  geom_col(position = 'dodge')
+
+interact_low_first<- p_low_first + labs(title="High Populated Countries", x="1996 Deaths")
+
+ggplotly(interact_low_first)
+
+#Low Population Deaths 2006 
+p_low_first <- all_low_first %>%
+  group_by(country) %>% 
+  ggplot(aes(x= country, y= Low_Deaths_06, fill = country)) +
+  scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
+  geom_col(position = 'dodge')
+
+interact_low_first<- p_low_first + labs(title="High Populated Countries", x="1996 Deaths")
+
+ggplotly(interact_low_first)
 ```

### 2022-05-09 Final Report.Rmd
second decade continent graphs

--- 2022-05-09 Final Report.Rmd
@@
 ```{r, echo=FALSE}
-#High Population Deaths 2007-2017
-p_high_second <- all_high_second %>%
-  group_by(country) %>% 
-  ggplot(aes(x= High_Deaths_07, y= High_Deaths_12, color= High_Deaths_17)) +
+#High Population Deaths 2007
+p_high_second_07 <- all_high_second %>%
+  group_by(country) %>% 
+  ggplot(aes(x= country, y= High_Deaths_07, fill=country)) +
   scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
-  geom_point() +
-  facet_wrap(~country)
-
-interact_high_second<- p_high_second + labs(title="High Populated Countries", x="2007 Deaths", y="2012 Deaths")
-
-ggplotly(interact_high_second)
-
-#Low Population Deaths 2007-2017
-p_low_second <- all_low_second %>%
-  group_by(country) %>% 
-  ggplot(aes(x= Low_Deaths_07, y= Low_Deaths_12, color= Low_Deaths_17)) +
+  geom_col(position = 'dodge')
+
+interact_high_second_07<- p_high_second_07 + labs(title="High Populated Countries", x="2007 Deaths")
+
+ggplotly(interact_high_second_07)
+
+#High Population Deaths 2012
+p_high_second_12 <- all_high_second %>%
+  group_by(country) %>% 
+  ggplot(aes(x= country, y= High_Deaths_12, fill=country)) +
   scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
-  geom_point() +
-  facet_wrap(~country)
-
-interact_low_second<- p_low_second + labs(title="Low Populated Countries", x="2007 Deaths", y="2012 Deaths")
-
-ggplotly(interact_low_second)
+  geom_col(position = 'dodge')
+
+interact_high_second_12<- p_high_second_12 + labs(title="High Populated Countries", x="2012 Deaths")
+
+ggplotly(interact_high_second_12)
+
+#High Population Deaths 2017
+p_high_second_17 <- all_high_second %>%
+  group_by(country) %>% 
+  ggplot(aes(x= country, y= High_Deaths_17, fill=country)) +
+  scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
+  geom_col(position = 'dodge')
+
+interact_high_second_17<- p_high_second_17 + labs(title="High Populated Countries", x="2017 Deaths")
+
+ggplotly(interact_high_second_17)
+
+#Low Population Deaths 2007
+p_low_second_07 <- all_low_second %>%
+  group_by(country) %>% 
+  ggplot(aes(x= country, y= Low_Deaths_07, fill=country)) +
+  scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
+  geom_col(position='dodge')
+
+interact_low_second_07<- p_low_second_07 + labs(title="Low Populated Countries", x="2007 Deaths")
+
+ggplotly(interact_low_second_07)
+
+#Low Population Deaths 2012
+p_low_second_12 <- all_low_second %>%
+  group_by(country) %>% 
+  ggplot(aes(x= country, y= Low_Deaths_12, fill=country)) +
+  scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
+  geom_col(position='dodge')
+
+interact_low_second_12<- p_low_second_12 + labs(title="Low Populated Countries", x="2012 Deaths")
+
+ggplotly(interact_low_second_12)
+
+#Low Population Deaths 2017
+p_low_second_17 <- all_low_second %>%
+  group_by(country) %>% 
+  ggplot(aes(x= country, y= Low_Deaths_17, fill=country)) +
+  scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
+  geom_col(position='dodge')
+
+interact_low_second_17<- p_low_second_17 + labs(title="Low Populated Countries", x="2017 Deaths")
+
+ggplotly(interact_low_second_17)
 

### 2022-05-09 Final Report.Rmd

@@
 ```{r, echo=FALSE}
+#High Population Deaths 1996-2006
+p_high_first <- all_high_first %>%
+  group_by(country) %>% 
+  ggplot(aes(x= High_Deaths_96, y= High_Deaths_01, color= High_Deaths_06)) +
+  scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
+  geom_point() +
+  facet_wrap(~country)
+
+interact_high_first<- p_high_first + labs(title="High Populated Countries", x="1996 Deaths", y="2001 Deaths")
+
+ggplotly(interact_high_first)
+
 #High Population Deaths 1996 
@@
 
-#High Population Deaths 1996
+#High Population Deaths 2001
+p_high_first <- all_high_first %>%
+  group_by(country) %>% 
+  ggplot(aes(x= country, y= High_Deaths_01, fill = country)) +
+  scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
+  geom_col(position = 'dodge')
+
+interact_high_first<- p_high_first + labs(title="High Populated Countries", x="2001 Deaths")
+
+ggplotly(interact_high_first)
+
+#High Population Deaths 2006
+p_high_first <- all_high_first %>%
+  group_by(country) %>% 
+  ggplot(aes(x= country, y= High_Deaths_06, fill = country)) +
+  scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
+  geom_col(position = 'dodge')
+
+interact_high_first<- p_high_first + labs(title="High Populated Countries", x="2006 Deaths")
+
+ggplotly(interact_high_first)
 

### 2022-05-09 Final Report.Rmd
second decade stuff

@@
 ```{r, echo=FALSE}
+#High Population Deaths 2007-2017
+p_high_second <- all_high_second %>%
+  group_by(country) %>% 
+  ggplot(aes(x= High_Deaths_07, y= High_Deaths_12, color= High_Deaths_17)) +
+  scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
+  geom_point() +
+  facet_wrap(~country)
+
+interact_high_second<- p_high_second + labs(title="High Populated Countries", x="2007 Deaths", y="2017 Deaths")
+
+ggplotly(interact_high_second)
+
 #High Population Deaths 2007
@@
 ggplotly(interact_high_second_17)
+
+#Low Population Deaths 2007-2017
+p_low_second <- all_low_second %>%
+  group_by(country) %>% 
+  ggplot(aes(x= Low_Deaths_07, y= Low_Deaths_12, color= Low_Deaths_17)) +
+  scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
+  geom_point() +
+  facet_wrap(~country)
+
+interact_low_second<- p_low_second + labs(title="Low Populated Countries", x="2007 Deaths", y="2017 Deaths")
+
+ggplotly(interact_low_second)
 

### 2022-05-09 Final Report.Rmd
gradient

@@
   geom_point() +
+  scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
   ggtitle("High Population Pollutant Type Death Counts")
@@
   geom_point() +
+  scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
   ggtitle("Low Population Pollutant Type Death Counts")

### 2022-05-09 Final Report.Rmd
intro

@@
 ---
-We were excited to do our report over this data because it was relatively tidy and had quite a few categorical variables and options for additional columns to graph.
+In our initial discussion over choosing a topic for our project, we narrowed it down to environmental-related data because we were interested in seeing possible trends over time and the vast quantity of environmental data is available.
 
@@
 # Deaths Data
+We were excited to do our report over this data because it was relatively tidy and had quite a few categorical variables and options for additional columns to graph.
+
 * Deaths from pollution:

### 2022-05-09 Final Report.Rmd
explanations

@@
 
+Our deaths due to air pollution data set was from Kaggle.
+The author is Akshat Giri and was last updated 2 years ago so it’s pretty relevant.
+When we first loaded in the data some of the column names were lengthy so we shortened them to: country, acronym, year, total deaths, indoor deaths, outdoor deaths, and ozone deaths.
+
 Import the deaths-due-to-air-pollution data
@@
 ```
-Fixed: use rename instead of colnames
 
@@
   + Charcoal, coal
-* __outdoor_deaths__: Outdoor air or ambient air are emissions caused by combustion processes from motor vehicles, solid fuel burning and industry 
+  This method of cooking is commonly seen in underdeveloped countries.
+* __outdoor_deaths__: Outdoor air or ambient air are emissions caused by combustion processes from motor vehicles, solid-fuel burning and industries
   + Ozone (O3)
@@
   + Sulfur dioxide (SO2)
+
+The data set takes a closer look at deaths caused by the ozone itself which is considered a component of outdoor air pollution.
+
 * __ozone_deaths__: Ozone is a gas that occurs both in Earth's upper atmosphere and at ground level.
@@
   + Last update: 2 Years ago
+
+The world population data set is also from Kaggel.
+The author is Devakumar K.
+P.
+and was last updated 2 years ago so it is also recent.
+From looking at a glimpse of the data set you can see the columns are country name, year, and count which refers to the population at that time
 

### 2022-05-10 Final Report.Rmd
next slides

@@
 ***
-We selected a high population from each continent and used the formula below to determine the low population.
-
-Low population = high population * .10
-
-  - We wanted to have a consistent variation between our selection of low and high populated countries. 
-  - Based on our calculation we selected the low populated country that was closest to 10% of the high populated country.
-  - EX: U.S (high populated) = 331002651. If we multiply by .10 we get 33100265.1 and Canada (low populated) was a population of 37742154 which was closest to 10% of the US population in 2020
+We selected a high-population and a low-population country from each continent, but we wanted a consistent variation between our selection of low and high population countries.
+So, we came up with a formula for calculating what the low-population country should be by multiplying the high-population country by .10.
+For example, when we chose the U.S (which had a population of 331002651 at the time the data was recorded), we multiplied this number by .10 to get 33100265.1 and find the country whose population most-closesly matched (in this case it was Canada with 37742154).
+
+We purposefully left out countries whose population numbers were higher than the majority because we didn't want those countries to skew the data (Russia, India, and China).
 
@@
 
+Continents:
+
+North America: U.S, Canada
+South America: Brazil, Chile
+Africa: Nigeria, Malawi
+Europe: Germany, Serbia
+Asia: Pakistan, Sri Lanka
+Oceania: Australia, New Zealand
+
 # Combine Data Sets
@@
 
-Combine the data based on continent.
+We also looked at how the data varied by continent.
 
@@
 
+This graph shows the population change over time.
+Something to note is Germany and Australia's line seem relatively flat, but a closer look will determine there is a gradual increase that is drastically slower than the other countries.
+
 ```{r, echo=FALSE}
@@
 ```
+These graphs are of the same information but we have added the percentage of air-pollution related deaths as the width of the line to demonstrate visually if deaths increased or decreased over time.
+It is easier to see in the high-populated countries but air-pollution related deaths do decrease over time.
 

### 2022-05-10 Final Report.Rmd
more explanations

@@
 ```
+We wanted to take a closer look at the death count and see which country has the highest average death count.
+In the tables we made you can see that Pakistan had the highest average death count at 144.33 for the high populated countries.
+Malawi had the highest average death count at 147.77 for the low populated countries, which is higher than Pakistan.
 
@@
 ```
+
+When we look at the average death count based on continent we can see that overall Oceania countries had the least amount of deaths.
+On average Australia had an average death count of roughly 17.8 and New Zealand had an average death count of 15.9.
+Whereas Africa countries had the most amount of deaths.
+On average Malawi had an average of 147.8 and Nigeria had an average of 112.3.
 
@@
 
+To get a better visualization we created a bar graph of the average deaths in both the high and low populated countries.
+In this high-population graph you can see that Pakistan is at the highest and Australia is at the lowest.
+In the low-population graph you can see that Malawi is at the highest and New Zealand is at the lowest.
+
 *** 
@@
 ```
+
+So now that we’ve looked at the deaths due to pollution we wanted to see what percentage of the population was actually affected by this.
+At the top we have a table depicting the average populations in both the high and low populated countries.
+You can see for the high populated countries that Pakistan is in the lead with 12.1% and for the low populated countries Malawi is in the lead with 130.9%.
 

### 2022-05-10 Final Report.Rmd
graph explanations

@@
 
-North America: U.S, Canada
-South America: Brazil, Chile
-Africa: Nigeria, Malawi
-Europe: Germany, Serbia
-Asia: Pakistan, Sri Lanka
-Oceania: Australia, New Zealand
+  + North America: U.S, Canada
+  + South America: Brazil, Chile
+  + Africa: Nigeria, Malawi
+  + Europe: Germany, Serbia
+  + Asia: Pakistan, Sri Lanka
+  + Oceania: Australia, New Zealand
 
@@
 
-Lastly, we will join the population and and deaths with its respected country.
+Lastly, we will join the population and and deaths with its respected country. The last column displays the 'Count' which represents the population of that country at that time period.
 
@@
 head(joined_all)
-
+```
+
+North America
+```{r}
 north_america <- joined_all %>% filter(country %in% c("United States", "Canada"))
 head(na.omit(north_america))
-
+```
+
+South America 
+```{r}
 south_america <- joined_all %>% filter(country %in% c("Brazil", "Chile"))
 head(na.omit(south_america))
-
+```
+
+Africa
+```{r}
 africa <- joined_all %>% filter(country %in% c("Nigeria", "Malawi"))
 head(na.omit(africa))
-
+```
+
+Europe
+```{r}
 europe <- joined_all %>% filter(country %in% c("Germany", "Serbia"))
 head(na.omit(europe))
-
+```
+
+Asia
+```{r}
 asia <- joined_all %>% filter(country %in% c("Pakistan", "Sri Lanka"))
 head(na.omit(asia))
-
+```
+
+Oceania
+```{r}
 oceania <- joined_all %>% filter(country %in% c("Australia", "New Zealand"))
@@
 ```
+
 These graphs are of the same information but we have added the percentage of air-pollution related deaths as the width of the line to demonstrate visually if deaths increased or decreased over time.
@@
 ```
+
 We wanted to take a closer look at the death count and see which country has the highest average death count.
@@
 
+Looking between indoor, outdoor, and ozone pollution deaths, we can see which pollutant-type had the greatest death count.
+
 ```{r, echo=FALSE, message=FALSE}
@@
 kable(high_poll)
-
+```
+The average indoor-death count is higher than other pollutant types, especially in Pakistan and Nigeria. In all three categories, Pakistan had the highest death count.
+
+```{r}
 #Low Population Pollutant Averages 
@@
 ```
+
+In the low populated countries, Malawi has a higher average indoor death than that of any of our other selected countries. It is also the highest in all three categories, as well.
 
@@
 ggplotly(l_indoor)
-
+```
+In these graphs, it is much easier to see the discrepancies in the indoor deaths of the countries. Malawi, Pakistan, and Nigeria have dots located high on the graph indicating high death counts.
+
+```{r}
 # High Outdoor Air Pollution
@@
 ```
+
+In these graphs, the outdoor pollution deaths are displayed on the y-axis with the amount of ozone-related deaths indicated by the color gradient.
+The dots highest-up on the graph have high outdoor-related deaths, and the darker the color of pink indicates a high ozone-related death.
+
+For our high-populated countries, Pakistan had high outdoor-deaths and ozone-deaths while the other high-populated countries were lower in ozone-deaths.
+In the low-populated countries, Serbia had the highest outdoor and ozone deaths, but Malawi in particular had the greatest ozone deaths.
 

### 2022-05-10 Final Report.Rmd
over time info

@@
 
+To see if there's been a change over time we looked at 2 decades.
+The first decade was from 1996-2006.
+Here we can see that there’s a general decrease in both high and low populated countries over the years even though some countries have a higher death count than others, such as Nigeria, Pakistan, Malawi, and Tonga.
+
 This is the first decade 1996-2006
@@
 
+The second decade was from 2007-2017 and there is still a decrease in both high and low populated countries over the years.
+You can also see that the death counts in the second decade are lower than the death counts in the first decade.
+If you look at Pakistan you can see the death count started with 155.4 in 1996 and went down to 143.8 in 2007.
+
 This is the second decade 2007-2017
@@
 ```
+
+Based on these tables, we can see that Oceania countries had the lowest death counts over time across both decades.
+In Europe, although Germany is a high-populated country its death count was significantly less than Serbia, a low-populated country.
+This was similarly seen in Africa, where Malawi, a low-populated country, had a higher death count overall in comparison to Nigeria, a high-populated country.
+In Africa, there was the greatest amount of decrease in their average death count.
+Malawi decreased roughly 4,600,000 from 1996 to 2006.
+
 ***
@@
 
-The first decade 1996-2006.
+In the first graph we faceted the high-populated countries and look at the average death count in 1996 on the x-axis, the average death count in 2001 on the y-axis, and color the points by the average death count in 2006.
+The darker the points are the higher the death count was in 2006.
+To get a closer look at the average death count in all 3 years we created individualized bar graphs for 1996, 2001, and 2006.
+
+High Population first decade (1996-2006).
 ```{r, echo=FALSE}
@@
   scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
+   xlab("Country")+
+  ylab("1996 Deaths")+
   geom_col(position = 'dodge')
 
-interact_high_first<- p_high_first + labs(title="High Populated Countries", x="1996 Deaths")
+interact_high_first<- p_high_first + labs(title="High Populated Countries")
 
@@
   scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
+  xlab("Country")+
+  ylab("2001 Deaths")+
   geom_col(position = 'dodge')
 
-interact_high_first<- p_high_first + labs(title="High Populated Countries", x="2001 Deaths")
+interact_high_first<- p_high_first + labs(title="High Populated Countries")
 
@@
   scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
+  xlab("Country")+
+  ylab("2006 Deaths")+
   geom_col(position = 'dodge')
 
-interact_high_first<- p_high_first + labs(title="High Populated Countries", x="2006 Deaths")
+interact_high_first<- p_high_first + labs(title="High Populated Countries")
 
 ggplotly(interact_high_first)
-
+```
+
+***
+
+In the first graph we faceted the low-populated countries and look at the average death count in 1996 on the x-axis, the average death count in 2001 on the y-axis, and color the points by the average death count in 2006.
+The darker the points are the higher the death count was in 2006.
+To get a closer look at the average death count in all 3 years we created individualized bar graphs for 1996, 2001, and 2006.
+
+Low Population first decade (1996-2006).
+```{r}
 #Low Population Deaths 1996-2006
@@
   scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
+  xlab("Country")+
+  ylab("1996 Deaths")+
   geom_col(position = 'dodge')
 
-interact_low_first<- p_low_first + labs(title="High Populated Countries", x="1996 Deaths")
+interact_low_first<- p_low_first + labs(title="Low Populated Countries")
 
@@
   scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
+  xlab("Country")+
+  ylab("2001 Deaths")+
   geom_col(position = 'dodge')
 
-interact_low_first<- p_low_first + labs(title="High Populated Countries", x="1996 Deaths")
+interact_low_first<- p_low_first + labs(title="Low Populated Countries")
 
@@
   scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
+  xlab("Country")+
+  ylab("2006 Deaths")+
   geom_col(position = 'dodge')
 
-interact_low_first<- p_low_first + labs(title="High Populated Countries", x="1996 Deaths")
+interact_low_first<- p_low_first + labs(title="Low Populated Countries")
 
@@
 
-This shows the second decade 2007-2017.
+In the first graph we faceted the high-populated countries and look at the average death count in 2007 on the x-axis, the average death count in 2012 on the y-axis, and color the points by the average death count in 2017.
+The darker the points are the higher the death count was in 2017.
+To get a closer look at the average death count in all 3 years we created individualized bar graphs for 2007, 2012, and 2017.
+
+High Population second decade (2007-2017).
 ```{r, echo=FALSE}
@@
 
-interact_high_second<- p_high_second + labs(title="High Populated Countries", x="2007 Deaths", y="2017 Deaths")
+interact_high_second<- p_high_second + labs(title="High Populated Countries", x="2007 Deaths", y="2012 Deaths")
 
@@
   scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
+  xlab("Country")+
+  ylab("2007 Deaths")+
   geom_col(position = 'dodge')
 
-interact_high_second_07<- p_high_second_07 + labs(title="High Populated Countries", x="2007 Deaths")
+interact_high_second_07<- p_high_second_07 + labs(title="High Populated Countries")
 
@@
   scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
+  xlab("Country")+
+  ylab("2012 Deaths")+
   geom_col(position = 'dodge')
 
-interact_high_second_12<- p_high_second_12 + labs(title="High Populated Countries", x="2012 Deaths")
+interact_high_second_12<- p_high_second_12 + labs(title="High Populated Countries")
 
@@
   scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
+  xlab("Country")+
+  ylab("2017 Deaths")+
   geom_col(position = 'dodge')
 
-interact_high_second_17<- p_high_second_17 + labs(title="High Populated Countries", x="2017 Deaths")
+interact_high_second_17<- p_high_second_17 + labs(title="High Populated Countries")
 
 ggplotly(interact_high_second_17)
-
+```
+
+***
+In the first graph we faceted the low-populated countries and look at the average death count in 2007 on the x-axis, the average death count in 2012 on the y-axis, and color the points by the average death count in 2017.
+The darker the points are the higher the death count was in 2017.
+To get a closer look at the average death count in all 3 years we created individualized bar graphs for 2007, 2012, and 2017.
+
+Low Population second decade (2007-2017)
+```{r}
 #Low Population Deaths 2007-2017
@@
 
-interact_low_second<- p_low_second + labs(title="Low Populated Countries", x="2007 Deaths", y="2017 Deaths")
+interact_low_second<- p_low_second + labs(title="Low Populated Countries", x="2007 Deaths", y="2012 Deaths")
 
@@
   scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
+  xlab("Country")+
+  ylab("2007 Deaths")+
   geom_col(position='dodge')
 
-interact_low_second_07<- p_low_second_07 + labs(title="Low Populated Countries", x="2007 Deaths")
+interact_low_second_07<- p_low_second_07 + labs(title="Low Populated Countries")
 
@@
   scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
+  xlab("Country")+
+  ylab("2012 Deaths")+
   geom_col(position='dodge')
 
-interact_low_second_12<- p_low_second_12 + labs(title="Low Populated Countries", x="2012 Deaths")
+interact_low_second_12<- p_low_second_12 + labs(title="Low Populated Countries")
 
@@
   scale_color_gradient2(low = "light pink", mid = "pink", high = "dark violet", aesthetics = "color") +
+  xlab("Country")+
+  ylab("2017 Deaths")+
   geom_col(position='dodge')
 
-interact_low_second_17<- p_low_second_17 + labs(title="Low Populated Countries", x="2017 Deaths")
+interact_low_second_17<- p_low_second_17 + labs(title="Low Populated Countries")
 

### 2022-05-10 Final Report.Rmd
indoor deaths

@@
 
-Indoor Deaths
+Let's focus on Indoor Deaths first:
 ```{r, echo=FALSE}
@@
 
-#Pakistan and Malawi were both mainly affected in 1996
 indoor_top<- deaths_df %>% 
@@
 ```
+It is interesting to see the decrease in both the high and low-populated countries over time.
+It is easier to see the difference in Malawi's indoor deaths.
+Pakistan and Malawi were the two countries with the highest indoor death count in our subset.
+Despite being a low-populated country, Malwai had much larger death counts.
 

### 2022-05-10 Final Report.Rmd
outdoor death

@@
 ```
+
 It is interesting to see the decrease in both the high and low-populated countries over time.
@@
 
-Outdoor Deaths
+Outdoor Deaths: 
 ```{r, echo= FALSE}
@@
 
-#Serbia and Pakistan were the top countries affected. 
-#2011 Pakistan had the greatest amount of Outdoor Deaths
 outdoor_top<- deaths_df %>% 
@@
 ```
+
+For outdoor deaths you can see that for high populated countries Pakistan is in the lead.
+There tends to be an interesting increase in Nigeria from 2014-2015.
+Also Germany and the United States are very neck in neck over the years.
+In the lower populated countries you can see that Serbia is in the lead.
+Sri Lanka does have a steep decline from 2015 to 2016 but before that it was really close with Tonga.
+When you look at the top 2 countries, Pakistan and Serbia, you can see that Pakistan is drastically higher.
+2011 was the greatest number of deaths for Pakistan.
+1997 was the greatest number of deaths for Serbia
 

### 2022-05-10 Final Report.Rmd
ozone

@@
 ```
+In the high-populated countries, the U.S.
+is the second-highest in ozone-related deaths and even increases over some years.
+We also see an increase in deaths in Pakistan, Chile, and Sri Lanka.
+Malawi is still one of the highest countries for ozone-related deaths, even though it saw the most decrease over time.
 

### 2022-05-10 Final Report.Rmd
finally the end!

--- 2022-05-10 Final Report.Rmd
@@
 author: "Alison Cheek & Elizabeth Bekele"
-date: '2022-05-06'
+date: '2022-05-10'
 output: html_document
 ---
-In our initial discussion over choosing a topic for our project, we narrowed it down to environmental-related data because we were interested in seeing possible trends over time and the vast quantity of environmental data is available.
+In our initial discussion over choosing a topic for our project, we narrowed it down to environmental-related data because we were interested in seeing possible trends over time and the vast quantity of environmental data that is available helped pique our curiosity.
 
@@
 
-* Overview
-  + We will see how different air pollution types affect the population 
+* After selecting our deaths data frame due to air pollution we were able to compile a list of questions to explore
+
+* Topics of Interest
+  + How different air pollution types affect the population 
   + compare past and present population numbers
@@
 
-We cannot conclude which is worse.
-
-* High Populated Countries: 
-  + Outdoor pollution seems to be more detrimental with the exception of two countries in this sample set.
-
-* Low Populated Countries: 
-  + it's half-and-half
+In the high populated countries we can see that 2 countries exhibit higher indoor air pollutant deaths whereas the others depict higher outdoor deaths.
+In the lower populated countries you can see that half show higher outdoor air pollutant deaths and half show higher indoor air pollutant deaths.
+Therefore, we cannot conclude which is worse.
 
@@
 * Which is worse - outdoor or indoor pollution?
-  + Inconclusive
+  + Based on our sample set, it is inconclusive
+  + "According to the EPA (Environmental Protection Agency), however, the levels of indoor air pollutants are often 2 to 5 times higher than outdoor levels, and in some cases these levels can exceed 100 times that of outdoor levels of the same pollutants.
+In other words, sometimes the air inside can be more harmful than the air outside."
+
+There are so many factors affecting environmental data, another area of interest would be to look into the cultural, geographical, and historical climate of the country to see if any significant events occurred to skew the numbers during those years where great increases and decreases occurred.
+
+To make the data more manageable we had to reduce the number of countries we looked at.
+This is why we selected one high populated country and one low populated country from each continent.
+We thought having a high and low populated country would give us a general idea on how air pollution affected each continent.
+Furthermore, we excluded Russia, China, and India because their populations were significantly large and wouldn't be a good representation of the average populations.
+
+Overall this project has sparked interest in environmental concerns.
+After seeing the impact we have in our environment's air pollution we are prone to be more conscious of how we treat our planet.
+We hope that this data has brought more awareness to others as well!
 

### 2022-05-11 Final Report.Rmd

@@
 ---
-title: "Final Report"
+title: "Final Report: Air Pollution Deaths"
 author: "Alison Cheek & Elizabeth Bekele"

### 2022-05-11 population_data.Rmd

@@
 
+#Let's compare the first decade 1996-2006 with second decade 2007-2017 
 high_pop_countries <- world_pop %>% 
-  filter(Year > 1999 & Country.Name %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
+  filter(Year > 1996 & Country.Name %in% c('United States', 'Brazil', 'Nigeria', 'Germany', 'Pakistan', 'Australia')) %>% 
   group_by(Year)
@@
 low_pop_countries <- world_pop %>% 
-  filter(Year > 1999 & Country.Name %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand'))%>% 
+  filter(Year > 1996 & Country.Name %in% c('Canada', 'Chile', 'Malawi', 'Serbia', 'Sri Lanka', 'New Zealand'))%>% 
   group_by(Year)
@@
 
-head(world_pop)
+#General overview of data 
+head(world_pop, 12)
 
